---
title: "Phylowave DENV"
# Ref: from Noémie Lefrancq https://www.medrxiv.org/content/10.1101/2023.12.23.23300456v1.full
# date: 2024-09-10
# author Douglas Fritz 2024-Nov-05
output:
  github_document:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Input


## Load codes and DENV data

#### Load index functions

```{r, echo = F, eval=T}
## Not printed, set directory
#setwd('~/Dropbox/THD/Index_paper_methods_Sc2_H3N2_BP_TB/paper-index-fitness-dynamics-trees/')
```

First, source all the necessary functions:
```{r, eval=T}
source(file = '2_Functions/2_1_Index_computation_20240909.R')
source(file = '2_Functions/2_2_Lineage_detection_20240909.R')
source(file = '2_Functions/2_3_Lineage_fitness_20240909.R')

# Function to extract the start of date ranges
extract_start_date <- function(date_entry) {
  # Check if the entry contains a range
  if (str_detect(date_entry, "\\(")) {
    # Extract the date before the '('
    start_date <- str_extract(date_entry, "^[^\\s]+")
  } else {
    # It's a single date
    start_date <- date_entry
  }
  
  return(start_date)
}
```

Load necessary packages
```{r, echo = T, eval=T, , results = 'hide', warning=FALSE, message=FALSE}
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("ggtree")

# install.packages("remotes")
# remotes::install_github(repo = "stan-dev/cmdstanr")

library(ape, quiet = T); library(phytools, quiet = T); library(stringr, quiet = T)
library(MetBrewer, quiet = T); library(parallel, quiet = T); library(mgcv, quiet = T)
library(cowplot, quiet = T); library(ggplot2, quiet = T); library(ggtree, quiet = T);
library(cmdstanr, quiet = T); library(binom, quiet = T); library(plotly) ; library(DescTools); 
library(readr, quiet = T); library(glue, quiet = T)
```

```{r debug area, eval = FALSE}
# names_seqs = data.frame(tree_DENV$tip.label)
# colnames(names_seqs) = c("strain")
# 
# matched_data <- names_seqs %>%
#   left_join(metadata %>% select(strain, start_date), by = "strain")
# times_seqs <- matched_data$start_date
# 
# times_seqs = data.frame(times_seqs)

# times_seqs <- data.frame(metadata$start_date)
# 
# times_seqs <- times_seqs$metadata.start_date
# 
# any(is.na(matched_indices)) #Should be false
# names_seqs[is.na(matched_indices)]  # Sequences without a match
# metadata$strain[is.na(matched_indices)]  # Strains without a match
# summary(metadata)
# summary(dataset_with_nodes)

#tip_matches <- grep("8546/93DEN2", tree_DENV$tip.label, value = TRUE)
#node_matches <- grep("8546/93DEN2", tree_DENV$node.label, value = TRUE)


```
 
 
 
#### Load data
Load the tree, in which all the tip name include: collection time, location and Pango lineage 
```{r, eval=T}
# tree_DENV = read.nexus('1_Data/1_5_DENV/nextstrain_dengue_all_genome_timetree.nexus')
# 
# #Tree had entries that messed with the nexus syntax: eg "(<names>)", ":", "," "-" in names have to be removed

##Nexus dowload tree
# tree_DENV = read.nexus('1_Data/1_5_DENV/nextstrain_dengue_all_genome_timetree_fixed.nexus')

##Ape created tree from nexus
tree_DENV = read.nexus("1_Data/1_5_DENV/nextstrain_dengue_all_genome_timetree_apefixed.nexus.nex")


## Make sure the tree is binary, and ladderized
tree_DENV = collapse.singles(ladderize(multi2di(tree_DENV, random = F), right = F))

#Some trees have numeric isolates, those will look like nodes to the model and will throw warnings.
#data cleaning: retrieve all the number only samples
numeric_labels <- grep("^[0-9]+$", tree_DENV$tip.label, value = TRUE)
# Add I (Isolate) to the front of them
tree_DENV$tip.label[tree_DENV$tip.label %in% numeric_labels] <- paste0("I", numeric_labels)

## Names all sequences
names_seqs = tree_DENV$tip.label
n_seq = length(names_seqs)

## Collection times of all sequences Only works with Nexus time trees, will need to add them via metadata
metadata_file = read_tsv("1_Data/1_5_DENV/nextstrain_dengue_all_genome_metadata.tsv")


#Extract the start of the date range 
metadata_file <- metadata_file %>%
  mutate(start_date = sapply(date, extract_start_date)) %>%
  mutate(start_date = as.Date(start_date, format="%Y-%m-%d")) 

#Remove spaces in the metadata to match tree tips
metadata_file$strain <- gsub(" ", "", metadata_file$strain)

#Apply the I to the numeric only samples like we did with the tree
metadata_file$strain[metadata_file$strain %in% numeric_labels] <- paste0("I", metadata_file$strain[metadata_file$strain %in% numeric_labels])

#Match metadata strains to tree labels
matched_indices <- match(names_seqs, metadata_file$strain)

#Assign times to times_seqs
times_seqs <- metadata_file$start_date[matched_indices]
times_seqs <- Date2decimal(times_seqs)

#Assign clade and serotypes 
clades_seqs <- metadata_file$genotype_nextclade[matched_indices]
serotype_seqs <- metadata_file$serotype_genbank[matched_indices]


## Nextstrain clades of all sequences
# clades_seqs = sapply(names_seqs, function(x)tail(str_split(x, pattern = '/')[[1]],1))

# #For other trees
# metadata_file <- read.csv("1_Data/1_4_TB/List_all_isolates_TB_Samara_Russia_20231216.csv")
# matched_indices <- match(names_seqs, metadata_file$Isolate)
# times_seqs <- metadata_file$Collection_time[matched_indices]
# clades_seqs <- metadata_file$Clade[matched_indices]

# # Convert times_seqs to numeric if necessary
# times_seqs <- as.numeric(times_seqs)
```


## Compute the index dynamics



To compute the index of all nodes, use the function *compute.index*, you will need different inputs: 

1. Data: 
    * the *timed_tree*
    * *metadata* dataframe (see below the details of the *dataset_with_nodes*)
    * *distance matrix*: can be computed from the timed tree with the function *dist.nodes.with.names*
2. Information on the pathogen genome (these are pathogen specific):
    * The genome length of the pathogen considered: *genome_length* (in bp).
    * The mutation rate of the pathogen considered: *mutation_rate* (in bp/genome/year), an average is fine.
      
3. Index parameters (these are both pathogen-specific and dataset-specific):
    * The timescale: *timescale* (in years), this will be used to compute the bandwidth, see more details below.
    * Window of time on which to search for samples in the population: *wind*, see more details below.

The function outputs a vector containing the index of each node (internal and terminal).


**timescale**: The timescale determines the kernel which enables to track lineage emergence dynamically, focusing on short distances between nodes (containing information about recent population dynamics) rather than long distances (containing information about past evolution). The timescale is tailored to the specific pathogen studied and its choice depends on the molecular signal, as well as the transmission rate. In the study, we used timescales ranging from months (typical of RNA viruses) to years (typical of bacteria). To determine a timescale suitable for your dataset, we recommend thinking about the generation time of the pathogen considered, its mutation rate, and the amount of diversity already accumulated. For example, at the time of the analysis, SARS-CoV-2 was a new pathogen, spreading quickly and accumulating diversity at a rate of ~2 mutations per month. Therefore, a small timescale of less than a year chosen (0.15 years). On the contrary, *Mycobacterium tuberculosis* is an older and relatively slowly spreading pathogen, which accumulates mutations at a rate of ~0.2 mutation per year. A much larger timescale was then chosen (30 years), to reflect this. Ultimately, the best timescale is one that maximises the visualisation of population dynamics. We recommend trying different values.

**wind**: The choice of *wind* will depend on the sampling intensity of the dataset. It defines the window of time around each node on which to search for samples in the population. Ultimately it smooths the index dynamics. As a mean of example, for SARS-CoV-2, we set *wind* to 15 days, as the dataset was intensely sampled. But for *Bordetella pertussis*, which is more sparsely sampled, we chose a *wind* of 1 year. If *wind* is too large, then all the nodes are considered to be part of the same time window. If *wind* is too small, then only the nodes in direct proximity of the node of interest will be considered in the time window, which can result in noisy index dynamics. We recommend choosing a *wind* value that enables to span multiple sampling times, for example if you have samples and nodes every week, you may choose a *wind* of ~1-2 months. If you have samples and nodes every year, you may choose a *wind* of ~2 years.



###Set the index parameters.
```{r, eval=T}
## Length genome 
genome_length = 10649 # reference  https://nextstrain.org/dengue/all/genome
## Mutation rate 
mutation_rate = 7.6e-4 # mutation rate #ref https://pmc.ncbi.nlm.nih.gov/articles/PMC9030598/

## Parameters for the index
timescale = 30 ## Timescale in years; raises index 
## Window of time on which to search for samples in the population
wind = 365 #days 
wind = wind/365 #Wind smooths lines


```


#### Compute pairwise distance matrix
Compute distance between each pair of sequences and internal nodes in the tree
```{r, eval=T}
genetic_distance_mat = dist.nodes.with.names(tree_DENV)
```

Get the time of each internal node
```{r, eval=T}
nroot = length(tree_DENV$tip.label) + 1 ## Root number
distance_to_root = genetic_distance_mat[nroot,]
root_height = times_seqs[which(names_seqs == names(distance_to_root[1]))] - distance_to_root[1]
nodes_height = root_height + distance_to_root[n_seq+(1:(n_seq-1))]
```

#### Preparation data tips and nodes

Prepare the main dataframe, where the index and lineages of all nodes (internal and terminal) are going to be stored.
```{r, eval=T}
# Meta-data with all nodes 
dataset_with_nodes = data.frame('ID' = c(1:n_seq, n_seq+(1:(n_seq-1))),
                                'name_seq' = c(names_seqs, n_seq+(1:(n_seq-1))),
                                'time' = c(times_seqs, nodes_height),
                                'is.node' = c(rep('no', n_seq), rep('yes', (n_seq-1))),
                                'clade' = c(clades_seqs, rep(NA, n_seq-1)))

# 
#                                 'serotype' = c(serotype_seqs, rep(NA, n_seq-1)))

# Maybe add serotype once it works?
```



#### Compute index of every tip and node
```{r, eval=T}
# debug(compute.index)
dataset_with_nodes$index = compute.index(time_distance_mat = genetic_distance_mat, 
                                         timed_tree = tree_DENV, 
                                         time_window = wind,
                                         metadata = dataset_with_nodes, 
                                         mutation_rate = mutation_rate,
                                         timescale = timescale,
                                         genome_length = genome_length)


```



#### Plot tree & index below, with colors from clades

First, generate the color key, based on the Nextstrain clade of each sequence.
```{r, eval=T}
## Color key for Nextstrain clades
colors_clade = met.brewer(name="Cross", n=length(levels(as.factor(dataset_with_nodes$clade))), type="continuous")

## Color of each node, based on the key
dataset_with_nodes$clade_color = as.factor(dataset_with_nodes$clade)
clade_labels = levels(dataset_with_nodes$clade_color)
levels(dataset_with_nodes$clade_color) = colors_clade
dataset_with_nodes$clade_color = as.character(dataset_with_nodes$clade_color)
```




Then plot the tree and index:
```{r, eval=T}
par(mfrow = c(2,1), oma = c(0,0,0,0), mar = c(4,4,0,0))

min_year = 1950
max_year = 2024

## Tree
plot(tree_DENV, show.tip.label = FALSE, 
     edge.color = 'grey', edge.width = 0.25,
     x.lim = c(min_year, max_year)-root_height)
tiplabels(pch = 16, col = dataset_with_nodes$clade_color, cex = 0.3)
axisPhylo_NL(side = 1, root.time = root_height, backward = F,
             at_axis = seq(min_year, max_year, 0.5)-root_height,
             lab_axis = seq(min_year, max_year, 0.5), lwd = 0.5)

## Index
plot(dataset_with_nodes$time, 
     dataset_with_nodes$index, 
     col = adjustcolor(dataset_with_nodes$clade_color, alpha.f = 1),
     bty = 'n', xlim = c(min_year, max_year), cex = 0.4,
     pch = 16, bty = 'n', ylim = c(0, 1), 
     main = paste0(''), 
     ylab = 'Index', xlab = 'Time (years)', xaxt = 'n', yaxt = 'n')
axis(2, las = 2, lwd = 0.5)
axis(1, lwd = 0.5)

# Color key
legend('topright', 
       legend = clade_labels,
       fill = colors_clade, border = colors_clade,
       cex = 0.5, bty = 'n', ncol = 5)
```


## Find DENGUE clades based on index dynamics 



#### Run the lineage detection algorithm on DENGUE data

Parameters for the detection: 
```{r, eval = T, eval=FALSE}
time_window_initial = 2030;
time_window_increment = 100;
p_value_smooth = 0.05
weight_by_time = 0.1
k_smooth = -1
plot_screening = F
min_descendants_per_tested_node = 30
min_group_size = 30
weighting_transformation = c('inv_sqrt')

parallelize_code = T
number_cores = 8

max_stepwise_deviance_explained_threshold = 0
max_groups_found = 13
stepwise_AIC_threshold = 0

keep_track = T
```


Run the detection function (this steps takes approximately <10 min on 2 cores):
```{r, eval=FALSE, results = 'hide', warning=FALSE, message=FALSE}
start_time = Sys.time()
potential_splits = find.groups.by.index.dynamics(timed_tree = tree_DENV,
                                                 metadata = dataset_with_nodes,
                                                 node_support = tree_DENV$edge.length[match((n_seq+1):(2*n_seq-1),
                                                                                            tree_DENV$edge[,2])],
                                                 threshold_node_support = 1/(29903*0.00081),
                                                 time_window_initial = time_window_initial,
                                                 time_window_increment = time_window_increment,
                                                 min_descendants_per_tested_node = min_descendants_per_tested_node,
                                                 min_group_size = min_group_size,
                                                 p_value_smooth = p_value_smooth,
                                                 stepwise_deviance_explained_threshold = max_stepwise_deviance_explained_threshold,
                                                 stepwise_AIC_threshold = stepwise_AIC_threshold,
                                                 weight_by_time = weight_by_time,
                                                 weighting_transformation = weighting_transformation,
                                                 k_smooth = k_smooth,
                                                 parallelize_code = parallelize_code,
                                                 number_cores = number_cores, 
                                                 plot_screening = plot_screening,
                                                 max_groups_found = max_groups_found, 
                                                 keep_track = keep_track)
end_time = Sys.time()
print(end_time - start_time)
```

Instead, you may wish to load or write the results:
```{r, eval = T}
# potential_splits = readRDS('README_files/potential_splits.rds')

##Write
#file_name <- glue("3_Outputs/potential_splits_{timescale}_{wind}.rds") 
#write_rds(potential_splits, file_name)
```

Look at the deviance explained by the models with different number of groups. Here for simplicity we directly chose max_groups_found = 13, which is the number of groups used in the original analysis. To decide on 13 groups, we initially ran the algorithm up to 30 groups.
```{r, eval = T}
df_explained_dev = data.frame('N_groups' = 0:length(potential_splits$best_dev_explained),
                              'Non_explained_deviance' = (1-c(potential_splits$first_dev, potential_splits$best_dev_explained)),
                              'Non_explained_deviance_log' = log(1-c(potential_splits$first_dev, potential_splits$best_dev_explained)))
df_explained_dev$Non_explained_deviance_log = df_explained_dev$Non_explained_deviance_log-min(df_explained_dev$Non_explained_deviance_log)

par(mfrow = c(1,2), oma = c(2,2,1,1), mar = c(2,2,2,0.5), mgp = c(0.75,0.25,0), cex.axis=0.5, cex.lab=0.5, cex.main=0.7, cex.sub=0.5)
plot(df_explained_dev$N_groups,
     df_explained_dev$Non_explained_deviance,
     bty = 'n', ylim = c(0, ceiling(10*max(df_explained_dev$Non_explained_deviance))/10),
     xaxt = 'n', yaxt = 'n', pch = 16, main = 'linear scale', cex = 0.5, 
     ylab = 'Non-explained deviance (%)', xlab = 'Number of groups')
axis(1, lwd = 0.5, tck=-0.02)
axis(2, las = 2, at = seq(0,ceiling(10*max(df_explained_dev$Non_explained_deviance))/10,0.1),
     labels = seq(0, ceiling(10*max(df_explained_dev$Non_explained_deviance))/10,0.1)*100, lwd = 0.5, tck=-0.02)

plot(df_explained_dev$N_groups,
     (df_explained_dev$Non_explained_deviance),
     log = 'y',
     ylim = c(0.01, 1),
     bty = 'n',
     xaxt = 'n', yaxt = 'n', pch = 16, main = 'log scale', cex = 0.5, 
     ylab = 'Non-explained deviance (%) - log scale', xlab = 'Number of groups')
axis(1, lwd = 0.5, tck=-0.02)
axis(2, las = 2, at = c(0.01, 0.1, 0.25, 0.5, 1),
     labels = c(0.01, 0.1, 0.25, 0.5, 1)*100, lwd = 0.5, tck=-0.02)
```


Optimize the number of groups: set the minimum number of sequences per group to 30, with a minimum frequency of 1%.
```{r, eval = T}
split = merge.groups(timed_tree = tree_DENV, metadata = dataset_with_nodes, 
                     initial_splits = potential_splits$potential_splits, 
                     group_count_threshold = 30, group_freq_threshold = 0.01)
```

Label sequences with these new groups, and assign a color to each of them.
```{r, eval = T}
## Label sequences with new groups
dataset_with_nodes$groups = as.factor(split$groups)
## Reorder labels by time of emergence
name_groups = levels(dataset_with_nodes$groups)
time_groups_world = NULL
for(i in 1:length(name_groups)){
  time_groups_world = c(time_groups_world, min(dataset_with_nodes$time[which(dataset_with_nodes$groups == name_groups[i] &
                                                                                   dataset_with_nodes$is.node == 'no')]))
}
levels(dataset_with_nodes$groups) = match(name_groups, order(time_groups_world, decreasing = T))
dataset_with_nodes$groups = as.numeric(as.character(dataset_with_nodes$groups))
dataset_with_nodes$groups = as.factor(dataset_with_nodes$groups)
## Update names in split list
split$tip_and_nodes_groups = match(split$tip_and_nodes_groups, order(time_groups_world, decreasing = T))
names(split$tip_and_nodes_groups) = 1:length(split$tip_and_nodes_groups)
split$groups = as.factor(split$groups)
levels(split$groups) = match(name_groups, order(time_groups_world, decreasing = T))
split$groups = as.numeric(as.character(split$groups))
## Choose color palette
n_groups <- length(name_groups)
colors_groups = (met.brewer(name="Cross", n=n_groups, type="continuous"))
## Color each group
dataset_with_nodes$group_color = dataset_with_nodes$groups
levels(dataset_with_nodes$group_color) = colors_groups
dataset_with_nodes$group_color = as.character(dataset_with_nodes$group_color)
```


#### Plot tree & index below, with colors from index-defined groups

Plot the tree and index colored with the new groups:
```{r, eval = T}
par(mfrow = c(2,1), oma = c(0,0,0,0), mar = c(4,4,0,0))

## Tree
tree_plot <- plot(tree_DENV, show.tip.label = FALSE, 
     edge.color = 'grey', edge.width = 0.25,
     x.lim = c(min_year, max_year)-root_height)
tiplabels(pch = 16, col = dataset_with_nodes$group_color, cex = 0.3)
axisPhylo_NL(side = 1, root.time = root_height, backward = F,
             at_axis = seq(min_year, max_year, 0.5)-root_height,
             lab_axis = seq(min_year, max_year, 0.5), lwd = 0.5)

tree_plot

## Index colored by group
index_plot <- plot(dataset_with_nodes$time, 
     dataset_with_nodes$index, 
     col = adjustcolor(dataset_with_nodes$group_color, alpha.f = 1),
     bty = 'n', xlim = c(min_year, max_year), cex = 0.5,
     pch = 16, bty = 'n', #ylim = c(0, 1), 
     main = paste0(''), #log = 'y',
     ylab = 'Index', xlab = 'Time (years)', yaxt = 'n')

axis(2, las = 2)
# Color key
legend('topright', 
       legend = name_groups,
       fill = colors_groups, border = colors_groups,
       cex = 0.5, bty = 'n', ncol = 5)


index_plot

```



#### Compare NextStrain groups and groups called with the index

Generate trees colored with each set of groups next to each other:
```{r, eval = T, results = 'hide', warning=FALSE, message=FALSE}
## Tree with index-defined groups
groups = matrix(dataset_with_nodes$groups[which(dataset_with_nodes$is.node == 'no')], ncol = 1)
colnames(groups) = 'groups'
rownames(groups) = dataset_with_nodes$name_seq[which(dataset_with_nodes$is.node == 'no')]
cols = as.character(colors_groups)
names(cols) = as.character(1:max(as.numeric(name_groups)))
plot_tree_world_groups <- ggtree(tree_DENV, mrsd=lubridate::date_decimal(max(times_seqs)), size = 0.10,
                   aes(color = as.character(dataset_with_nodes$groups), label = label)) + 
  scale_color_manual(values = cols)+theme_tree2()
plot_tree_world_groups = gheatmap(plot_tree_world_groups, groups, offset=0.1, width=0.10, 
                    colnames=FALSE, legend_title="Group", color=NA) +
  scale_fill_manual(values = (cols))+scale_y_continuous(expand=c(0, 0.3))+theme(legend.position = 'none')

## Tree with NextStrain clades
Nextstrain = matrix(dataset_with_nodes$clade[which(dataset_with_nodes$is.node == 'no')], ncol = 1)
colnames(Nextstrain) = 'groups'
rownames(Nextstrain) = dataset_with_nodes$name_seq[which(dataset_with_nodes$is.node == 'no')]
cols_NextStrain = as.character(colors_clade)
names(cols_NextStrain) = clade_labels
plot_tree_world_Nextstrain <- ggtree(tree_DENV, mrsd=lubridate::date_decimal(max(times_seqs)), size = 0.10,
                                          aes(color = as.character(dataset_with_nodes$clade), label = label)) + 
  scale_color_manual(values = cols_NextStrain)+
  theme_tree2(legend = 'none')
plot_tree_world_Nextstrain = gheatmap(plot_tree_world_Nextstrain, Nextstrain, offset=0.1, width=0.10, 
                     colnames=FALSE, legend_title="Group", color=NA) +
  scale_fill_manual(values = cols_NextStrain, na.value = 'white')+
  scale_x_reverse() + 
  scale_y_continuous(expand=c(0, 0.3))+
  theme(legend.position = 'none')
```

Plot the generated  trees:
```{r, eval = T}
plot_grid(plot_tree_world_groups, plot_tree_world_Nextstrain,
          rel_widths = c(1, 1), labels = c('Automatic clades', 'NextStrain clades'), label_size = 10, label_x = c(0.1, 0.25), ncol = 2)

ggplotly(subplot(plot_tree_world_groups, plot_tree_world_Nextstrain,
         widths = c(0.5, 0.5)))
```


## Quantify the fitness of detected DENV lineage

#### Run the fitness model
Quantify the fitness of each group you can run the code (this steps takes approximately <5 min on 3 cores):
```{r, eval = F, results = 'hide', warning=FALSE, message=FALSE}
start_time = Sys.time()
## Load and compile stan code (this can take a few minutes)
## If you get an error about path mapping to a previous user 
##make sure to go into the model/scripts folder and delete the old unix model before making this one
model_compiled <- cmdstan_model(stan_file = '2_Functions/Model_multinomial_logistic_birthdeath_lineage_fitness_20231220.stan')
## Run model on the groups
res_fitness = estimate_rel_fitness_groups_with_branches(dataset_with_nodes = dataset_with_nodes,
                                                        tree = tree_DENV,
                                                        min_year = 2020, 
                                                        window = 30/365,
                                                        model_compiled = model_compiled,
                                                        iter_warmup = 250, iter_sampling = 500, refresh = 50, seed = 1)
end_time = Sys.time()
print(end_time - start_time)
```
You might encounter a warning saying that '*alpha_true_GA*' has a missing init value - this is normal as those groups (ancestral groups that are not present at the start of the time series) do not always exist and therefore there is no default initial value. This is does not impact the model run. The seed has been set to 1 so allow for reproducible results.

To save some time, you may wish to load the results:
```{r, eval = T}
# res_fitness = readRDS('README_files/res_fitness.rds')
##Write
file_name_fitness <- glue("3_Outputs/res_fitness_{timescale}_{wind}.rds") 
write_rds(res_fitness, file_name_fitness)

```

#### Plot the fits and estimated parameters
Plot the fits:
```{r}
order_colors = order(as.numeric(split$tip_and_nodes_groups))
colour_lineage = colors_groups[match(split$tip_and_nodes_groups[order_colors], name_groups)]

plot_fit_data_new(data = res_fitness$data,
                  Chains = res_fitness$chains,
                  colour_lineage = colour_lineage,
                  xmin = 2020, xmax = 2023.5)
```

Plot the predicted vs observed proportions:
```{r}
plot_observed_vs_predicted(data = res_fitness$data,
                           Chains = res_fitness$chains,
                           colour_lineage = colour_lineage)
```

Plot raw fitness estimates:
```{r}
plot_estimated_fitness_ref_ancestral(data = res_fitness$data,
                                     Chains = res_fitness$chains,
                                     colour_lineage = colour_lineage, 
                                     gentime = 1)
```






