---
title: "Streamlined Analysis and Data Parsing"
author: "Douglas Fritz"
date: "2025-10-13"
# Ref: from Noémie Lefrancq https://www.medrxiv.org/content/10.1101/2023.12.23.23300456v1.full
output: html_document
# output:
#   github_document:
#     toc: false
# editor_options: 
#   markdown: 
#     wrap: 72
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Input

## Load codes and DENV data

#### Load index functions

```{r here setup, echo = F, eval=T}
## Not printed, set directory
here::i_am("DENV_Streamlined.Rmd")
st <- format(Sys.time(), "%Y-%m-%d_%H%M")


```

### Load necessary packages

```{r packages, echo = T, eval=T, , results = 'hide', warning=FALSE, message=FALSE}
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("ggtree")

# install.packages("remotes")
# remotes::install_github(repo = "stan-dev/cmdstanr")

library(ape, quiet = T); library(phytools, quiet = T); library(stringr, quiet = T)
library(MetBrewer, quiet = T); library(parallel, quiet = T); library(mgcv, quiet = T)
library(cowplot, quiet = T); library(ggplot2, quiet = T); library(ggtree, quiet = T);
library(cmdstanr, quiet = T); library(binom, quiet = T); library(plotly) ; library(DescTools); 
library(readr, quiet = T); library(glue, quiet = T)
library(grid)
library(png)
library(vDiveR)
library(vcfR)
library(reticulate)
library(tidyverse)
library(foreach)
library(extrafont)
library(dsmisc)
library(phangorn)
library(glue)
library(readr)
library(svglite)
library(here)
library(svglite)
library(teal.modules.general)
library(treedataverse)

# font_import()
loadfonts(device="all") 

```

First, source all the necessary functions:

```{r functions, eval=T}
source(file = '2_Functions/2_1_Index_computation_20240909.R')
source(file = '2_Functions/2_2_Lineage_detection_20240909.R')
source(file = '2_Functions/2_3_Lineage_fitness_20240909.R')
source(file = '2_Functions/DF_edits_create_vcf_from_fasta.R')

# Function to extract the start of date ranges
extract_start_date <- function(date_entry) {
  # Check if the entry contains a range
  if (str_detect(date_entry, "\\(")) {
    # Extract the date before the '('
    start_date <- str_extract(date_entry, "^[^\\s]+")
  } else {
    # It's a single date
    start_date <- date_entry
  }
  
  return(start_date)
}


#Reconstructs ancestral states (e.g., mutation presence/absence) along phylogenetic trees based on SNP data.

reconstruct_node_states = function(tree, dataset_tips, dataset_with_nodes, min_prop, max_prop, names_seqs){
  # Meta-data with nodes 
  dataset_with_inferred_resonstruction = dataset_with_nodes[,1:4]
  possible_snps = names(dataset_tips[,4:ncol(dataset_tips)])
  
  ## Filter SNP on frequency: only keep SNP that are present in:
  ## >1% of dataset
  ## <99% of dataset
  prevalence = t(apply(dataset_tips[,4:ncol(dataset_tips)], MARGIN = 2, function(x)table(factor(x, levels = c("0", "1")))))
  prevalence_prop = prevalence[,2]/(prevalence[,1]+prevalence[,2])
  prevalence = cbind(prevalence, prevalence_prop)
  possible_snps = names(which(prevalence[,3] >= min_prop & prevalence[,3] <= max_prop))
  
  ## Filter dataset_tips accordingly
  a = which(is.na(match(colnames(dataset_tips), possible_snps)) == F)
  dataset_tips = dataset_tips[,c(1:3, a)]
  
  print(paste0('Going though ', length(possible_snps), ' snps'))
  k=1
  for(i in possible_snps){
    print(paste0(k, ' / ', length(possible_snps), ' snps'))
    
    snp_data = dataset_tips[,which(colnames(dataset_tips) == i)]
    snp_data = as.factor(snp_data)
    names(snp_data) = names_seqs
    
    ## Perform reconstruction for this position
    # rec = phytools::fastAnc(tree = tree, x = snp_data, CI = TRUE, vars = T) ## factAnc lots faster than ace.ML
    rec = ape::ace(phy = tree, x = snp_data, method = "pic")
    rec_all = c(snp_data, rec$ace) ## List of all states: first all tips, then all nodes
    
    ## Find first state, to set it to 0, always
    first_state = rec_all[which(dataset_with_inferred_resonstruction$ID == length(names_seqs) + 1)]
    first_state = round(first_state, digits = 0)
    
    ## Write reconstruction in the big dataset
    if(first_state < 1.5){ ## First state is 0: all good
      dataset_with_inferred_resonstruction = cbind(dataset_with_inferred_resonstruction, rec_all - 1)     
    }
    if(first_state > 1.5){ ## First state is 1: have to change to 0
      dataset_with_inferred_resonstruction = cbind(dataset_with_inferred_resonstruction,  2 - rec_all)
    }
    
    ## Set column name to position name
    colnames(dataset_with_inferred_resonstruction)[which(colnames(dataset_tips) == i) +1] = i
    
    k=k+1
  }
  return(list('dataset_with_inferred_resonstruction' = dataset_with_inferred_resonstruction,
              'snp_prevalence' = prevalence,
              'possible_snps' = possible_snps))
}


## SNP defining Mutations Function
association_scores_per_group = function(dataset_with_nodes, dataset_with_inferred_reconstruction, tree, 
                                        possible_snps, upstream_window, downstream_window, 
                                        virus = NULL, gene = NULL, level = NULL){
  ## Set list to store results
  ## Extracts unique group names and initializes an empty list (scores) to store results for each group.
  group_names = levels(as.factor(dataset_with_nodes$groups))
  scores = as.list(rep(NA, length(group_names)-1))
  n_tips = length(tree$tip.label)
  
  
  
  ## For each group (except the initial group, which is the root), look at snp association
  for(j in 1:(length(group_names)-1)){ 
    cat("Working on group", j, "of", length(group_names)-1, "\n")
    
    ## Find members of the group and MRCA
    ## Identifies members of the current group and its MRCA. Ensures the MRCA is included in the list of members.
    members = dataset_with_nodes$ID[which(dataset_with_nodes$groups == group_names[j])]
    
    mrca_node = getMRCA(tree, dataset_with_nodes$name_seq[which(dataset_with_nodes$groups == group_names[j] & dataset_with_nodes$is.node == 'no')])
    if (is.null(mrca_node) || is.na(mrca_node)) {
      cat("⚠️ getMRCA failed for group", j, "in",dataset_with_nodes, "\n")
      next
      }
    members = unique(c(members, mrca_node))
    
    ## Update members list, with strains downstream, within the time window
    ## Identifies members of the current group and its MRCA. Ensures the MRCA is included in the list of members.
    time_mrca = dataset_with_nodes$time[which(dataset_with_nodes$ID == mrca_node)] ## Reference time for window
    # tmp = getDescendants(tree, mrca)
    # tmp = tmp[which(dataset_with_nodes$time[tmp] < time_mrca + downstream_window)]
    # members = unique(c(tmp, members))
    
    ## Update members list, with strains upstream, within the time window
    ancest = Ancestors(x = tree, node = mrca_node, type = 'all')
    
    if (length(ancest) == 0) {
      cat("⚠️ Group", j, "in", virus, gene, level, "has MRCA at root → no ancestors available\n")
      
      # Use just members for downstream analysis (no ancestors to add)
      all_nodes_of_interest <- members
    } else {
      # Normal case: Use members + valid ancestors
      time_ancest = dataset_with_nodes$time[ancest]
      time_ancest[1] = time_mrca
      tmp = which(time_ancest < time_mrca - upstream_window)
      if(length(tmp) > 0) ancest = ancest[-tmp]
      if(length(ancest) == 0) ancest = Ancestors(tree, node = mrca, type = 'all')[1]
      groups_ancest = dataset_with_nodes$groups[ancest]
      gr = min(as.numeric(as.character(groups_ancest)))
      ancest = ancest[which(groups_ancest == gr)]
      time_ancest = dataset_with_nodes$time[ancest]
      
      all_nodes_of_interest <- unique(c(ancest, members))
    }
    
    ## Branches of interest
    branches = tree$edge
    branches = branches[match(all_nodes_of_interest, branches[,2]),] ## Take all tips, from ancests, members
    tmp = which(is.na(match(branches[,1], c(ancest, members))))
    if(length(tmp) > 0){  branches = branches[-tmp, ]} ## Remove nodes that are not in ancest or members
    
    ## Branches time
    branches_time = branches
    branches_time[,1] = dataset_with_nodes$time[branches[,1]]
    branches_time[,2] = dataset_with_nodes$time[branches[,2]]
    
    ## Branches group (checked: ok)
    branches_group = branches
    branches_group[,1] = dataset_with_nodes$groups[branches[,1]]
    branches_group[,2] = dataset_with_nodes$groups[branches[,2]]
    
    ## Make branch group matrix binary: 1=group of interest, 0=other group (eg ancestral)
    branches_group[which(branches_group == j, arr.ind = T)] = 1
    branches_group[which(branches_group > j, arr.ind = T)] = 0
    
    snps = time_diff = snps_props_within = snps_props_whole = names_possibles_snps = NULL
    
    
    cat("👉 DEBUG: group", j, "(", group_names[j], ") in", virus, gene, level, "\n")
    cat("Number of members:", length(members), "\n")
    cat("Number of ancestors:", length(ancest), "\n")
    cat("Branches:", nrow(branches), "\n")
    cat("Possible SNPs:", length(possible_snps), "\n")
    ## Extracts SNP states for the start and end nodes of each branch for the current SNP.
    
    scores[[j]] <- numeric(length(possible_snps))
    
    for (i in 1:length(possible_snps)) {
      snp_name <- possible_snps[i]
      col_idx <- which(colnames(dataset_with_inferred_reconstruction) == snp_name)
      branches_snp <- branches
      branches_snp[, 1] <- dataset_with_inferred_reconstruction[branches[, 1], col_idx]
      branches_snp[, 2] <- dataset_with_inferred_reconstruction[branches[, 2], col_idx]
      
      if (length(ancest) == 0) {
        ancestral_state <- dataset_with_inferred_reconstruction[mrca_node, col_idx]
      } else {
        k <- 1
        ancestral_state <- branches_snp[which(branches[, 1] == rev(ancest)[k]), 1]
        k <- 2
        while (str_detect(ancestral_state, pattern = 'n|X') == TRUE & k <= length(ancest) - 1) {
          current_ancestor <- rev(ancest)[k]
          idx <- which(branches[, 1] == current_ancestor)
          if (length(idx) == 0) {
            cat("⚠️ No match for ancestor node in branches[,1] at SNP", i, "\n")
            break
          }
          ancestral_state <- branches_snp[idx, 1]
          k <- k + 1
        }
        if (str_detect(ancestral_state, pattern = 'n|X') == TRUE) {
          ancestral_state <- branches_snp[which(branches[, 1] == mrca_node), 1][1]
        }
      }
      
      branches_snp[str_detect(branches_snp, pattern = 'n|X')] = ancestral_state
      
      ## Converts SNP states into a binary format indicating differences from the ancestral state. 
      ## Calculates the association score (score3) for the SNP.
      branches_snp_bin = (branches_snp!=ancestral_state)
      
      tmp = which(branches_group[,1] == 1 & branches_group[,2] == 1)
      Px = as.numeric(branches_group[tmp,])
      ## Px = 1 → If the branch belongs to the phylogenetic group.
      ## Px = 0 → If the branch is outside the group.
      Sx = as.numeric(branches_snp_bin[tmp,])
      ## Sx = 1 → If the SNP is mutated at that position.
      ##Sx = 0 → If the SNP retains the ancestral state.
      if (length(Px) == 0) {
        scores[[j]][i] <- NA
        names_possibles_snps <- c(names_possibles_snps, paste0(ancestral_state, '|', snp_name, '|NA'))
        next
      }
      score3 <- (sum((1 - Px)*(1 - Sx), na.rm=TRUE) + sum(Px*Sx, na.rm=TRUE)) / length(Px)
      ## Px: Presence/absence of the phylogenetic group.
      ## Sx: Presence/absence of the SNP.
      ## score3: Measures how well the mutation segregates within the group 
      ## (higher values suggest stronger association).
      ## Entries where the score is 1 indicate a perfect association (mutation is fully linked with the group).
      
      scores[[j]][i] = score3
      
      t <- table(branches_snp)
      if (length(t) == 0 || all(names(t) == ancestral_state)) {
        names_possibles_snps <- c(names_possibles_snps, paste0(ancestral_state, '|', snp_name, '|NA'))
      } else {
        t <- t[names(t) != ancestral_state]
        derived_state <- names(t)[which.max(t)]
        names_possibles_snps <- c(names_possibles_snps, paste0(ancestral_state, '|', snp_name, '|', derived_state))
      }
      
    }
    scores[[j]] = scores[[j]]#- median(scores[[j]])
    names(scores[[j]]) = names_possibles_snps
    ## Each entry for a group has both a name and a value:
    ## A name that follows the pattern "X|Y|Z", which corresponds to:
      ## X: Ancestral state (e.g., 0).
      ## Y: SNP position index (e.g., 1, 2, 3, ...).
      ## Z: Derived state (present for some SNPs, missing for others).
    ## A value that represents a computed score for SNP association
    
    ## For example, in 1.06490532281936|15|1 - Value 1 
    ## The score is 1.06490532281936 (a high association score)
    ## 15 is the SNP position
    ## 1 is the derived state (different from the ancestral state 0)
    ## Value of 1 Suggests Linieage defining mutaiton
    ## As compared to 0.0642177183586905|136|0 - Value 1
    
  }
  return(scores)
  ## calculate association scores between SNPs and phylogenetic groups, considering evolutionary and temporal relationships.
  ## The scores quantify how well specific SNPs distinguish groups of nodes/tips in the tree.
}


read_clean_tree <- function(file) {
  tr <- ape::read.nexus(file)
  tr <- collapse.singles(ladderize(multi2di(tr, random = FALSE), right = FALSE))
  tr
}

```

**Explainer: For the association scores:**

`score3 <- (sum((1 - Px)*(1 - Sx), na.rm=TRUE) + sum(Px*Sx, na.rm=TRUE)) / length(Px)`

$$
\text{score3} = \frac{\sum (1 - P_x)(1 - S_x) + \sum P_x S_x}{\text{length}(P_x)}
$$

**First term: `sum((1 - Px) * (1 - Sx))`**

-   Counts cases where **neither** the group nor the SNP mutation is
    present

**Second term: `sum(Px * Sx)`**

-   Counts cases where **both** the group and the SNP mutation are
    present.

**Denominator: `length(Px)`**

-   Normalizes by the number of branches considered.

##### **What Does `score3` Represent?**

-   If **score = 1** → **Perfect association** (mutation occurs
    exclusively within the group).

-   If **score = 0** → No association (mutation occurs randomly outside
    the group).

-   Intermediate values (e.g., `0.5`) → Partial association (mutation is
    somewhat linked to the group).

### Debug

```{r debug area, eval = FALSE}


```

### Name code format

"year.decimal_location_clade_id/suffix"

For example: "1994.019_BKK_3II_00092/94"

-   Year + Decimal: 1994.019
-   Location: BKK \*\* Bangkok (BKK), Kamphaeng Phet. (KPP) and
    Ratchaburi (RTB)
-   Clade: 3II
-   ID?: 00092
-   Year Suffix: 94

#### Load data

Load the tree, in which all the tip name include: collection time,
location and Pango lineage


```{r load tree, eval=T}
tree_path  <- "1_Data/1_5_DENV/best_tree_d%d_Lin.nexus"

for (d in 1:4) {
  file <- sprintf(tree_path, d)
  tr   <- read_clean_tree(file)

  # Assign tree
  assign(sprintf("tree_DENV%d", d), tr)

  # Names & counts
  nm <- tr$tip.label
  assign(sprintf("names_seqs_%d", d), nm)
  assign(sprintf("n_seq_%d", d), length(nm))

  # Parse YEAR and CLADE 
  years  <- suppressWarnings(as.numeric(stringr::word(nm, 1, sep = stringr::fixed("_"))))
  clades <- stringr::word(nm, 3, sep = stringr::fixed("_"))

  assign(sprintf("times_seqs_%d", d), years)
  assign(sprintf("clades_seqs_%d", d), clades)
}

```

## Explainer: Compute the index dynamics

To compute the index of all nodes, use the function *compute.index*, you
will need different inputs:

1.  Data:
    -   the *timed_tree*
    -   *metadata* dataframe (see below the details of the
        *dataset_with_nodes*)
    -   *distance matrix*: can be computed from the timed tree with the
        function *dist.nodes.with.names*
2.  Information on the pathogen genome (these are pathogen specific):
    -   The genome length of the pathogen considered: *genome_length*
        (in bp).
    -   The mutation rate of the pathogen considered: *mutation_rate*
        (in bp/genome/year), an average is fine.
3.  Index parameters (these are both pathogen-specific and
    dataset-specific):
    -   The timescale: *timescale* (in years), this will be used to
        compute the bandwidth, see more details below.
    -   Window of time on which to search for samples in the population:
        *wind*, see more details below.

The function outputs a vector containing the index of each node
(internal and terminal).

**timescale**: The timescale determines the kernel which enables to
track lineage emergence dynamically, focusing on short distances between
nodes (containing information about recent population dynamics) rather
than long distances (containing information about past evolution). The
timescale is tailored to the specific pathogen studied and its choice
depends on the molecular signal, as well as the transmission rate. In
the study, we used timescales ranging from months (typical of RNA
viruses) to years (typical of bacteria). To determine a timescale
suitable for your dataset, we recommend thinking about the generation
time of the pathogen considered, its mutation rate, and the amount of
diversity already accumulated. For example, at the time of the analysis,
SARS-CoV-2 was a new pathogen, spreading quickly and accumulating
diversity at a rate of \~2 mutations per month. Therefore, a small
timescale of less than a year chosen (0.15 years). On the contrary,
*Mycobacterium tuberculosis* is an older and relatively slowly spreading
pathogen, which accumulates mutations at a rate of \~0.2 mutation per
year. A much larger timescale was then chosen (30 years), to reflect
this. Ultimately, the best timescale is one that maximises the
visualisation of population dynamics. We recommend trying different
values.

**wind**: The choice of *wind* will depend on the sampling intensity of
the dataset. It defines the window of time around each node on which to
search for samples in the population. Ultimately it smooths the index
dynamics. As a mean of example, for SARS-CoV-2, we set *wind* to 15
days, as the dataset was intensely sampled. But for *Bordetella
pertussis*, which is more sparsely sampled, we chose a *wind* of 1 year.
If *wind* is too large, then all the nodes are considered to be part of
the same time window. If *wind* is too small, then only the nodes in
direct proximity of the node of interest will be considered in the time
window, which can result in noisy index dynamics. We recommend choosing
a *wind* value that enables to span multiple sampling times, for example
if you have samples and nodes every week, you may choose a *wind* of
\~1-2 months. If you have samples and nodes every year, you may choose a
*wind* of \~2 years.

### Set the index parameters.

```{r index params, eval=T}
## Length genome 
genome_length = 11000 # reference  https://nextstrain.org/dengue/all/genome Could also just use 11kb or something highly specific
## Mutation rate 
mutation_rate = 7.6e-4
  #7.6e-4 # mutation rate #ref https://pmc.ncbi.nlm.nih.gov/articles/PMC9030598/
mutation_rate_sci <- formatC(mutation_rate, format = "e", digits = 2)
## Parameters for the index
timescale = 7 ## Timescale in years; raises index 
## Window of time on which to search for samples in the population
wind = 365 #days 
wind = wind/365 #Wind smooths lines


```

#### Compute pairwise distance matrix

Compute distance between each pair of sequences and internal nodes in
the tree

```{r pairwise distance, eval=T}

for (d in 1:4) {
  tree_var <- sprintf("tree_DENV%d", d)
  tree     <- get(tree_var, inherits = TRUE)
  gmat <- dist.nodes.with.names(tree)
  assign(sprintf("genetic_distance_mat_%d", d), gmat)
}
```

Get the time of each internal node

```{r internal node times , eval=T}

for (d in 1:4) {
  tree         <- get(sprintf("tree_DENV%d", d))
  gmat         <- get(sprintf("genetic_distance_mat_%d", d))
  n_seq        <- get(sprintf("n_seq_%d", d))
  names_seqs   <- get(sprintf("names_seqs_%d", d))
  times_seqs   <- get(sprintf("times_seqs_%d", d))

  # Root index: tips are 1..Ntip, so root is Ntip + 1 (ape convention)
  nroot <- length(tree$tip.label) + 1
  distance_to_root <- gmat[nroot, ]

  # Use a reference tip that exists in the distance vector names (robust if first entry isn't a tip)
  ref_tip <- intersect(names_seqs, names(distance_to_root))[1]

  root_height <- times_seqs[which(names_seqs == ref_tip)] - distance_to_root[ref_tip]

  # Internal nodes are indexed right after tips, sets height in time relative to tree root
  nodes_height <- root_height + distance_to_root[n_seq + (1:(n_seq - 1))] 

  assign(sprintf("nroot_%d", d), nroot)
  assign(sprintf("distance_to_root_%d", d), distance_to_root)
  assign(sprintf("root_height_%d", d), root_height)
  assign(sprintf("nodes_height_%d", d), nodes_height)
}
  


```

#### Preparation data tips and nodes and Compute index

Prepare the main dataframe, where the index and lineages of all nodes
(internal and terminal) are going to be stored.

```{r dataset with nodes, eval=T}

# Metadata dataframe
for (d in 1:4) {
  n_seq       <- get(sprintf("n_seq_%d", d))
  names_seqs  <- get(sprintf("names_seqs_%d", d))
  times_seqs  <- get(sprintf("times_seqs_%d", d))
  clades      <- get(sprintf("clades_seqs_%d", d))
  nodes_height <- get(sprintf("nodes_height_%d", d))
  
  tip_ids     <- seq_len(n_seq)  # 1, 2, ..., Ntip
  node_ids    <- n_seq + seq_len(n_seq - 1)    # Ntip+1, ..., Ntip+(Ntip-1)

  df <- data.frame(
    ID       = c(tip_ids, node_ids),
    name_seq = c(names_seqs, node_ids),  # tips keep names; nodes get numeric IDs
    time     = c(times_seqs, nodes_height),  # tip times + internal node heights
    is.node  = c(rep("no", n_seq), rep("yes", n_seq-1)),
    clade    = c(clades, rep(NA, n_seq-1)),
    stringsAsFactors = FALSE
  )

  # Compute index of every tip and node
   idx <- compute.index(
    time_distance_mat = get(sprintf("genetic_distance_mat_%d", d)),
    timed_tree        = get(sprintf("tree_DENV%d", d)),
    time_window       = wind,
    metadata          = df,
    mutation_rate     = mutation_rate,
    timescale         = timescale,
    genome_length     = genome_length
  )
  
  df$index <- idx
  
  
  assign(sprintf("dataset_with_nodes_%d", d), df)
}


```


Several of the DENV 3 strains are missing clade asignments, these were assigned by eye. 
```{r fixing D3XX}
#Fixing 3XX

three_xx_tips <- dataset_with_nodes_3 %>%
  filter(grepl("3XX", clade)) %>%
  pull(name_seq)  # Extract sequence names with 3XX clade


three_iii_reassign <- c(
  "2007.063_H07_3XX_006.1",
  "2007.563_R07_3XX_266.1",
  "2008.73_P08_3XX_268.1",
  "2009.914_R09_3XX_505.1",
  "2010.647_P10_3XX_011.1"
)

dataset_with_nodes_3 <- dataset_with_nodes_3 %>%
  mutate(clade = case_when(
    name_seq %in% three_iii_reassign ~ "3III",  # Assign 3III to specific sequences
    grepl("3XX", clade) ~ "3II",  # Default: Assign all other 3XX to 3II
    TRUE ~ clade  # Keep existing assignments
  ))

# Sanity check
three_xx_tips_post <- dataset_with_nodes_3 %>%
  filter(grepl("3XX", clade)) %>%
  pull(name_seq)  # Extract sequence names with 3XX clade




```

####Hunting for node clades to color: ie: inferring internal node clades
based on descendant tips

```{r clade inference, eval = TRUE}
# Dynamically inferring the clades
for (d in 1:4) {
  
  tree <- get(paste0("tree_DENV", d))
  dataset_with_nodes <- get(paste0("dataset_with_nodes_", d))
  
  # Total number of tips
  ntips <- length(tree$tip.label)
  
  # Create inferred clade column
  dataset_with_nodes$clade_inferred <- dataset_with_nodes$clade
  
  # Traverse internal nodes
  for (node in (ntips + 1):(ntips + tree$Nnode)) {
    
    desc_tips <- unlist(Descendants(tree, node, type = "tips"))
    clade_vals <- dataset_with_nodes$clade[desc_tips]
    
    if (length(unique(clade_vals[!is.na(clade_vals)])) == 1) {
      dataset_with_nodes$clade_inferred[node] <- unique(clade_vals[!is.na(clade_vals)])
    } else {
      dataset_with_nodes$clade_inferred[node] <- NA
    }
  }
  
  # Save the updated dataset back into the environment
  assign(paste0("dataset_with_nodes_", d), dataset_with_nodes)
}
```

#### Plot tree & index below, with colors from clades

First, generate the color key, based on the Nextstrain clade of each
sequence.

```{r old clade plots, eval=T}

for (d in 1:4) {
  
  tree <- get(paste0("tree_DENV", d))
  dataset_with_nodes <- get(paste0("dataset_with_nodes_", d))
  
  colors_of_clades = met.brewer(name="Cross", n=length(levels(as.factor(dataset_with_nodes$clade_inferred))),
                            type="continuous")
  
  
  dataset_with_nodes$clade_color = as.factor(dataset_with_nodes$clade_inferred)
  clade_labels = levels(dataset_with_nodes$clade_color)
  levels(dataset_with_nodes$clade_color) = colors_of_clades
  dataset_with_nodes$clade_color = as.character(dataset_with_nodes$clade_color)
  dataset_with_nodes$clade_color[is.na(dataset_with_nodes$clade_color)] <- "grey"
  
  assign(paste0("clade_labels_", d), clade_labels)
  assign(paste0("colors_clade_", d), colors_of_clades)
  assign(paste0("dataset_with_nodes_", d), dataset_with_nodes)
}


```

# Figure 1 Plot tree & index below, with colors from clades



```{r clade and index plotting funcitons}

old_tree_plot <- function(
    dengue = 1:4,
    min_year = NULL,
    max_year = 2015,
    output_dir = "3_Output_Figures/3_5_DENV/3_5_1_Figure 1",
    filename = "Classical_Clades_plot_DENV_%d.svg",
    width = 8,
    height = 5){

  for (d in dengue) {
    
  # Retrieve relevant variables
    tree <- get(paste0('tree_DENV', d)) 
    dataset_with_nodes <- get(paste0('dataset_with_nodes_', d))
    root_height <- get(paste0('root_height_', d))
    clade_labels <- get(paste0('clade_labels_', d))
    colors_clade <- get(paste0('colors_clade_', d))
    
    #need new dummy variable to not overight in the loop, is rounded lowest year if NULL and is user specified in else occasion
    min_year_d <- if (is.null(min_year)) floor(min(dataset_with_nodes$time, na.rm = TRUE)) - 1 else min_year 
    
  
    
     # 1 panel clades plot setup
    file_path <- file.path(output_dir, sprintf(filename, d))
    svglite(filename = file_path, width = width, height = height)
    par(mfrow = c(1, 1), oma = c(0, 0, 0, 0), mar = c(4, 4, 0, 0))
    
    ## Tree plot
    plot(tree, show.tip.label = FALSE, 
         edge.color = 'grey', edge.width = 0.25,
         x.lim = c(min_year_d, max_year) - root_height)
    tiplabels(pch = 16, col = dataset_with_nodes$clade_color, cex = 0.3)
    axisPhylo_NL(side = 1, root.time = root_height, backward = FALSE,
                 at_axis = seq(min_year_d, max_year, 0.5) - root_height,
                 lab_axis = seq(min_year_d, max_year, 0.5), lwd = 0.5)
    
    title(main = paste0(
      "DENV", d, "\n",
      "Genome Length: ", genome_length, " bp\n",
      "Mutation Rate: ", mutation_rate_sci, "\n",
      "Timescale: ", timescale, " years \n",
      "Window: ", round(wind, 2), " years"
    ), line = -3.9, adj = 0, cex.main = 0.8, col.main = "grey")
     
    # Save the plot
    dev.off()
    
  }
}

old_index_plot <- function(
    dengue = 1:4,
    min_year = NULL,
    max_year = 2015,
    output_dir = "3_Output_Figures/3_5_DENV/3_5_1_Figure 1",
    filename = "Classical_Index_plot_DENV_%d.svg",
    width = 8,
    height = 5){
  
  for (d in dengue) {
    # Retrieve relevant variables
    tree <- get(paste0('tree_DENV', d)) 
    dataset_with_nodes <- get(paste0('dataset_with_nodes_', d))
    root_height <- get(paste0('root_height_', d))
    clade_labels <- get(paste0('clade_labels_', d))
    colors_clade <- get(paste0('colors_clade_', d))
    min_year_d <- if (is.null(min_year)) floor(min(dataset_with_nodes$time, na.rm = TRUE)) - 1 else min_year
    
    
     # 1 panel index plot setup
    file_path <- file.path(output_dir, sprintf(filename, d))
    svglite(filename = file_path, width = width, height = height)
    par(mfrow = c(1, 1), oma = c(0, 0, 0, 0), mar = c(4, 4, 0, 0))
    ## Index plot
    plot(dataset_with_nodes$time, 
         dataset_with_nodes$index, 
         col = adjustcolor(dataset_with_nodes$clade_color, alpha.f = 1),
         bty = 'n', xlim = c(min_year_d, max_year), cex = 0.4,
         pch = 16, ylim = c(0, 1), 
         ylab = 'Index', xlab = 'Time (years)', xaxt = 'n', yaxt = 'n')
    
    title(main = paste0(
      "DENV", d, " \n",
      "Genome Length: ", genome_length, " bp\n",
      "Mutation Rate: ", mutation_rate_sci, "\n",
      "Timescale: ", timescale, " years \n",
      "Window: ", round(wind, 2), " years"
    ), line = -3.9, adj = 0, cex.main = 0.8, col.main = "grey")
    
    axis(2, las = 2, lwd = 0.5)
    axis(1, lwd = 0.5)
    
    # Color key
    legend('topright', 
           legend = clade_labels,
           fill = colors_clade, border = colors_clade,
           cex = 0.5, bty = 'n', ncol = 5)
    # Save the plot
    dev.off()
    
  }
}


old_clade_and_index_panel_plot <- function(
    dengue = 1:4,
    min_year = NULL,
    max_year = 2015,
    output_dir = "3_Output_Figures/3_5_DENV/3_5_1_Figure 1",
    filename = "Classical_Clades_and_Index_panel_plot_DENV_%d.svg",
    width = 8,
    height = 10){
  
  for (d in dengue) {
    # Retrieve relevant variables
    tree <- get(paste0('tree_DENV', d)) 
    dataset_with_nodes <- get(paste0('dataset_with_nodes_', d))
    root_height <- get(paste0('root_height_', d))
    clade_labels <- get(paste0('clade_labels_', d))
    colors_clade <- get(paste0('colors_clade_', d))
    min_year_d <- if (is.null(min_year)) floor(min(dataset_with_nodes$time, na.rm = TRUE)) - 1 else min_year
    
     # 1 panel index plot setup
    file_path <- file.path(output_dir, sprintf(filename, d))
    svglite(filename = file_path, width = width, height = height)

  
  # Two-panel plot setup
  par(mfrow = c(2, 1), oma = c(0, 0, 0, 0), mar = c(4, 4, 0, 0))
  
  ## Tree plot
  plot(tree, show.tip.label = FALSE, 
       edge.color = 'grey', edge.width = 0.25,
       x.lim = c(min_year_d, max_year) - root_height)
  tiplabels(pch = 16, col = dataset_with_nodes$clade_color, cex = 0.3)
  axisPhylo_NL(side = 1, root.time = root_height, backward = FALSE,
               at_axis = seq(min_year_d, max_year, 0.5) - root_height,
               lab_axis = seq(min_year_d, max_year, 0.5), lwd = 0.5)
  
  title(main = paste0(
    "DENV", d, "\n",
    "Genome Length: ", genome_length, " bp\n",
    "Mutation Rate: ", mutation_rate_sci, "\n",
    "Timescale: ", timescale, " years \n",
    "Window: ", round(wind, 2), " years"
  ), line = -3.9, adj = 0, cex.main = 0.8, col.main = "grey")
  
  ## Index plot
  plot(dataset_with_nodes$time, 
       dataset_with_nodes$index, 
       col = adjustcolor(dataset_with_nodes$clade_color, alpha.f = 1),
       bty = 'n', xlim = c(min_year_d, max_year), cex = 0.4,
       pch = 16, ylim = c(0, 1), 
       ylab = 'Index', xlab = 'Time (years)', xaxt = 'n', yaxt = 'n')
  
  title(main = paste0(
    "DENV", d, " \n",
    "Genome Length: ", genome_length, " bp\n",
    "Mutation Rate: ", mutation_rate_sci, "\n",
    "Timescale: ", timescale, " years \n",
    "Window: ", round(wind, 2), " years"
  ), line = -3.9, adj = 0, cex.main = 0.8, col.main = "grey")
  
  axis(2, las = 2, lwd = 0.5)
  axis(1, lwd = 0.5)
  
  # Color key
  legend('topright', 
         legend = clade_labels,
         fill = colors_clade, border = colors_clade,
         cex = 0.5, bty = 'n', ncol = 5)
  
  
  # Save the plot
  
  dev.off()
  }
}




```

### TODO: Maxtime or set all to 1965?
```{r clade and index plots old clades}

old_tree_plot(
    dengue = 1:4,
    min_year = NULL,  #set to the lowest year in the tree
    max_year = 2015,
    output_dir = "3_Output_Figures/3_5_DENV/3_5_1_Figure 1",
    filename = "Classical_Clades_plot_alltime_DENV_%d.svg",
    width = 8,
    height = 5
)

old_index_plot(
    dengue = 1:4,
    min_year = NULL,
    max_year = 2015,
    output_dir = "3_Output_Figures/3_5_DENV/3_5_1_Figure 1",
    filename = "Classical_Index_plot_alltime_DENV_%d.svg",
    width = 8,
    height = 5)

 old_clade_and_index_panel_plot(
    dengue = 1:4,
    min_year = NULL,
    max_year = 2015,
    output_dir = "3_Output_Figures/3_5_DENV/3_5_1_Figure 1",
    filename = "Classical_Clades_and_Index_panel_plot_alltime_DENV_%d.svg",
    width = 8,
    height = 10)


#plotting since 1965 
old_tree_plot(
    dengue = 1:4,
    min_year = 1965,
    max_year = 2015,
    output_dir = "3_Output_Figures/3_5_DENV/3_5_1_Figure 1",
    filename = "Classical_Clades_plot_1965start_DENV_%d.svg",
    width = 8,
    height = 5
)

old_index_plot(
    dengue = 1:4,
    min_year = 1965,
    max_year = 2015,
    output_dir = "3_Output_Figures/3_5_DENV/3_5_1_Figure 1",
    filename = "Classical_Index_plot_1965start_DENV_%d.svg",
    width = 8,
    height = 5)

old_clade_and_index_panel_plot(
    dengue = 1:4,
    min_year = 1965,
    max_year = 2015,
    output_dir = "3_Output_Figures/3_5_DENV/3_5_1_Figure 1",
    filename = "Classical_Clades_and_Index_panel_plot_1965start_DENV_%d.svg",
    width = 8,
    height = 10)



```


### Suplement Wind and Timescale robustness: Subplot comparison

```{r subplot_comparison, eval=FALSE}

for (d in 1:4) {
  
# Retrieve relevant variables
  dataset_with_nodes <- get(paste0('dataset_with_nodes_', d))
  
# Define global plot parameters
# min_year <- round(min(dataset_with_nodes$time) - 1)
min_year <- 1965
max_year <- 2015
  
# Define wind values (x-axis )
wind_values <- seq(62, 730, length.out = 5) / 365  # Convert days to years

# Define timescale values (y-axis)
timescale_values <- seq(2, 20, length.out = 5)  # Years

par(mfrow = c(5, 5), oma = c(2, 2, 2, 2), mar = c(0.5, 0.5, 0.7, 0.5), mgp = c(2, 0.5, 0))  # Adjust margins

for (t in timescale_values) {
  for (w in wind_values) {
    # Update parameters
    timescale <- t
    wind <- w
    
    # Compute the index for the given wind and timescale
    
    dataset_with_nodes$index <- compute.index(
    time_distance_mat = get(sprintf("genetic_distance_mat_%d", d)),
    timed_tree        = get(sprintf("tree_DENV%d", d)),
    time_window       = wind,
    metadata          = dataset_with_nodes,
    mutation_rate     = mutation_rate,
    timescale         = timescale,
    genome_length     = genome_length
  )
    
    # Plot the index
    plot(dataset_with_nodes$time, 
         dataset_with_nodes$index, 
         col = adjustcolor(dataset_with_nodes$clade_color, alpha.f = 1),
         bty = 'n', xlim = c(min_year, max_year), cex = 0.4,
         pch = 16, ylim = c(0, 1), 
         xaxt = 'n', yaxt = 'n',
         main = paste0("TS: ", timescale, " | W: ", round(wind, 2))
    )
    # Add axes for the leftmost and bottom subplots only
    if (w == wind_values[1]) axis(2, las = 2, lwd = 0.5)  # Left y-axis
    if (t == timescale_values[5]) axis(1, lwd = 0.5)      # Bottom x-axis
  }
}

mtext("Window (years)", side = 1, outer = TRUE, line = 1.2, cex = 0.6)
mtext("Timescale (years)", side = 2, outer = TRUE, line = 1.2, cex= 0.6)
mtext(paste0(
  "Index Plots for Various Timescales and Windows",
  "Genome Length: ", genome_length,
  "Mutation Rate: ", mutation_rate_sci), side = 3, outer = TRUE, line = 1, cex = 0.6, font = 1)
  

# After generating the 5x5 subplot grid
file_name_subplot <- glue("3_Output_Figures/3_5_DENV/3_5_99_Supplement/index_plots_DENV{d}_{mutation_rate_sci}_TS{min(timescale_values)}-{max(timescale_values)}_W{round(min(wind_values),2)}-{round(max(wind_values),2)}.pdf")

# Save the current plot to a PDF
dev.copy(pdf, file = file_name_subplot, width = 10, height = 10)  # Adjust dimensions as needed
dev.off()  # Close the PDF device

}

```

## Find DENGUE clades based on index dynamics

#### Run the lineage detection algorithm on DENGUE data

To run the lineage detection algorithm, use the function
*find.groups.by.index.dynamics*, you will need different inputs:

1.  Data: *timed_tree* (must be the same as the one used in
    compute.index) and *metadata* (*dataset_with_nodes*, with the index
    values of each node)
2.  Lineage detection parameters:
    -   *min_descendants_per_tested_node*: to start the analysis, start
        from nodes that have this minimum number of sequences
    -   *min_group_size*: minimum group size, when creating a new
        potential split
    -   *node_support*: numeric value of support of each node (e.g.
        mutations on the branch leading to the node, or bootstrap
        support)
    -   *threshold_node_support*: threshold on the node support for the
        nodes to be considered in the detection algorithm
    -   *weight_by_time*: size of the window of time on which to compute
        the weights (NULL or numeric, in years)
    -   *weighting_transformation*: type of weighting to use (NULL,
        inv_freq, inv_sqrt, or inv_log)
    -   *max_groups_found*: maximum number of groups to find (Integer)
3.  Technical parameters: they do not necessarily need to be updated
    (see the function documentation for details): *p_value_smooth*,
    *stepwise_deviance_explained_threshold*, *stepwise_AIC_threshold*,
    *k_smooth*, *parallelize_code*, *number_cores*, *plot_screening*,
    *keep_track* and *log_y*.
    
#### More notes on Parameters for the detection:

Threshold node support: Max lineages is the same as number of lineages. Idea is restrict search space to certain level. Trying to restrict space to only well supported nodes. Each node has support value: 1. bootstrap/posterior (similar to Tb tree) support comes from data Creates sparsing issues in viruses that don't change much. 2. Proxy of how many mutations ahead of branch. If two nodes are separated by a short amount of time then the hypothesis that both nodes are distinct is less likely. Uses branch lenght between nodes. Only permit nodes that are at least 1 mutation apart. Length must be how long it takes for 1 mutation.


Using time_window_initial + increment: sometimes helpful if wanting to look at newer lineages. But elbow plots are harder to interpret as depeneding on time params. Not recommended.  

min_descendants_per_tested_node: only considering nodes with n tip and node descendents. A filter or partition of the tree. 

Index dynamics are made of many splines (each with a k), p values for the spline need to better explain dynamics. k is the number of knots 

Weight by time: Seqiences are a time series, this compensates for some times where there are no sequences. Thus the model is trying to overexplain the densely time sampled variants. Penalty for higher samples in higher sampled times. Weight by time is the bucket/slice of time in years. 
weighting_transformation: how the penalty is applied. Can try all of these but can't really compare between compare since AIC and BIC are poor measure of GAM comparison.  

The function outputs multiple elements in a list:

-   potential_splits: vector of the nodes included in most complex model
    tested
-   best_dev_explained: vector of the deviance explained by the best
    models for each number of groups
-   first_dev: the null deviance of the initial model (when no lineage
    is present)
-   best_AIC : vector of the AIC of the best models for each number of
    groups
-   best_BIC: vector of the BIC of the best models for each number of
    groups
-   best_summary: list of the summaries of the best models for each
    number of groups
-   best_mod: list of the best models for each number of groups
-   best_groups: list of the groups used in the best models for each
    number of groups
-   best_nodes_names: list of the nodes included in the best models for
    each number of groups

Typically, one chooses a value of *max_groups_found* greater than the
expected number of lineages. The algorithm then runs until it finds all
those groups, or until it cannot find any significant split anymore. The
user can then check the deviance explained by all the models with
increasing complexity and choose an adequate number of groups.

Once the split nodes have been defined, the user can then extract the
group ID for each node using the function *merge.groups*. One can choose
to refine these groups if needed, by setting a minimum number of nodes
per group (*group_count_threshold*) or a minimum frequency of each group
(*group_freq_threshold*).




Parameters for the detection:

```{r model params, eval =TRUE}
# Time it takes to have a single denv mut. Nodes must be at least this far apart
threshold_node_support = 1/(mutation_rate*genome_length) 
time_window_initial = 2030 # Consider the whole tree, don't increment the window
time_window_increment = 100 # Not used since whole tree is being considered
min_descendants_per_tested_node = 10
min_group_size = 10
p_value_smooth = 0.05 # default is 0.05 but often need to increase to allow more smooth
stepwise_deviance_explained_threshold = 0
stepwise_AIC_threshold = 0 # AIC is poor comparison of GAMs, kept at inactive
# weight_by_time: Time (in years) to consider density penalty buckets, 
# if set too low can poorly fit early data
weight_by_time = 10
weighting_transformation = c('inv_sqrt')
k_smooth = -1 #Allows model to pick, usually model sets at 10 
parallelize_code = T
number_cores = 8
plot_screening = T
plot_gif = T
gif_output_dir = "3_Output_Figures/3_5_DENV/3_5_99_Supplement"
max_groups_found_default =  30
keep_track = T

# max_groups_found_3 = 6
# max_groups_found_2 = 3
# max_groups_found_1 = 8
# max_groups_found_4 = 2







```

Go back and set the max_groups_found to a reasonable value after doing
the deviance analysis

DENV3 Run the detection function (this steps takes approximately \<10
min on 2 cores):

```{r potential splits 3, eval=FALSE, results = 'hide', warning=FALSE, message=FALSE}
start_time = Sys.time()
potential_splits_reasonable_3 = find.groups.by.index.dynamics(timed_tree = tree_DENV3,
                                                 metadata = dataset_with_nodes_3,
                                                 node_support = tree_DENV3$edge.length[match((n_seq_3+1):(2*n_seq_3-1),
                                                                                            tree_DENV3$edge[,2])],
                                                 threshold_node_support = threshold_node_support,
                                                 time_window_initial = time_window_initial,
                                                 time_window_increment = time_window_increment,
                                                 min_descendants_per_tested_node = min_descendants_per_tested_node,
                                                 min_group_size = min_group_size,
                                                 p_value_smooth = p_value_smooth,
                                                 stepwise_deviance_explained_threshold = stepwise_deviance_explained_threshold,
                                                 stepwise_AIC_threshold = stepwise_AIC_threshold,
                                                 weight_by_time = weight_by_time,
                                                 weighting_transformation = weighting_transformation,
                                                 k_smooth = k_smooth,
                                                 parallelize_code = parallelize_code,
                                                 number_cores = number_cores, 
                                                 plot_screening = plot_screening,
                                                 plot_gif = plot_gif,
                                                 gif_output_dir = gif_output_dir,
                                                 gif_filename = "reasonable_params_maxgroups_D3.gif",
                                                 max_groups_found = max_groups_found_default, 
                                                 # max_groups_found = max_groups_found_3, 
                                                 keep_track = keep_track
                                                 )

end_time = Sys.time()
print(end_time - start_time)


attr(potential_splits_reasonable_3, "params") <- list(
  head_note = "Run to with reasonable params on D3 for max lineages",
  node_support = tree_DENV3$edge.length[match((n_seq_3+1):(2*n_seq_3-1),
                                              tree_DENV3$edge[,2])],
  threshold_node_support = threshold_node_support,
  time_window_initial = time_window_initial,
  time_window_increment = time_window_increment,
  min_descendants_per_tested_node = min_descendants_per_tested_node,
  min_group_size = min_group_size,
  p_value_smooth = p_value_smooth,
  stepwise_deviance_explained_threshold = stepwise_deviance_explained_threshold,
  stepwise_AIC_threshold = stepwise_AIC_threshold,
  weight_by_time = weight_by_time,
  weighting_transformation = weighting_transformation,
  k_smooth = k_smooth,
  parallelize_code = parallelize_code,
  number_cores = number_cores, 
  plot_screening = plot_screening,
  plot_gif = plot_gif,
  gif_output_dir = gif_output_dir,
  gif_filename = "reasonable_params_maxgroups_D3.gif",
  max_groups_found = max_groups_found_default, 
  # max_groups_found = max_groups_found_3, 
  keep_track = keep_track
)
```

DENV2 Run the detection function (this steps takes approximately \<10
min on 2 cores):

```{r potential splits 2, eval=FALSE, results = 'hide', warning=FALSE, message=FALSE}
start_time = Sys.time()
potential_splits_reasonable_2 = find.groups.by.index.dynamics(timed_tree = tree_DENV2,
                                                 metadata = dataset_with_nodes_2,
                                                 node_support = tree_DENV2$edge.length[match((n_seq_2+1):(2*n_seq_2-1),
                                                                                            tree_DENV2$edge[,2])],
                                                 threshold_node_support = threshold_node_support,
                                                 time_window_initial = time_window_initial,
                                                 time_window_increment = time_window_increment,
                                                 min_descendants_per_tested_node = min_descendants_per_tested_node,
                                                 min_group_size = min_group_size,
                                                 p_value_smooth = p_value_smooth,
                                                 stepwise_deviance_explained_threshold = stepwise_deviance_explained_threshold,
                                                 stepwise_AIC_threshold = stepwise_AIC_threshold,
                                                 weight_by_time = weight_by_time,
                                                 weighting_transformation = weighting_transformation,
                                                 k_smooth = k_smooth,
                                                 parallelize_code = parallelize_code,
                                                 number_cores = number_cores, 
                                                 plot_screening = plot_screening,
                                                 plot_gif = plot_gif,
                                                 gif_output_dir = gif_output_dir,
                                                 gif_filename = "reasonable_params_maxgroups_D2.gif",
                                                 max_groups_found =  max_groups_found_default, 
                                                 keep_track = keep_track)
end_time = Sys.time()
print(end_time - start_time)


attr(potential_splits_reasonable_2, "params") <- list(
  head_note = "Run to with reasonable params on D2 for max lineages",
  node_support = tree_DENV2$edge.length[match((n_seq_2+1):(2*n_seq_2-1),
                                              tree_DENV2$edge[,2])],
  threshold_node_support = threshold_node_support,
  time_window_initial = time_window_initial,
  time_window_increment = time_window_increment,
  min_descendants_per_tested_node = min_descendants_per_tested_node,
  min_group_size = min_group_size,
  p_value_smooth = p_value_smooth,
  stepwise_deviance_explained_threshold = stepwise_deviance_explained_threshold,
  stepwise_AIC_threshold = stepwise_AIC_threshold,
  weight_by_time = weight_by_time,
  weighting_transformation = weighting_transformation,
  k_smooth = k_smooth,
  parallelize_code = parallelize_code,
  number_cores = number_cores, 
  plot_screening = plot_screening,
  plot_gif = plot_gif,
  gif_output_dir = gif_output_dir,
  gif_filename = "reasonable_params_maxgroups_D2.gif",
  max_groups_found =  max_groups_found_default, 
  keep_track = keep_track
)
```

DENV1 Run the detection function (this steps takes approximately \<10
min on 2 cores):

```{r potential splits 1, eval=FALSE, results = 'hide', warning=FALSE, message=FALSE}
start_time = Sys.time()
potential_splits_reasonable_1 = find.groups.by.index.dynamics(timed_tree = tree_DENV1,
                                                 metadata = dataset_with_nodes_1,
                                                 node_support = tree_DENV1$edge.length[match((n_seq_1+1):(2*n_seq_1-1),
                                                                                            tree_DENV1$edge[,2])],
                                                 threshold_node_support = threshold_node_support,
                                                 time_window_initial = time_window_initial,
                                                 time_window_increment = time_window_increment,
                                                 min_descendants_per_tested_node = min_descendants_per_tested_node,
                                                 min_group_size = min_group_size,
                                                 p_value_smooth = p_value_smooth,
                                                 stepwise_deviance_explained_threshold = stepwise_deviance_explained_threshold,
                                                 stepwise_AIC_threshold = stepwise_AIC_threshold,
                                                 weight_by_time = weight_by_time,
                                                 weighting_transformation = weighting_transformation,
                                                 k_smooth = k_smooth,
                                                 parallelize_code = parallelize_code,
                                                 number_cores = number_cores, 
                                                 plot_screening = plot_screening,
                                                 plot_gif = plot_gif,
                                                 gif_output_dir = gif_output_dir,
                                                 gif_filename = "reasonable_params_maxgroups_D1.gif",
                                                 max_groups_found = max_groups_found_default, 
                                                 keep_track = keep_track)
end_time = Sys.time()
print(end_time - start_time)



attr(potential_splits_reasonable_1, "params") <- list(
  head_note = "Run to with reasonable params on D1 for max lineages",
  node_support = tree_DENV1$edge.length[match((n_seq_1+1):(2*n_seq_1-1),
                                              tree_DENV1$edge[,2])],
  threshold_node_support = threshold_node_support,
  time_window_initial = time_window_initial,
  time_window_increment = time_window_increment,
  min_descendants_per_tested_node = min_descendants_per_tested_node,
  min_group_size = min_group_size,
  p_value_smooth = p_value_smooth,
  stepwise_deviance_explained_threshold = stepwise_deviance_explained_threshold,
  stepwise_AIC_threshold = stepwise_AIC_threshold,
  weight_by_time = weight_by_time,
  weighting_transformation = weighting_transformation,
  k_smooth = k_smooth,
  parallelize_code = parallelize_code,
  number_cores = number_cores, 
  plot_screening = plot_screening,
  plot_gif = plot_gif,
  gif_output_dir = gif_output_dir,
  gif_filename = "reasonable_params_maxgroups_D1.gif",
  max_groups_found = max_groups_found_default, 
  keep_track = keep_track
)
```

DENV4 Run the detection function (this steps takes approximately \<10
min on 2 cores):

```{r potential splits 4, eval=FALSE, results = 'hide', warning=FALSE, message=FALSE}
start_time = Sys.time()
potential_splits_reasonable_4 = find.groups.by.index.dynamics(timed_tree = tree_DENV4,
                                                 metadata = dataset_with_nodes_4,
                                                 node_support = tree_DENV4$edge.length[match((n_seq_4+1):(2*n_seq_4-1),
                                                                                            tree_DENV4$edge[,2])],
                                                 threshold_node_support = threshold_node_support,
                                                 time_window_initial = time_window_initial,
                                                 time_window_increment = time_window_increment,
                                                 min_descendants_per_tested_node = min_descendants_per_tested_node,
                                                 min_group_size = min_group_size,
                                                 p_value_smooth = p_value_smooth,
                                                 stepwise_deviance_explained_threshold = stepwise_deviance_explained_threshold,
                                                 stepwise_AIC_threshold = stepwise_AIC_threshold,
                                                 weight_by_time = weight_by_time,
                                                 weighting_transformation = weighting_transformation,
                                                 k_smooth = k_smooth,
                                                 parallelize_code = parallelize_code,
                                                 number_cores = number_cores, 
                                                 plot_screening = plot_screening,
                                                 plot_gif = plot_gif,
                                                 gif_output_dir = gif_output_dir,
                                                 gif_filename = "reasonable_params_maxgroups_D4.gif",
                                                 max_groups_found =  max_groups_found_default,
                                                 keep_track = keep_track)
end_time = Sys.time()
print(end_time - start_time)

attr(potential_splits_reasonable_4, "params") <- list(
  head_note = "Run to with reasonable params on D4 for max lineages",
  node_support = tree_DENV4$edge.length[match((n_seq_4+1):(2*n_seq_4-1),
                                              tree_DENV4$edge[,2])],
  threshold_node_support = threshold_node_support,
  time_window_initial = time_window_initial,
  time_window_increment = time_window_increment,
  min_descendants_per_tested_node = min_descendants_per_tested_node,
  min_group_size = min_group_size,
  p_value_smooth = p_value_smooth,
  stepwise_deviance_explained_threshold = stepwise_deviance_explained_threshold,
  stepwise_AIC_threshold = stepwise_AIC_threshold,
  weight_by_time = weight_by_time,
  weighting_transformation = weighting_transformation,
  k_smooth = k_smooth,
  parallelize_code = parallelize_code,
  number_cores = number_cores, 
  plot_screening = plot_screening,
  plot_gif = plot_gif,
  gif_output_dir = gif_output_dir,
  gif_filename = "reasonable_params_maxgroups_D4.gif",
  max_groups_found =  max_groups_found_default,
  keep_track = keep_track
)

attr(potential_splits_reasonable_4, "params")
```

Saving the Potential Splits and treedata
```{r save splits, }


for (d in 1:4){
  potential_splits <- get(sprintf("potential_splits_reasonable_%d", d))
  splits_file_name <- glue("4_Output_Data/Max_groups_potential_splits_reasonable_DENV{d}_timescale{timescale}_wind{wind}_{st}.rds")
  write_rds(potential_splits, splits_file_name)

  tree <- get(paste0('tree_DENV', d)) 
  tree_file_name <- glue("4_Output_Data/tree_DENV{d}_{st}.rds")
  write_rds(tree, tree_file_name)
  
  dataset_with_nodes <- get(paste0('dataset_with_nodes_', d))
  ds_file_name <- glue("4_Output_Data/dataset_with_nodes_index_DENV{d}_{st}.rds")
  write_rds(dataset_with_nodes, ds_file_name)
  
}

```



Instead, you may wish to load or write the results:

```{r, eval = T}



# potential_splits_3 = readRDS("4_Output_Data/potential_splits_DENV3_timescale7_wind1_2025-06-20_0018.rds")
# potential_splits_2 = readRDS("4_Output_Data/potential_splits_DENV2_timescale7_wind1_2025-06-20_0018.rds")
# potential_splits_1 = readRDS("4_Output_Data/potential_splits_DENV1_timescale7_wind1_2025-06-20_0018.rds")
# potential_splits_4 = readRDS("4_Output_Data/potential_splits_DENV4_timescale7_wind1_2025-06-20_0018.rds")
# 


potential_splits_1 <- readRDS("4_Output_Data/Max_groups_potential_splits_reasonable_DENV1_timescale7_wind1_2025-10-20_1143.rds")
potential_splits_2 <- readRDS("4_Output_Data/Max_groups_potential_splits_reasonable_DENV2_timescale7_wind1_2025-10-20_1143.rds")
potential_splits_3 <- readRDS("4_Output_Data/Max_groups_potential_splits_reasonable_DENV3_timescale7_wind1_2025-10-20_1143.rds")
potential_splits_4 <- readRDS("4_Output_Data/Max_groups_potential_splits_reasonable_DENV4_timescale7_wind1_2025-10-20_1143.rds")

dataset_with_nodes_1 <- readRDS("4_Output_Data/dataset_with_nodes_index_DENV1_2025-10-20_1143.rds")
dataset_with_nodes_2 <- readRDS("4_Output_Data/dataset_with_nodes_index_DENV2_2025-10-20_1143.rds")
dataset_with_nodes_3 <- readRDS("4_Output_Data/dataset_with_nodes_index_DENV3_2025-10-20_1143.rds")
dataset_with_nodes_4 <- readRDS("4_Output_Data/dataset_with_nodes_index_DENV4_2025-10-20_1143.rds")

tree_1 <- readRDS("4_Output_Data/tree_DENV1_2025-10-20_1143.rds")
tree_2 <- readRDS("4_Output_Data/tree_DENV2_2025-10-20_1143.rds")
tree_3 <- readRDS("4_Output_Data/tree_DENV3_2025-10-20_1143.rds")
tree_4 <- readRDS("4_Output_Data/tree_DENV4_2025-10-20_1143.rds")


##Write
# file_name_3 <- glue("4_Output_Data/potential_splits_DENV3_timescale{timescale}_wind{wind}_{st}.rds")
# write_rds(potential_splits_3, file_name_3)
# 
# file_name_2 <- glue("4_Output_Data/potential_splits_DENV2_timescale{timescale}_wind{wind}_{st}.rds")
# write_rds(potential_splits_2, file_name_2)
# 
# file_name_1 <- glue("4_Output_Data/potential_splits_DENV1_timescale{timescale}_wind{wind}_{st}.rds")
# write_rds(potential_splits_1, file_name_1)
# 
# file_name_4 <- glue("4_Output_Data/potential_splits_DENV4_timescale{timescale}_wind{wind}_{st}.rds")
# write_rds(potential_splits_4, file_name_4)

```

Look at the deviance explained by the models with different number of
groups. Here for simplicity we directly chose max_groups_found_3 = 7. To
decide on this many groups, we initially ran the algorithm up to 30
groups.




How many lineages to choose is a practical question not something that is easy to ground in math. Elbow plot analysis is basically arbitrary. First elbow is the known dynamics, the second elbow is more exploratory. Stopping at the first linear point on horizontal line. 

Use knees to detect elbow. 


###TODO Why is D3 log less than 1 (log1 = 0)

```{r linlog group plot}
max_groups_found_line_3 = 6 #Knees + 1 say 9
max_groups_found_line_2 = 3 #Knees + 1 say 9
max_groups_found_line_1 = 8 #Knees + 1 say 10
max_groups_found_line_4 = 4 #Knees + 1 say 4

for (i in 1:4) {
  
 svglite(sprintf("3_Output_Figures/3_5_DENV/3_5_99_Supplement/DENV_%d_linlog.svg", i), width = 12, height = 6)
  par(mfrow = c(1, 2), oma = c(2, 2, 1, 1), mar = c(2, 2, 2, 0.5),
      mgp = c(0.75, 0.25, 0), cex.axis = 0.5, cex.lab = 0.5,
      cex.main = 0.7, cex.sub = 0.5)
  ## Dynamically retrieve the potential splits for the current group
  potential_splits <- get(paste0('potential_splits_', i))
  max_groups_found_line_i <- get(paste0('max_groups_found_line_', i))
  
  ## Dynamically create or update the data frame for the current group
  df_explained_dev_i <- data.frame(
    'N_groups' = 0:length(potential_splits$best_dev_explained),
    'Non_explained_deviance' = (1 - c(potential_splits$first_dev, potential_splits$best_dev_explained)),
    'Non_explained_deviance_log' = log10(1 - c(potential_splits$first_dev, potential_splits$best_dev_explained))
  )
  
  ## Linear scale plot
  plot(df_explained_dev_i$N_groups,
       df_explained_dev_i$Non_explained_deviance,
       bty = 'n', ylim = c(0, 1),
       xaxt = 'n', yaxt = 'n', pch = 16, 
       main = paste('Linear scale DENV', i), cex = 0.5,
       ylab = 'Non-explained deviance (%)', xlab = 'Number of groups')
  axis(1, lwd = 0.5, tck = -0.02)
  axis(2, las = 2, at = seq(0, 1, 0.1), labels = seq(0, 1, 0.1) * 100, lwd = 0.5, tck = -0.02)
  abline(v = max_groups_found_line_i, col = "purple", lty = 2, lwd = 0.5)
  
  ## Log scale plot
  plot(df_explained_dev_i$N_groups,
       df_explained_dev_i$Non_explained_deviance,
       log = 'y', # This is log10
       ylim = c(0.0001, 1),
       bty = 'n',
       xaxt = 'n', yaxt = 'n', pch = 16, 
       main = paste('Log scale DENV', i), cex = 0.5,
       ylab = 'Non-explained deviance (%) - log scale', xlab = 'Number of groups')
  axis(1, lwd = 0.5, tck = -0.02)
  yticks <- c(0.0001, 0.001, 0.005, 0.01, 0.05, 0.1, 0.25, 0.5, 1)
  yticks <- yticks[yticks >= 0.0001 & yticks <= 1]
  axis(2, las = 2, at = yticks, labels = yticks * 100, lwd = 0.5, tck = -0.02)
  abline(v = max_groups_found_line_i, col = "purple", lty = 2, lwd = 0.5)
  
  dev.off()
}


```

```{r AIC BIC plots}

for (i in 1:4) {
  
  i = 3
  
  potential_splits <- get(paste0('potential_splits_', i))
  mods <- potential_splits[["best_mod"]]
  output_dir <- "3_Output_Figures/3_5_DENV/3_5_99_Supplement"

df <- tibble(
  step = seq_along(mods),
  AIC  = sapply(mods, AIC),
  BIC = sapply(mods, BIC),
  dev  = sapply(mods, function(m) summary(m)$dev.expl)
)

df <- df %>%
  mutate(
    dev_pct  = dev * 100,
    nonexpl  = pmax(1 - dev, 1e-8)
  )

kmin <- df$step[which.min(df$AIC)]
kmin_BIC <- df$step[which.min(df$BIC)]

p1 <- ggplot(df, aes(step, AIC)) +
  geom_line() + geom_point() +
  geom_vline(xintercept = kmin, linetype = 2, color = "#012169aa") +
  labs(x = "Number of groups (step index)", y = "AIC",
       title = "Model AIC across tested splits",
       subtitle = sprintf("Min AIC at k=%d (AIC=%.1f)", kmin, df$AIC[kmin]))

p2 <- ggplot(df, aes(step, dev_pct)) +
  geom_line() + geom_point() +
  labs(x = "Number of groups (step index)", y = "Deviance explained (%)",
       title = glue("DENV{i} Deviance across tested splits"))


p3 <- ggplot(df, aes(step, BIC)) +
  geom_line() + geom_point() +
  geom_vline(xintercept = kmin_BIC, linetype = 2, color = "#C8102Eaa") +
  labs(x = "Number of groups (step index)", y = "BIC",
       title = glue("DENV{i} Model BIC across tested splits"),
       subtitle = sprintf("Min BIC at k=%d (BIC=%.1f)", kmin_BIC, df$BIC[kmin_BIC])) 

p4 <- ggplot(df, aes(step)) +
  geom_line(aes(y = AIC), colour = "#012169FF") + geom_point(aes(y = AIC), colour = "#012169FF") +
  geom_line(aes(y = BIC), colour = "#C8102EFF") + geom_point(aes(y = BIC), colour = "#C8102EFF") +
  geom_vline(xintercept = kmin, linetype = 2, color = "#012169aa") +
  geom_vline(xintercept = kmin_BIC, linetype = 2, color = "#C8102Eaa") +
  labs(x = "Number of groups (step index)", y = "Value",
       title = glue("DENV{i} Model AIC & BIC across tested splits"),
       subtitle = glue::glue("Min AIC at k={kmin} (AIC={round(df$AIC[kmin],1)})\n",
                      "Min BIC at k={kmin_BIC} (BIC={round(df$BIC[kmin_BIC],1)})") )


p6 <- ggplot(df, aes(step, nonexpl)) +
  geom_line() + geom_point() +
  scale_y_continuous(
    trans  = "log10",
    breaks = scales::log_breaks(n = 6),                       # data-aware log ticks
    labels = function(x) scales::percent(x, accuracy = 0.1)   # show as %
  ) +
  labs(x = "Number of groups (step index)", y = "Non-explained Deviance (%) - log scale",
       title = glue("DENV{i} Non-explained Deviance explained across tested splits (log)"))

p7 <- ggplot(df, aes(step, nonexpl*100)) +
  geom_line() + geom_point() +
  labs(x = "Number of groups (step index)", y = "Non-explained Deviance (%)",
       title = glue("DENV{i} Non-explained Deviance explained across tested splits"))



p8 <- p2 + p6 + p7



p5 <- p4 + p6 + p7

print(p1); print(p2); print(p3); print(p4); print(p5); print(p6); print(p7); print(p8)

# save if you want

# ggsave(glue("{output_dir}/aic_by_step_DENV{i}.png"), p1, width = 6, height = 4, dpi = 300)
# ggsave(glue("{output_dir}/devexp_by_step_DENV{i}.png"), p2, width = 6, height = 4, dpi = 300)
# ggsave(glue("{output_dir}/bic_by_step_DENV{i}.png"), p3, width = 6, height = 4, dpi = 300)
ggsave(glue("{output_dir}/AIC_BIC_devexp_by_step_DENV{i}.png"), p5, width = 8, height = 11, dpi = 300)
ggsave(glue("{output_dir}/Devexp_by_step_DENV{i}.png"), p8, width = 8, height = 11, dpi = 300)

}

```

Optimize the number of groups: set the minimum number of sequences per
group to 30, with a minimum frequency of 1%.


#TODO: Decide if we should do more groups, and do sensitivty analysis on differnt numbers of groups

```{r minimum groups, eval = T}

for (i in 1:4) {
  ## Dynamically retrieve the relevant timed tree and dataset_with_nodes
  timed_tree <- get(paste0('tree_DENV', i))
  dataset_with_nodes <- get(paste0('dataset_with_nodes_', i))
  potential_splits <- get(paste0('potential_splits_', i))
  
  ## Call the merge.groups function
  split <- merge.groups(
    timed_tree = timed_tree,
    metadata = dataset_with_nodes,
    initial_splits = potential_splits$potential_splits,
    group_count_threshold = 30,
    group_freq_threshold = 0.01
  )
  
  ## Dynamically assign the result to split_1, split_2, etc.
  assign(paste0('split_', i), split)
}

```

Label sequences with these new groups, and assign a color to each of
them.

```{r, eval = T}

for (i in 1:4) {
  ## Dynamically retrieve the dataset and split for the current iteration
  dataset_with_nodes <- get(paste0('dataset_with_nodes_', i))
  split <- get(paste0('split_', i))
  
  ## Label sequences with new groups
  dataset_with_nodes$groups <- as.factor(split$groups)
  
  ## Reorder labels by time of emergence
  name_groups <- levels(dataset_with_nodes$groups)
  time_groups_world <- NULL
  for (j in 1:length(name_groups)) {
    time_groups_world <- c(
      time_groups_world, 
      min(dataset_with_nodes$time[which(dataset_with_nodes$groups == name_groups[j] & 
                                        dataset_with_nodes$is.node == 'no')])
    )
  }
  
  ## Update group levels
  levels(dataset_with_nodes$groups) <- match(name_groups, order(time_groups_world, decreasing = TRUE))
  dataset_with_nodes$groups <- as.numeric(as.character(dataset_with_nodes$groups))
  dataset_with_nodes$groups <- as.factor(dataset_with_nodes$groups)
  
  ## Update names in the split list
  split$tip_and_nodes_groups <- match(split$tip_and_nodes_groups, order(time_groups_world, decreasing = TRUE))
  names(split$tip_and_nodes_groups) <- 1:length(split$tip_and_nodes_groups)
  split$groups <- as.factor(split$groups)
  levels(split$groups) <- match(name_groups, order(time_groups_world, decreasing = TRUE))
  split$groups <- as.numeric(as.character(split$groups))
  
  ## Choose color palette
  n_groups <- length(name_groups)
  colors_groups <- met.brewer(name = "Cross", n = n_groups, type = "continuous")
  
  ## Color each group
  dataset_with_nodes$group_color <- dataset_with_nodes$groups
  levels(dataset_with_nodes$group_color) <- colors_groups
  dataset_with_nodes$group_color <- as.character(dataset_with_nodes$group_color)
  
  ## Dynamically save group names and colors for this iteration
  assign(paste0('name_groups_', i), name_groups)
  assign(paste0('colors_groups_', i), colors_groups)
  
  ## Save the updated variables dynamically
  assign(paste0('dataset_with_nodes_', i), dataset_with_nodes)
  assign(paste0('split_', i), split)
  assign(paste0('colors_groups_', i), colors_groups)
}



```

#### Plot tree & index below, with colors from index-defined groups

Plot the tree and index colored with the new groups:

```{r, eval = T}

# TODO: this works but it's clunky to save as other formats. Ideally you'd save this via http://stackoverflow.com/questions/48132169/export-r-plot-to-multiple-formats
# OR you'd convert to ggplot and use ggsave


for (i in 1:4) {
  ## Dynamically retrieve the relevant variables
  tree <- get(paste0('tree_DENV', i))
  dataset_with_nodes <- get(paste0('dataset_with_nodes_', i))
  root_height <- get(paste0('root_height_', i))
  min_year <- get('min_year')  # Assuming min_year and max_year are global
  max_year <- get('max_year')
  name_groups <- get(paste0('name_groups_', i))
  colors_groups <- get(paste0('colors_groups_', i)) 
  genome_length <- get('genome_length')  # Global variables
  mutation_rate_sci <- get('mutation_rate_sci')
  timescale <- get('timescale')
  wind <- get('wind')
  
  ## Plotting
  
  svglite(filename = glue("3_Output_Figures/3_5_DENV/3_5_1_Figure 1/DENV{i}_New_Clades_MaxGroups_{st}.svg" ), width = 8, height = 10)
  
  par(mfrow = c(2, 1), oma = c(0, 0, 0, 0), mar = c(4, 4, 0, 0))
  
  ## Tree
  tree_plot <- plot(tree, show.tip.label = FALSE, 
                    edge.color = 'grey', edge.width = 0.25,
                    x.lim = c(min_year, max_year) - root_height)
  tiplabels(pch = 16, col = dataset_with_nodes$group_color, cex = 0.3)
  axisPhylo_NL(side = 1, root.time = root_height, backward = FALSE,
               at_axis = seq(min_year, max_year, 0.5) - root_height,
               lab_axis = seq(min_year, max_year, 0.5), lwd = 0.5)
  
  ## Index colored by group
  index_plot <- plot(dataset_with_nodes$time, 
                     dataset_with_nodes$index, 
                     col = adjustcolor(dataset_with_nodes$group_color, alpha.f = 1),
                     bty = 'n', xlim = c(min_year, max_year), cex = 0.5,
                     pch = 16, ylab = 'Index', xlab = 'Time (years)', yaxt = 'n')
  title(main = paste0(
    "DENV_", i, "\n",
    "Genome Length: ", genome_length, " bp\n",
    "Mutation Rate: ", mutation_rate_sci, "\n",
    "Timescale: ", timescale, " years \n",
    "Window: ", round(wind, 2), " years"
  ), line = -3.9, adj = 0, cex.main = 0.8, col.main = "grey")
  
  axis(2, las = 2, lwd = 0.5)
  axis(1, lwd = 0.5)
  
  
  ## Color key
  legend('topright', 
         legend = name_groups,
         fill = colors_groups, border = colors_groups,
         cex = 0.5, bty = 'n', ncol = 5)
  
  ## Save
  dev.off()
}






```


#### Compare NextStrain groups and groups called with the index

Generate trees colored with each set of groups next to each other:

```{r, eval = T, results = 'hide', warning=FALSE, message=FALSE}
## Tree with index-defined groups

# DENV 3
groupings_3 = matrix(dataset_with_nodes_3$groups[which(dataset_with_nodes_3$is.node == 'no')], ncol = 1)
colnames(groupings_3) = 'groups'
rownames(groupings_3) = dataset_with_nodes_3$name_seq[which(dataset_with_nodes_3$is.node == 'no')]
cols_3 = as.character(colors_groups_3)
names(cols_3) = as.character(1:max(as.numeric(name_groups_3)))
plot_tree_world_groups_3 <- ggtree(tree_DENV3, mrsd=lubridate::date_decimal(max(times_seqs_3)), size = 0.10,
                   aes(color = as.character(dataset_with_nodes_3$groups), label = label)) + 
  scale_color_manual(values = cols_3)+theme_tree2()
plot_tree_world_groups_3 = gheatmap(plot_tree_world_groups_3, groupings_3, offset=0.1, width=0.10, 
                    colnames=FALSE, legend_title="Group", color=NA) +
  scale_fill_manual(values = (cols_3))+scale_y_continuous(expand=c(0, 0.3))+theme(legend.position = 'none')

## Tree with NextStrain clades
Nextstrain_3 = matrix(dataset_with_nodes_3$clade[which(dataset_with_nodes_3$is.node == 'no')], ncol = 1)
colnames(Nextstrain_3) = 'groups'
rownames(Nextstrain_3) = dataset_with_nodes_3$name_seq[which(dataset_with_nodes_3$is.node == 'no')]
cols_NextStrain_3 = as.character(colors_clade_3)
names(cols_NextStrain_3) = clade_labels_3
plot_tree_world_Nextstrain_3 <- ggtree(tree_DENV3, mrsd=lubridate::date_decimal(max(times_seqs_3)), size = 0.10,
                                          aes(color = as.character(dataset_with_nodes_3$clade), label = label)) + 
  scale_color_manual(values = cols_NextStrain_3)+
  theme_tree2(legend = 'none')
plot_tree_world_Nextstrain_3 = gheatmap(plot_tree_world_Nextstrain_3, Nextstrain_3, offset=0.1, width=0.10, 
                     colnames=FALSE, legend_title="Group", color=NA) +
  scale_fill_manual(values = cols_NextStrain_3, na.value = 'white')+
  scale_x_reverse() + 
  scale_y_continuous(expand=c(0, 0.3))+
  theme(legend.position = 'none')

plot_tree_world_groups_3
plot_tree_world_Nextstrain_3

# DENV 2
groupings_2 = matrix(dataset_with_nodes_2$groups[which(dataset_with_nodes_2$is.node == 'no')], ncol = 1)
colnames(groupings_2) = 'groups'
rownames(groupings_2) = dataset_with_nodes_2$name_seq[which(dataset_with_nodes_2$is.node == 'no')]
cols_2 = as.character(colors_groups_2)
names(cols_2) = as.character(1:max(as.numeric(name_groups_2)))
plot_tree_world_groups_2 <- ggtree(tree_DENV2, mrsd=lubridate::date_decimal(max(times_seqs_2)), size = 0.10,
                   aes(color = as.character(dataset_with_nodes_2$groups), label = label)) + 
  scale_color_manual(values = cols_2)+theme_tree2()
plot_tree_world_groups_2 = gheatmap(plot_tree_world_groups_2, groupings_2, offset=0.1, width=0.10, 
                    colnames=FALSE, legend_title="Group", color=NA) +
  scale_fill_manual(values = (cols_2))+scale_y_continuous(expand=c(0, 0.3))+theme(legend.position = 'none')

## Tree with NextStrain clades
Nextstrain_2 = matrix(dataset_with_nodes_2$clade[which(dataset_with_nodes_2$is.node == 'no')], ncol = 1)
colnames(Nextstrain_2) = 'groups'
rownames(Nextstrain_2) = dataset_with_nodes_2$name_seq[which(dataset_with_nodes_2$is.node == 'no')]
cols_NextStrain_2 = as.character(colors_clade_2)
names(cols_NextStrain_2) = clade_labels_2
plot_tree_world_Nextstrain_2 <- ggtree(tree_DENV2, mrsd=lubridate::date_decimal(max(times_seqs_2)), size = 0.10,
                                          aes(color = as.character(dataset_with_nodes_2$clade), label = label)) + 
  scale_color_manual(values = cols_NextStrain_2)+
  theme_tree2(legend = 'none')
plot_tree_world_Nextstrain_2 = gheatmap(plot_tree_world_Nextstrain_2, Nextstrain_2, offset=0.1, width=0.10, 
                     colnames=FALSE, legend_title="Group", color=NA) +
  scale_fill_manual(values = cols_NextStrain_2, na.value = 'white')+
  scale_x_reverse() + 
  scale_y_continuous(expand=c(0, 0.3))+
  theme(legend.position = 'none')

plot_tree_world_groups_2
plot_tree_world_Nextstrain_2

# DENV1
groupings_1 = matrix(dataset_with_nodes_1$groups[which(dataset_with_nodes_1$is.node == 'no')], ncol = 1)
colnames(groupings_1) = 'groups'
rownames(groupings_1) = dataset_with_nodes_1$name_seq[which(dataset_with_nodes_1$is.node == 'no')]
cols_1 = as.character(colors_groups_1)
names(cols_1) = as.character(1:max(as.numeric(name_groups_1)))
plot_tree_world_groups_1 <- ggtree(tree_DENV1, 
                                   mrsd=lubridate::date_decimal(max(times_seqs_1)), 
                                   size = 0.10,
                   aes(color = as.character(dataset_with_nodes_1$groups), label = label)) + 
  scale_color_manual(values = cols_1)+theme_tree2()
plot_tree_world_groups_1 = gheatmap(plot_tree_world_groups_1, groupings_1, offset=0.1, width=0.10, 
                    colnames=FALSE, legend_title="Group", color=NA) +
  scale_fill_manual(values = (cols_1))+scale_y_continuous(expand=c(0, 0.3))+theme(legend.position = 'none')

## Tree with NextStrain clades
Nextstrain_1 = matrix(dataset_with_nodes_1$clade[which(dataset_with_nodes_1$is.node == 'no')], ncol = 1)
colnames(Nextstrain_1) = 'groups'
rownames(Nextstrain_1) = dataset_with_nodes_1$name_seq[which(dataset_with_nodes_1$is.node == 'no')]
cols_NextStrain_1 = as.character(colors_clade_1)
names(cols_NextStrain_1) = clade_labels_1
plot_tree_world_Nextstrain_1 <- ggtree(tree_DENV1, mrsd=lubridate::date_decimal(max(times_seqs_1)), size = 0.10,
                                          aes(color = as.character(dataset_with_nodes_1$clade), label = label)) + 
  scale_color_manual(values = cols_NextStrain_1)+
  theme_tree2(legend = 'none')
plot_tree_world_Nextstrain_1 = gheatmap(plot_tree_world_Nextstrain_1, Nextstrain_1, offset=0.1, width=0.10, 
                     colnames=FALSE, legend_title="Group", color=NA) +
  scale_fill_manual(values = cols_NextStrain_1, na.value = 'white')+
  scale_x_reverse() + 
  scale_y_continuous(expand=c(0, 0.3))+
  theme(legend.position = 'none')

plot_tree_world_groups_1
plot_tree_world_Nextstrain_1


# DENV 4
groupings_4 = matrix(dataset_with_nodes_4$groups[which(dataset_with_nodes_4$is.node == 'no')], ncol = 1)
colnames(groupings_4) = 'groups'
rownames(groupings_4) = dataset_with_nodes_4$name_seq[which(dataset_with_nodes_4$is.node == 'no')]
cols_4 = as.character(colors_groups_4)
names(cols_4) = as.character(1:max(as.numeric(name_groups_4)))
plot_tree_world_groups_4 <- ggtree(tree_DENV4, mrsd=lubridate::date_decimal(max(times_seqs_4)), size = 0.10,
                   aes(color = as.character(dataset_with_nodes_4$groups), label = label)) + 
  scale_color_manual(values = cols_4)+theme_tree2()
plot_tree_world_groups_4 = gheatmap(plot_tree_world_groups_4, groupings_4, offset=0.1, width=0.10, 
                    colnames=FALSE, legend_title="Group", color=NA) +
  scale_fill_manual(values = (cols_4))+scale_y_continuous(expand=c(0, 0.3))+theme(legend.position = 'none')

## Tree with NextStrain clades
Nextstrain_4 = matrix(dataset_with_nodes_4$clade[which(dataset_with_nodes_4$is.node == 'no')], ncol = 1)
colnames(Nextstrain_4) = 'groups'
rownames(Nextstrain_4) = dataset_with_nodes_4$name_seq[which(dataset_with_nodes_4$is.node == 'no')]
cols_NextStrain_4 = as.character(colors_clade_4)
names(cols_NextStrain_4) = clade_labels_4
plot_tree_world_Nextstrain_4 <- ggtree(tree_DENV4, mrsd=lubridate::date_decimal(max(times_seqs_4)), size = 0.10,
                                          aes(color = as.character(dataset_with_nodes_4$clade), label = label)) + 
  scale_color_manual(values = cols_NextStrain_4)+
  theme_tree2(legend = 'none')
plot_tree_world_Nextstrain_4 = gheatmap(plot_tree_world_Nextstrain_4, Nextstrain_4, offset=0.1, width=0.10, 
                     colnames=FALSE, legend_title="Group", color=NA) +
  scale_fill_manual(values = cols_NextStrain_4, na.value = 'white')+
  scale_x_reverse() + 
  scale_y_continuous(expand=c(0, 0.3))+
  theme(legend.position = 'none')




```


This doesn't work due to aes issues coloring the whole tree

```{r dyanmic classical vs new}
for (i in 1:4) {

## Dynamically retrieve the relevant variables
  tree <- get(paste0('tree_DENV', i))
  dataset_with_nodes <- get(paste0('dataset_with_nodes_', i))
  root_height <- get(paste0('root_height_', i))
  min_year <- get('min_year')  # Assuming min_year and max_year are global
  max_year <- get('max_year')
  name_groups <- get(paste0('name_groups_', i))
  colors_groups <- get(paste0('colors_groups_', i)) 
  colors_clade <- get(paste0('colors_clade_', i))
  genome_length <- get('genome_length')  # Global variables
  mutation_rate_sci <- get('mutation_rate_sci')
  times_seqs <- get(paste0('times_seqs_', i ))
  names_seqs <- get(paste0('names_seqs_', i ))
  clade_labels <- get(paste0('clade_labels_', i))
  timescale <- get('timescale')
  wind <- get('wind')
  
  
tip_df <- filter(dataset_with_nodes, is.node == "no") %>% 
  transmute(
    label  = name_seq,         
    groups = factor(groups)
  )
  


groupings = matrix(dataset_with_nodes$groups[which(dataset_with_nodes$is.node == 'no')], ncol = 1)
colnames(groupings) = 'groups'
rownames(groupings) = dataset_with_nodes$name_seq[which(dataset_with_nodes$is.node == 'no')]
cols = as.character(colors_groups)
names(cols) = as.character(1:max(as.numeric(name_groups)))

plot_tree_world_groups <- ggtree(tree, 
                                   mrsd=lubridate::date_decimal(max(times_seqs)), 
                                   size = 0.10,
                   aes(color = as.character(dataset_with_nodes$groups), label = label)) + 
  geom_tippoint(color = dataset_with_nodes$group_color[which(dataset_with_nodes$is.node == 'no')], size = 0.25) +
  scale_color_manual(values = cols)+theme_tree2()

## build a proper 1-column heatmap table
hm <- data.frame(Group = factor(tip_df$groups, levels = levels(tip_df$groups)))
rownames(hm) <- tip_df$label

## enforce row order == tree$tip.label
hm <- hm[tree$tip.label, , drop = FALSE]

plot_tree_world_groups = gheatmap(plot_tree_world_groups, hm, 
                                  offset=0.1, 
                                  width=0.10, 
                                  colnames=FALSE, 
                                  legend_title="Group", 
                                  color=NA
                                  ) +
  scale_fill_manual(values = (cols)) +
  scale_y_continuous(expand=c(0, 0.3)) + 
  theme(legend.position = 'none')


## Tree with NextStrain clades
Nextstrain = matrix(dataset_with_nodes$clade[which(dataset_with_nodes$is.node == 'no')], ncol = 1)
colnames(Nextstrain) = 'groups'
rownames(Nextstrain) = dataset_with_nodes$name_seq[which(dataset_with_nodes$is.node == 'no')]
cols_NextStrain = as.character(colors_clade)
names(cols_NextStrain) = clade_labels
plot_tree_world_Nextstrain <- ggtree(tree, 
                                     mrsd=lubridate::date_decimal(max(times_seqs)), 
                                     size = 0.10,
                                     aes(color = as.character(dataset_with_nodes$clade), label = label)) +
  geom_tippoint(color = dataset_with_nodes$clade_color[which(dataset_with_nodes$is.node == 'no')], size = 0.25) +
  scale_color_manual(values = cols_NextStrain)+theme_tree2()
  

plot_tree_world_Nextstrain = gheatmap(plot_tree_world_Nextstrain, Nextstrain, offset=0.1, width=0.10, 
                     colnames=FALSE, legend_title="Group", color=NA) +
  scale_fill_manual(values = cols_NextStrain, na.value = 'white')+
  scale_x_reverse() + 
  scale_y_continuous(expand=c(0, 0.3))+
  theme(legend.position = 'none')
  
  
assign(paste0('plot_tree_world_Nextstrain_', i), plot_tree_world_Nextstrain)

assign(paste0('plot_tree_world_groups_', i), plot_tree_world_groups)
  




}

```

This works but doesn't label the tree (below)

```{r}


for (i in 1:4) {

  tree  <- get(paste0('tree_DENV', i))
  dwn   <- get(paste0('dataset_with_nodes_', i))
  times <- get(paste0('times_seqs_', i))
  cols_groups <- as.character(get(paste0('colors_groups_', i)))
  names(cols_groups) <- levels(factor(dwn$groups[dwn$is.node == "no"]))

  # tip metadata
  tip_df <- dplyr::filter(dwn, is.node == "no") |>
    dplyr::transmute(
      label  = name_seq,
      groups = factor(groups)
    )

  # base tree: no color mapping here
  p <- ggtree(tree,
              mrsd = lubridate::date_decimal(max(times)),
              size = 0.10) +
       theme_tree2()

  # attach tip data and color ONLY on the tip layer
  p <- (p %<+% tip_df) +
       geom_tippoint(aes(color = groups), size = 0.25) +
       scale_color_manual(values = cols_groups, drop = FALSE)

  # heatmap table (1 col, rownames = tip labels, same order)
  hm <- data.frame(Group = tip_df$groups)
  rownames(hm) <- tip_df$label
  hm <- hm[tree$tip.label, , drop = FALSE]

  p_hm <- gheatmap(p, hm,
                   offset = 0.1, width = 0.10,
                   colnames = FALSE, legend_title = "Group",
                   color = NA) +
          scale_fill_manual(values = cols_groups, drop = FALSE) +
          scale_y_continuous(expand = c(0, 0.3)) +
          theme(legend.position = "none")

  assign(paste0('plot_tree_world_groups_', i), p_hm, inherits = TRUE)
}


plot_tree_world_groups_1
plot_tree_world_Nextstrain_1

plot_tree_world_groups_2
plot_tree_world_Nextstrain_2

plot_tree_world_groups_3
plot_tree_world_Nextstrain_3

plot_tree_world_groups_4
plot_tree_world_Nextstrain_4


```




Dynamically doing the above (Doesn't work due to aethetics inside the
loop issues)

Plot the generated trees side by side Dynamcially

```{r}

for (i in 1:4) {
  
  # Dynamically retrieve the group and NextStrain plots
  plot_tree_world_groups <- get(paste0('plot_tree_world_groups_', i))
  plot_tree_world_Nextstrain <- get(paste0('plot_tree_world_Nextstrain_', i))
  
  # Side-by-side static plot using plot_grid
  combined_plot <- plot_grid(
    plot_tree_world_groups, 
    plot_tree_world_Nextstrain,
    rel_widths = c(1, 1), 
    labels = c(paste('Phylowave clades DENV', i, sep = ''), paste('Classical clades DENV', i, sep = '')), 
    label_size = 10, 
    label_x = c(0.1, 0.25), 
    ncol = 2
  )
  
  # Save the static combined plot
  ggsave(filename = paste0("3_Output_Figures/3_5_DENV/3_5_1_Figure 1/","DENV_", i, "_", "Classical_vs_Phylowave_Clades_combined_plot_", st, ".svg"), plot = combined_plot, width = 12, height = 6)
  
  # Side-by-side interactive plot using ggplotly and subplot
  interactive_plot <- ggplotly(
    subplot(plot_tree_world_groups, plot_tree_world_Nextstrain, widths = c(0.5, 0.5))
  )

  
  # Save the interactive plot as an HTML file
  htmlwidgets::saveWidget(interactive_plot, file = glue("3_Output_Figures/3_5_DENV/3_5_1_Figure 1/Interactives/DENV_{i}_Classical_vs_Phylowave_Clades_interactive_plot_{st}.html"))
}



```


## Quantify the fitness of detected DENV lineage

#### Run the fitness model

```{r debug model, eval=FALSE}

min_year_model <- 1985
max_year_model <- 2015

options(mc.cores = 8) #Set to coresMore actions
start_time = Sys.time()
## Load and compile stan code (this can take a few minutes)


## If you get an error about path mapping to a previous user 
##make sure to go into the model/scripts folder and delete the old unix model before making this one or run: cmdstanr::rebuild_cmdstan()
model_compiled <- cmdstan_model(stan_file = '2_Functions/Model_multinomial_logistic_birthdeath_lineage_fitness_20231220.stan')
## Run model on the groups

# model_compiled$print()

res_fitness_3 = estimate_rel_fitness_groups_with_branches(dataset_with_nodes = dataset_with_nodes_3,
                                                        tree = tree_DENV3,
                                                        min_year = min_year_model, 
                                                        window = wind, 
                                                        model_compiled = model_compiled,
                                                        iter_warmup = 1000, 
                                                        iter_sampling = 2000, 
                                                        refresh = 50, 
                                                        seed = 1) #increased warmup and sampling from 250 and 500


res_fitness_2 = estimate_rel_fitness_groups_with_branches(dataset_with_nodes = dataset_with_nodes_2,
                                                        tree = tree_DENV2,
                                                        min_year = min_year_model, 
                                                        window = wind, 
                                                        model_compiled = model_compiled,
                                                        iter_warmup = 1000, 
                                                        iter_sampling = 2000, 
                                                        refresh = 50, 
                                                        seed = 1)

res_fitness_1 = estimate_rel_fitness_groups_with_branches(dataset_with_nodes = dataset_with_nodes_1,
                                                        tree = tree_DENV1,
                                                        min_year = min_year_model, 
                                                        window = wind, 
                                                        model_compiled = model_compiled,
                                                        iter_warmup = 1000, 
                                                        iter_sampling = 2000, 
                                                        refresh = 50, 
                                                        seed = 1)


res_fitness_4 = estimate_rel_fitness_groups_with_branches(dataset_with_nodes = dataset_with_nodes_4,
                                                        tree = tree_DENV4,
                                                        min_year = min_year_model, 
                                                        window = wind, 
                                                        model_compiled = model_compiled,
                                                        iter_warmup = 1000, 
                                                        iter_sampling = 2000, 
                                                        refresh = 50, 
                                                        seed = 1)
end_time = Sys.time()
print(end_time - start_time)
```

```{r save fitness model}

for (i in 1:4) {
res_fitness <- get(glue("res_fitness_{i}"))
write_rds(x = res_fitness, file = glue("4_Output_Data/res_fintess_{i}_max_groups{st}.rds"))

}

```



```{r main model params}

library(loo)

source(file = '2_Functions/2_1_Index_computation_20240909.R')
source(file = '2_Functions/2_2_Lineage_detection_20240909.R')
source(file = '2_Functions/2_3_Lineage_fitness_20240909.R')



min_year <- min_year_model
max_year <- max_year_model


options(mc.cores = 9) #Set to 14 cores

## Load and compile stan code (this can take a few minutes)

## If you get an error about path mapping to a previous user 
##make sure to go into the model/scripts folder and delete the old unix model before making this one or run: cmdstanr::rebuild_cmdstan()
# Define a list of serotypes and associated variables
serotypes <- list(
  DENV1 = list(dataset = dataset_with_nodes_1, tree = tree_DENV1),
  DENV2 = list(dataset = dataset_with_nodes_2, tree = tree_DENV2),
  DENV3 = list(dataset = dataset_with_nodes_3, tree = tree_DENV3),
  DENV4 = list(dataset = dataset_with_nodes_4, tree = tree_DENV4)
)

# Compile both models
model_static <- cmdstan_model("2_Functions/Model_multinomial_logistic_birthdeath_lineage_fitness_20231220.stan")
model_timeshift <- cmdstan_model("2_Functions/Model_multinomial_logistic_birthdeath_lineage_fitness_timeshift_20250617.stan")

breakpoints <- seq(min_year_model, max_year_model, by = 0.5)

```

# On Loo:

From: <https://users.aalto.fi/~ave/CV-FAQ.html> \#### 21 How are LOO and
WAIC related? LOO is an cross-validation approach which can be used to
estimate expected utility and loss functions. If log score is used, LOO
can be used to estimate ELPD and we write elpd_loo.

WAIC is an computational method to estimate ELPD and we could write
elpd_waic. Watanabe used log score, but as WAIC has often be represented
as an alternative to DIC which used -2 \* log score, WAIC is sometimes
use to estimate -2*ELPD. See also How are LOOIC and elpd_loo related?
Why LOOIC is -2*elpd_loo?.

In theory, the computational method used in WAIC, which corresponds to a
truncated Taylor series approximation of leave-one-out cross-validation,
could be used with other smooth utilities and loss functions than log
score (Vehtari and Ojanen, 2012), but we’re not aware of people doing
that and we don’t recommend it as PSIS-LOO has better diagnostics.

All limitations when LOO is valid or sensible hold also for WAIC (see
When is cross-validation valid?, Can cross-validation be used for
hierarchical/multilevel models?, Can cross-validation be used for time
series?). Thinking in terms of LOO cross-validation, it is easier to
move to other cross-validation data division schemes.

Vehtari, Gelman and Gabry (2017) show that PSIS-LOO has usually smaller
error in estimating ELPD than WAIC. The exception is the case when p_loo
≪𝑁 , as then WAIC tends to have slightly smaller error, but in that case
both PSIS-LOO and WAIC have very small error and it doesn’t matter which
computational approximation is used. On the other hand, for flexible
models WAIC fails more easily, has significant bias and is less easy to
diagnose for failures. WAIC has been included in loo package only for
comparison purposes and to make it easy to replicate the results by
Vehtari, Gelman and Gabry (2017).

#### 12 What is the interpretation of ELPD / elpd_loo / elpd_diff?

Log densities and log probabilities can be transformed to densities and
probabilities which have intrinsic interpretation, although most people
are not well calibrated for the values as they are not used to think in
densities and probabilities and even less in log densities and log
probabilities.

The log probabilities are easier. For example, Guido Biele had a problem
computing elpd_loo with a beta-binomial model for data with 22
categories. Computed individual elpd_loo values for observations were
around -461. For discrete model with uniform probability for 22
categories log probabilities would be log(1/22)≈−3.1 , and thus there
was two orders of magnitude error in log scale. With the fixed code
individual elpd_loo values were about −2.3\>−3.1 , that is, the model
was beating the uniform distribution.

The log densities are more difficult as they require knowing possible
scaling or transformations of the data. See more in Can cross-validation
be used to compare different observation models / response distributions
/ likelihoods?.

Although ELPD is good for model comparison as it measures the goodness
of the whole predictive distribution, the difference in ELPD is even
more difficult to interpret without some practice, and thus we recommend
to also use application specific utility or loss functions. See more in
Can other utility and loss functions be used than log predictive
density?.

As quick rule: If elpd difference (elpd_diff in loo package) is less
than 4, the difference is small (Sivula, Magnusson and Vehtari, 2020, p.
McLatchie+etal:2023). If elpd difference (elpd_diff in loo package) is
larger than 4, then compare that difference to standard error of
elpd_diff (provided e.g. by loo package) (Sivula, Magnusson and Vehtari,
2020). The value for deciding what is small or large can be based on
connection to Pseudo-BMA+-weights (Yao et al., 2018, p.
McLatchie+etal:2023). See also How to interpret in Standard error (SE)
of elpd difference (elpd_diff)? and Case studies for the paper
Uncertainty in Bayesian LOO-CV Model Comparison

Quantify the fitness of each group you can run the code (this steps
takes approximately 12hrs on 14 cores):

```{r main model, eval = F, results = 'hide', warning=FALSE, message=FALSE}


# Loop over each serotype
results <- list()
res_fitness_all <- list()        # save no-breakpoint models
res_fitness_breakpoints <- list() # save models for each breakpoint

start_time = Sys.time()

for (i in seq_along(serotypes)) {
  serotype_name <- names(serotypes)[i]
  cat("Running for", serotype_name, "\n")

  dataset <- serotypes[[i]]$dataset
  tree <- serotypes[[i]]$tree

  # No-breakpoint model
  res_no_bp <- estimate_rel_fitness_groups_with_branches(
    dataset_with_nodes = dataset,
      tree = tree,
      min_year = min_year_model,
      window = wind,
      model_compiled = model_static,
      iter_warmup = 1000,
      iter_sampling = 2000,
      seed = 1,
      t_breakpoint = NULL,
      max_treedepth = 15,       # increased from 12
      adapt_delta = 0.99        # higher for complex posteriors
  )
  
  saveRDS(res_no_bp, file = paste0("4_Output_Data/", serotype_name, "_no_bp_", min_year_model, ".rds"))

  # Save no-breakpoint model result
  res_fitness_all[[i]] <- res_no_bp
  assign(paste0("res_fitness_", i), res_no_bp, envir = .GlobalEnv)

  loo_scores <- list()
  loo_scores[["no_breakpoint"]] <- loo(res_no_bp$fit$draws("log_lik"))

  # Store all models for each breakpoint
  model_fits_bp <- list()

  for (bp in breakpoints) {
    cat("  └── Fitting ", serotype_name, " breakpoint model at year:", bp, "\n")

    fit_result <- estimate_rel_fitness_groups_with_branches(
      dataset_with_nodes = dataset,
      tree = tree,
      min_year = min_year_model,
      window = wind,
      model_compiled = model_timeshift,
      iter_warmup = 1000,
      iter_sampling = 2000,
      seed = 1,
      t_breakpoint = bp,
      max_treedepth = 15,       # increased from 12
      adapt_delta = 0.99        # higher for complex posteriors
    )

    loo_scores[[as.character(bp)]] <- loo(fit_result$fit$draws("log_lik"))
    model_fits_bp[[as.character(bp)]] <- fit_result
  }

  # Save all breakpoint models
  res_fitness_breakpoints[[i]] <- model_fits_bp
  
  saveRDS(model_fits_bp, file = paste0("4_Output_Data/", serotype_name, "_bps_", min_year_model, ".rds"))

  # LOO comparison
  loo_comparison <- loo::loo_compare(loo_scores)
  loo_df <- as.data.frame(loo_comparison)
  loo_df$breakpoint <- rownames(loo_df)

  loo_df <- loo_df %>%
    mutate(
      breakpoint_num = as.numeric(breakpoint),
      breakpoint_num = ifelse(breakpoint == "no_breakpoint", 2030, breakpoint_num)
    )
  
  saveRDS(loo_df, file = paste0("4_Output_Data/", serotype_name, "_loo_df.rds"))
  
  best_year <- loo_df %>%
    filter(elpd_diff == max(elpd_diff, na.rm = TRUE)) %>%
    pull(breakpoint_num)

  results[[serotype_name]] <- list(
    loo_df = loo_df,
    best_year = best_year,
    res_no_bp = res_no_bp,
    res_bps = model_fits_bp
  )
}


#Help to assess degree of imporvement
# Understanding elpd_diff / se_diff
# This ratio behaves like a standardized effect size for model comparison:
# It tells us how much better the top model is than others, relative to uncertainty.
# Analogous to a z-score:
# >1.96 → ~95% confidence one model is better than the other.
# However, the LOO authors caution:
# "The elpd_diff / se_diff can be optimistic because elpd_diff is computed from a shared dataset, so models are not truly independent."
# — Vehtari, Gelman, Gabry (2017), "Practical Bayesian model evaluation using leave-one-out cross-validation and WAIC" (arXiv:1507.04544)

annotate_loo_df <- function(loo_df) {
  loo_df <- loo_df %>%
    mutate(
      elpd_ratio = elpd_diff / se_diff,
      evidence_strength = case_when(
        is.na(elpd_ratio) ~ "NA",
        abs(elpd_ratio) > 5 ~ "Strong",
        abs(elpd_ratio) > 1.96 ~ "Moderate",
        TRUE ~ "Weak"
      )
    )
  return(loo_df)
}

DENV4_loo_df <-  annotate_loo_df(DENV4_loo_df)

```



Alternatively load the previous results

```{r, eval = FALSE}
# Assume: names(serotypes) == c("DENV1", "DENV2", "DENV3", "DENV4")
# Assume: min_year_model is defined (same as when saving files)

results <- list()  # Reinitialize results

for (i in seq_along(serotypes)) {
  serotype_name <- names(serotypes)[i]
  cat("Loading saved results for", serotype_name, "\n")

  # Load from previously saved RDS files
  res_no_bp <- readRDS(paste0("4_Output_Data/", serotype_name, "_no_bp_", min_year_model, ".rds"))
  res_bps   <- readRDS(paste0("4_Output_Data/", serotype_name, "_bps_", min_year_model, ".rds"))
  loo_df    <- readRDS(paste0("4_Output_Data/", serotype_name, "_loo_df.rds"))

  # Re-identify best breakpoint year
  best_year <- loo_df %>%
    filter(elpd_diff == max(elpd_diff, na.rm = TRUE)) %>%
    pull(breakpoint_num)

  # Store in results using original structure
  results[[serotype_name]] <- list(
    loo_df = loo_df,
    best_year = best_year,
    res_no_bp = res_no_bp,
    res_bps = res_bps
  )
}
```


```{r plot breakpoints, eval = F, results = 'hide', warning=FALSE, message=FALSE}

 

plot_list <- list()

for (serotype in names(results)) {
  loo_df <- results[[serotype]]$loo_df
  best_year <- results[[serotype]]$best_year
  
  png(filename = paste0("3_Output_Figures/breakpoint_plot_DENV_", serotype, "_" , st, min_year_model, ".png"), width = 1600, height = 1400, res = 300)
  
  p <- ggplot(loo_df, aes(x = breakpoint_num, y = elpd_diff)) +
    geom_point() +
    geom_errorbar(aes(ymin = elpd_diff - se_diff, ymax = elpd_diff + se_diff), width = 0.1) +
    geom_vline(xintercept = best_year, linetype = "dashed", color = "blue") +
    geom_vline(xintercept = 2030, linetype = "dotted", color = "red") +
    annotate("text", x = best_year, y = max(loo_df$elpd_diff, na.rm = TRUE),
             label = paste("Best model:", best_year), hjust = -0.1, vjust = -1, color = "blue") +
    annotate("text", x = 2030, y = max(loo_df$elpd_diff, na.rm = TRUE),
             label = "No Breakpoint", hjust = 1.1, vjust = 1, color = "red") +
    geom_hline(yintercept = 0, linetype = "dashed") +
    labs(
      x = "Breakpoint Year",
      y = "ELPD Difference vs. Best Model",
      title = paste("Model Fit Across Breakpoints -", serotype)
    ) +
    theme_minimal()
  
  print(p)  # Show plot for each serotype
  
    
  # Close the PNG device
  dev.off()
  
  plot_list[[serotype]] <- p  # Save ggplot object in list
}

plot_list[["DENV1"]]
plot_list[["DENV2"]]
plot_list[["DENV3"]]
plot_list[["DENV4"]]
```

You might encounter a warning saying that '*alpha_true_GA*' has a
missing init value - this is normal as those groups (ancestral groups
that are not present at the start of the time series) do not always
exist and therefore there is no default initial value. This is does not
impact the model run. The seed has been set to 1 so allow for
reproducible results.

To save some time, you may wish to load the results:

```{r, eval = T}
# res_fitness_1 = readRDS("4_Output_data/res_fitness_DENV1_timescale7_wind1_2025-01-06_2234.rds")
# res_fitness_2 = readRDS("4_Output_data/res_fitness_DENV2_timescale7_wind1_2025-01-06_2234.rds")
# res_fitness_3 = readRDS("4_Output_data/res_fitness_DENV3_timescale7_wind1_2025-01-06_2234.rds")
# res_fitness_4 = readRDS("4_Output_data/res_fitness_DENV4_timescale7_wind1_2025-01-06_2234.rds")




# #Write
# file_name_fitness_3 <- glue("4_Output_data/res_fitness_DENV3_timescale{timescale}_wind{wind}_{st}.rds")
# file_name_fitness_2 <- glue("4_Output_data/res_fitness_DENV2_timescale{timescale}_wind{wind}_{st}.rds")
# file_name_fitness_1 <- glue("4_Output_data/res_fitness_DENV1_timescale{timescale}_wind{wind}_{st}.rds")
# file_name_fitness_4 <- glue("4_Output_data/res_fitness_DENV4_timescale{timescale}_wind{wind}_{st}.rds")
# 
# write_rds(res_fitness_3, file_name_fitness_3)
# write_rds(res_fitness_2, file_name_fitness_2)
# write_rds(res_fitness_1, file_name_fitness_1)
# write_rds(res_fitness_4, file_name_fitness_4)

```

#### Plot the fits and estimated parameters

Plot the fits:

```{r, eval = FALSE}
order_colors = order(as.numeric(split_i$tip_and_nodes_groups))

#assign order_colors_i here

colour_lineage = colors_groups_i[match(split_i$tip_and_nodes_groups[order_colors], name_groups_i)]
#assign color_lineage_i here

plot_fit_data_new(data = res_fitness_i$data,
                  Chains = res_fitness_i$chains,
                  colour_lineage = colour_lineage,
                  xmin = min_year_model, xmax = max_year)

# Color key
legend('topright',
       legend = name_groups_i,
       fill = colour_lineage, border = colour_lineage,
       cex = 0.5, bty = 'n', ncol = 5)
```

```{r}
for (i in 1:4) {
  # Retrieve relevant variables
  split <- get(paste0('split_', i))
  res_fitness <- get(paste0('res_fitness_', i))
  name_groups <- get(paste0('name_groups_', i))
  colors_groups <- get(paste0('colors_groups_', i))
  
  # Order colors dynamically
  order_colors <- order(as.numeric(split$tip_and_nodes_groups))
  assign(paste0('order_colors_', i), order_colors)
  
  # Assign lineage colors dynamically
  colour_lineage <- colors_groups[match(split$tip_and_nodes_groups[order_colors], name_groups)]
  assign(paste0('colour_lineage_', i), colour_lineage)
  
  png(filename = glue("3_Output_Figures/3_5_DENV/3_5_2_Figure 2/Fitness_plot_DENV_{i}_{st}_min_year_{min_year_model}.png"), width = 1600, height = 1400, res = 300)
  
  # Plot fitness data
  plot_fit_data_new(
    data = res_fitness$data,
    Chains = res_fitness$chains,
    colour_lineage = colour_lineage,
    xmin = min_year_model, 
    xmax = max_year_model
  )
  
  # Add legend
  legend(
    'topright', 
    legend = name_groups,
    fill = colour_lineage, 
    border = colour_lineage,
    cex = 0.5, bty = 'n', ncol = 5
  )
  
  # Close the PNG device
  dev.off()
}


## Best year model

for (i in 1:4) {
  serotype_name <- paste0("DENV", i)

# Extract all results for this serotype
  res_bps <- results[[serotype_name]]$res_bps
  loo_df <- results[[serotype_name]]$loo_df

  # Get best breakpoint year (exclude 2030)
  best_year <- loo_df %>%
    filter(breakpoint_num != 2030) %>%
    slice_max(elpd_diff, n = 1, with_ties = FALSE) %>%
    pull(breakpoint)
  
  assign(paste0('best_year_', i), best_year)

  # Confirm it's a character and matches res_bps keys
  if (!best_year %in% names(res_bps)) {
    stop(paste("Breakpoint", best_year, "not found in res_bps for", serotype_name))
  }

  res_fit <- res_bps[[best_year]]
  assign(paste0('res_fitness_best_year_', i), res_fit)

  # Retrieve plotting variables
  split <- get(paste0('split_', i))
  name_groups <- get(paste0('name_groups_', i))
  colors_groups <- get(paste0('colors_groups_', i))
  order_colors <- get(paste0('order_colors_', i))
  colour_lineage <- get(paste0('colour_lineage_', i))

  # Define actual xmin/xmax from model results
  xmin_actual <- max(min_year, min(res_fit$data$t, na.rm = TRUE))
  xmax_actual <- max(res_fit$data$t, na.rm = TRUE)

  # Plot
  png(filename = paste0("3_Output_Figures/fitness_plot_DENV_", i, "_bp_", best_year, "_min_year_", min_year_model, ".png"), width = 1600, height = 1400, res = 300)

  plot_fit_data_new(
    data = res_fit$data,
    Chains = res_fit$chains,
    colour_lineage = colour_lineage,
    xmin = xmin_actual,
    xmax = xmax_actual
  )

  legend(
    'topright',
    legend = name_groups,
    fill = colour_lineage,
    border = colour_lineage,
    cex = 0.5, bty = 'n', ncol = 5
  )

  dev.off()
}
```

Plot the predicted vs observed proportions:

```{r observed vs predicted single, eval=FALSE}
plot_observed_vs_predicted(data = res_fitness$data,
                           Chains = res_fitness$chains,
                           colour_lineage = colour_lineage)

```

```{r observed vs predicted dynamic loop}
for (i in 1:4) {
  res_fitness <- get(paste0('res_fitness_', i))
  colour_lineage <- get(paste0('colour_lineage_', i))
  png(filename = paste0("3_Output_Figures/observed_vs_predicted_DENV_", i , "_", st, "_min_year_", min_year_model, ".png"), width = 1600, height = 1400, res = 300)

  plot_observed_vs_predicted(
    data = res_fitness$data,
    Chains = res_fitness$chains,
    colour_lineage = colour_lineage
  )
  
  dev.off()
}

## For best year model
for (i in 1:4) {

  # Retrieve plotting variables
  serotype_name <- paste0("DENV", i)
  best_year <- get(paste0('best_year_', i))
  res_fit <- get(paste0('res_fitness_best_year_', i))
  colour_lineage <- get(paste0('colour_lineage_', i))
  
  png(filename = paste0("3_Output_Figures/observed_vs_predicted_DENV_", i, "min_year_", min_year, "_bp_", best_year, "_" ,st, ".png"), width = 1600, height = 1400, res = 300)

  plot_observed_vs_predicted(
    data = res_fit$data,
    Chains = res_fit$chains,
    colour_lineage = colour_lineage
  )
  
  dev.off()
  
}

```

Plot raw fitness estimates:

```{r relative fitness dynamic loop}
for (i in 1:4) {
  res_fitness <- get(paste0('res_fitness_', i))
  colour_lineage <- get(paste0('colour_lineage_', i))
  # Save the plot as a PNG
  png(filename = paste0("3_Output_Figures/estimated_fitness_ref_ancestral_DENV_", i, "min_year_", min_year, "_", st, ".png"), width = 1600, height = 1200, res = 300)
  
  # Plot estimated fitness relative to ancestral reference
  plot_estimated_fitness_ref_ancestral(
    data = res_fitness$data,
    Chains = res_fitness$chains,
    colour_lineage = colour_lineage,
    gentime = 1
  )
  

  dev.off()
}




#Best Yyear 
for (i in 1:4) {
  res_fitness <- get(paste0('res_fitness_best_year_', i))
  colour_lineage <- get(paste0('colour_lineage_', i))
  best_year <- get(paste0('best_year_', i))
  # Save the plot as a PNG
  png(filename = paste0("3_Output_Figures/estimated_fitness_ref_ancestral_DENV_", i, "min_year_", min_year, "_bp_", best_year, "_" , st, ".png"), width = 1600, height = 1200, res = 300)
  
  # Plot estimated fitness relative to ancestral reference
  plot_estimated_fitness_ref_ancestral_breakpoint(
    data = res_fitness$data,
    Chains = res_fitness$chains,
    colour_lineage = colour_lineage,
    gentime = 1 # Gentime is set to 1 year 
  )
  

  dev.off()
}



plot_estimated_fitness_ref_ancestral(
    data = res_fitness_3$data,
    Chains = res_fitness_3$chains,
    colour_lineage = colour_lineage_3,
    gentime = 1
  )

plot_estimated_fitness_ref_ancestral_breakpoint(
    data = res_fitness_best_year_3$data,
    Chains = res_fitness_best_year_3$chains,
    colour_lineage = colour_lineage_3,
    gentime = 1
  )

## TODO: Ask noemie why gentime is 1 rather than actually generation time and why the ancestral strain set to 0 does not correspond with the most ancestral linage

```

Alternatively you may wish to plot the model, observed vs predicted and
relative fitness graphs together

```{r three panel fitness plots }

for (i in 1:4) {

## ------ Model Plot ------ ##

# Retrieve relevant variables
  split <- get(paste0('split_', i))
  res_fitness <- get(paste0('res_fitness_', i))
  name_groups <- get(paste0('name_groups_', i))
  colors_groups <- get(paste0('colors_groups_', i))
  
  # Order colors dynamically
  order_colors <- order(as.numeric(split$tip_and_nodes_groups))
  assign(paste0('order_colors_', i), order_colors)
  
  # Assign lineage colors dynamically
  colour_lineage <- colors_groups[match(split$tip_and_nodes_groups[order_colors], name_groups)]
  assign(paste0('colour_lineage_', i), colour_lineage)
  
  svglite(filename = paste0("3_Output_Figures/threepanel_fitness_plot_DENV_", i, "_", st, ".svg"), width = 3.33)
  
  par(mfrow = c(3, 1), oma = (c(1,1,2,1)+0.1), mar=(c(4,4,1,1)+ 0.1)) #c(bottom, left, top, right)
  
  # Plot fitness data
  
  plot_fit_data_new(
    data = res_fitness$data,
    Chains = res_fitness$chains,
    colour_lineage = colour_lineage,
    xmin = min_year, 
    xmax = max_year
  )
  
  # Add legend
  legend(
    'topright', 
    legend = name_groups,
    fill = colour_lineage, 
    border = colour_lineage,
    cex = 0.5, bty = 'n', ncol = 5
  )
  
## ------ Obs vs predicted Plot ------ ##
  plot_observed_vs_predicted(
    data = res_fitness$data,
    Chains = res_fitness$chains,
    colour_lineage = colour_lineage
  )

## ------ Relative Fitness Plot ------ ##
  # Plot estimated fitness relative to ancestral reference
  plot_estimated_fitness_ref_ancestral(
    data = res_fitness$data,
    Chains = res_fitness$chains,
    colour_lineage = colour_lineage,
    gentime = 1
  )
  
  mtext(paste0("DENV", i), line=0, side=3, outer=TRUE, cex=1)
  
  dev.off()
 
} 



# Breakpoint
for (i in 1:4) {

## ------ Model Plot ------ ##

# Retrieve relevant variables
  split <- get(paste0('split_', i))
  res_fitness <- get(paste0('res_fitness_best_year_', i))
  name_groups <- get(paste0('name_groups_', i))
  colors_groups <- get(paste0('colors_groups_', i))
  best_year <- get(paste0('best_year_', i))
  
  # Order colors dynamically
  order_colors <- order(as.numeric(split$tip_and_nodes_groups))
  assign(paste0('order_colors_', i), order_colors)
  
  # Assign lineage colors dynamically
  colour_lineage <- colors_groups[match(split$tip_and_nodes_groups[order_colors], name_groups)]
  assign(paste0('colour_lineage_', i), colour_lineage)
  
  svglite(filename = paste0("3_Output_Figures/threepanel_fitness_plot_DENV_", i, "bp_", best_year,"_" , st, ".svg"), width = 3.33)
  
  par(mfrow = c(3, 1), oma = (c(1,1,2,1)+0.1), mar=(c(4,4,1,1)+ 0.1)) #c(bottom, left, top, right)
  
  # Plot fitness data
  
  plot_fit_data_new(
    data = res_fitness$data,
    Chains = res_fitness$chains,
    colour_lineage = colour_lineage,
    xmin = min_year, 
    xmax = max_year
  )
  
  # Add legend
  legend(
    'topright', 
    legend = name_groups,
    fill = colour_lineage, 
    border = colour_lineage,
    cex = 0.5, bty = 'n', ncol = 5
  )
  
## ------ Obs vs predicted Plot ------ ##
  plot_observed_vs_predicted(
    data = res_fitness$data,
    Chains = res_fitness$chains,
    colour_lineage = colour_lineage
  )

## ------ Relative Fitness Plot ------ ##
  # Plot estimated fitness relative to ancestral reference
  plot_estimated_fitness_ref_ancestral_breakpoint(
    data = res_fitness$data,
    Chains = res_fitness$chains,
    colour_lineage = colour_lineage,
    gentime = 1
  )
  
  mtext(paste0("DENV", i), line=0, side=3, outer=TRUE, cex=1)
  
  dev.off()
 
} 
  
  
```

# Lineage-defining mutations

## Getting SNP Data

```{r location_reconstruction}




## Locations
##Originally Noemie had different tree locaitons, World, Thailand, etc for SarsCov2
##We will look at serotypes instead here
locations = c("Dengue_1", "Dengue_2", "Dengue_3", "Dengue_4")

#optionally load in new trees via readnexus here and process as we did at the begining:
# Example: tree_DENV1_world = read.nexus("~/Library/CloudStorage/OneDrive-NationalInstitutesofHealth/R/Pylowave Dengue/1_Data/1_5_DENV/best_tree_d1_Lin.nexus")

tree_DENV1_thai <-  tree_DENV1
tree_DENV2_thai <-  tree_DENV2
tree_DENV3_thai <-  tree_DENV3
tree_DENV4_thai <-  tree_DENV4



## Timed-tree
tree_DENV_all = list(tree_DENV1_thai,tree_DENV2_thai, tree_DENV3_thai, tree_DENV4_thai
)
names(tree_DENV_all) = locations

## Names all sequences
names_seqs_all = list(names_seqs_1,names_seqs_2, names_seqs_3, names_seqs_4)
names(names_seqs_all) = locations

## Collection times of all sequences
times_seqs_all = list(times_seqs_1, times_seqs_2, times_seqs_3, times_seqs_4)

names(times_seqs_all) = locations


#####01/24/25  TODO Are we sure this is the right/best metadata version? Length is diffent than tree tips
# 
# 
# ## Correspondences Virus ID // EPI ID (GISAID)
# correspondence_ID_EPI_Denv_1 = read.csv("~/Library/CloudStorage/OneDrive-NationalInstitutesofHealth/Collaborators_Data/denv_seqs_wgs/whole_genome/d1_n1026_geneious_metadata.csv", header = T, 
#                                        col.names = c("Name",	"Created Date",	"Description",	"Imported From: Filename",	"Imported From: Path",	"Sequence", "A", "B"))
# 
# correspondence_ID_EPI_Denv_2 = read.csv("~/Library/CloudStorage/OneDrive-NationalInstitutesofHealth/Collaborators_Data/denv_seqs_wgs/whole_genome/d2_n796_geneious_metadata.csv", header = T, 
#                                        col.names = c("Name",	"Created Date",	"Description",	"Imported From: Filename",	"Imported From: Path",	"Sequence", "A", "B"))
# 
# correspondence_ID_EPI_Denv_3 = read.csv("~/Library/CloudStorage/OneDrive-NationalInstitutesofHealth/Collaborators_Data/denv_seqs_wgs/whole_genome/d3_n625_degap_geneious_metadata.csv", header = T, 
#                                        col.names = c("Name",	"Created Date",	"Description",	"Imported From: Filename",	"Imported From: Path",	"Sequence", "A", "B"))
# 
# correspondence_ID_EPI_Denv_4 = read.csv("~/Library/CloudStorage/OneDrive-NationalInstitutesofHealth/Collaborators_Data/denv_seqs_wgs/whole_genome/d4_n477_geneious_metadata.csv", header = T, 
#                                        col.names = c("Name",	"Created Date",	"Description",	"Imported From: Filename",	"Imported From: Path",	"Sequence", "A", "B"))
# 
# 
# correspondence_ID_EPI_all <- list(correspondence_ID_EPI_Denv_1, 
#                                   correspondence_ID_EPI_Denv_2, 
#                                   correspondence_ID_EPI_Denv_3, 
#                                   correspondence_ID_EPI_Denv_4)
# 
# # Apply transformations to all data frames in the list, create longname and make it look liek the vcf file names
# correspondence_ID_EPI_all <- lapply(correspondence_ID_EPI_all, function(df) {
#   # Create the 'longname' column dynamically
#   df$longname <- gsub(" ", "_", paste(df$Name, df$Description, sep = "_"))
#   
#   # Reorder columns to put 'longname' first
#   df <- df[, c("longname", setdiff(names(df), "longname"))]
#   
#   return(df)
# })
# 
# names(correspondence_ID_EPI_all) <- locations



## Datasets with nodes and tips
dataset_with_nodes_all = list(dataset_with_nodes_1,
                              dataset_with_nodes_2,
                              dataset_with_nodes_3,
                              dataset_with_nodes_4)

names(dataset_with_nodes_all) = locations
```

Known issue: need to add underscoreses to the headers of the fasta
files, and rebuild the vcf files. For some reason the R and the python
code I use to find the VCF snp matrix does not like the spaces in the
fasta names. This could be addressed in python or R. I did it in
texteditor because it's a simple find and replace. You will only have to
do this once.

## Generating vcf files from fasta alignments

Shown for the sake of example, outputs are already in the data folder

```{r example Vcf from fasta , eval=FALSE}
  
# Example: define gene start-end ranges (in amino acid positions)
gene_positions <- list(
  C = c(start = 1, end = 342),
  prM = c(start = 343, end = 840),
  E = c(start = 841, end = 2325),
  NS1 = c(start = 2326, end = 3381),
  NS2A = c(start = 3382, end = 4038), #Some insertions
  NS2B = c(start = 4039, end = 4428),
  NS3 = c(start = 4429, end = 6285),
  NS4A = c(start = 6286, end = 6666),
  frag2k = c(start = 6667, end = 6735),
  NS4B = c(start = 6736, end = 7482),
  NS5 = c(start = 7483, end = 10188),
  whole_genome = c(start = 1, end = 10188)
)

file_mapping <- list(
  "1" = list(
    fasta = here("1_Data", "1_5_DENV", "Seq_vcf_data", "denv_seqs_wgs", "whole_genome", "d1_n1026_underscore.fas"),
    csv   = here("1_Data", "1_5_DENV", "Seq_vcf_data", "denv_seqs_wgs", "whole_genome", "d1_n1026_geneious_metadata.csv")
  ),
  "2" = list(
    fasta = here("1_Data", "1_5_DENV", "Seq_vcf_data", "denv_seqs_wgs", "whole_genome", "d2_n796_underscore.fas"),
    csv   = here("1_Data", "1_5_DENV", "Seq_vcf_data", "denv_seqs_wgs", "whole_genome", "d2_n796_geneious_metadata.csv")
  ),
  "3" = list(
    fasta = here("1_Data", "1_5_DENV", "Seq_vcf_data", "denv_seqs_wgs", "whole_genome", "d3_n625_degap_underscore.fas"),
    csv   = here("1_Data", "1_5_DENV", "Seq_vcf_data", "denv_seqs_wgs", "whole_genome", "d3_n625_degap_geneious_metadata.csv")
  ),
  "4" = list(
    fasta = here("1_Data", "1_5_DENV", "Seq_vcf_data", "denv_seqs_wgs", "whole_genome", "d4_n477_underscore.fas"),
    csv   = here("1_Data", "1_5_DENV", "Seq_vcf_data", "denv_seqs_wgs", "whole_genome", "d4_n477_geneious_metadata.csv")
  )
)

for (serotype in 1:4) {
  fasta_path <- file_mapping[[as.character(serotype)]]$fasta
  out_vcf_dir  <- here("1_Data", "1_5_DENV", "Seq_vcf_data", "denv_seqs_wgs", "vcf_from_fasta", "Test")
  
  create_vcf_from_fasta(fasta_path = fasta_path, 
                        gene_positions = gene_positions, 
                        output_path = out_vcf_dir, 
                        serotype = serotype)
  
}


# # Outputs are here: 
#   "1_Data/1_5_DENV/Seq_vcf_data/denv_seqs_wgs/vcf_from_fasta/d1.vcf",
#   "1_Data/1_5_DENV/Seq_vcf_data/denv_seqs_wgs/vcf_from_fasta/d4.vcf",
#   "1_Data/1_5_DENV/Seq_vcf_data/denv_seqs_wgs/vcf_from_fasta/d3.vcf",
#   "1_Data/1_5_DENV/Seq_vcf_data/denv_seqs_wgs/vcf_from_fasta/d2.vcf"



```

### New Code Load in reconstructions from 2_5_reconstruct_AA_changes

```{r , eval=TRUE}
#Loading In reconstrcutions 


# Example: define gene start-end ranges (in amino acid positions)
gene_positions <- list(
  C = c(start = 1, end = 342),
  prM = c(start = 343, end = 840),
  E = c(start = 841, end = 2325),
  NS1 = c(start = 2326, end = 3381),
  NS2A = c(start = 3382, end = 4038), #Some insertions
  NS2B = c(start = 4039, end = 4428),
  NS3 = c(start = 4429, end = 6285),
  NS4A = c(start = 6286, end = 6666),
  frag2k = c(start = 6667, end = 6735),
  NS4B = c(start = 6736, end = 7482),
  NS5 = c(start = 7483, end = 10188),
  whole_genome = c(start = 1, end = 10188)
)

gene_df <- do.call(rbind, lapply(names(gene_positions), function(name) {
  data.frame(
    Protein = name,
    Genome_Start = gene_positions[[name]]["start"],
    Genome_End = gene_positions[[name]]["end"]
  )
}))

# Directory where all VCFs are stored
vcf_dir <- "1_Data/1_5_DENV/Seq_vcf_data/denv_seqs_wgs/vcf_from_fasta/"

# List all .vcf files in that directory
vcf_files <- list.files(vcf_dir, pattern = "\\.vcf$", full.names = TRUE)
vcf_files <- vcf_files[grepl("_", basename(vcf_files))] #removing other vcfs in directory not needed

# create names for the list (e.g., "serotype_1_C", "serotype_3_NS5", etc.)
vcf_keys <- str_remove(basename(vcf_files), "\\.vcf$")  # Remove .vcf extension

# Initialize list to store results
vcf_data <- list()

# Loop through and read
for (i in seq_along(vcf_files)) {
  file_path <- vcf_files[i]
  key <- vcf_keys[i]
  
  cat("Processing:", key, "\n")
  
  data_vcf <- read.vcfR(file_path)
  fixed <- as.data.frame(data_vcf@fix)
  gt <- as.data.frame(data_vcf@gt)
  combined_data <- cbind(fixed, gt)
  
  vcf_data[[key]] <- combined_data
}

# Extract the serotype part of each key (e.g., "serotype_1")
serotype_groups <- str_extract(names(vcf_data), "serotype_\\d+")

# Use split() to group by serotype
vcf_data_by_serotype <- split(vcf_data, serotype_groups)




#Because Noemie's vcf files were made in R they can be parsed as csvs. Consider if using noemie vcf script:
# data_vcf_world_M = read.csv(file = "2_analysis_index/3_snp_association/vcfs_and_AA_reconstructions/World/nextstrain_ncov_gisaid_global_all-time_timetree_M.vcf", sep = '\t')

## Reconstructions, setting the idx_min or the earliest observed sample in the dataset. Reference for aligning time points.
idx_mins <- list()

for (x in seq_along(dataset_with_nodes_all)) {
  # current iteration
  current_dataset <- dataset_with_nodes_all[[x]]
  
  # Calculate idx_min
  idx_min <- which.min(current_dataset$time[which(current_dataset$is.node == 'no')])
  
  # Dynamically name
  idx_min_name <- paste0("idx_min_", x)
  
  # Assign
  assign(idx_min_name, idx_min)
  
  #Add to list
  idx_mins[[idx_min_name]] <- idx_min
}

for (loc in locations) {
  current_dataset <- dataset_with_nodes_all[[loc]]  
  
  if (!is.null(current_dataset)) {
    idx_min <- which.min(current_dataset$time[which(current_dataset$is.node == 'no')])
    idx_mins[[loc]] <- idx_min
  }
}


reconstruction_locations <- c("DENV1", "DENV2", "DENV3", "DENV4")
# Define reconstruction file paths dynamically

orf_mapping <- as.list(names(gene_positions))
names(orf_mapping) <- names(gene_positions)

reconstruction_files <- list()
for (loc in reconstruction_locations) {
  for (orf in names(orf_mapping)) {
    file_path <- here("4_Output_Data", "4_5_DENV", paste0(loc, "_Thai_timetree_reconstruction_", orf,"06_30_25.rds"))
    reconstruction_files[[paste0(loc, "_", orf)]] <- file_path
  }
}

# Load reconstruction data
reconstructions <- list()
for (key in names(reconstruction_files)) {
  file_path <- reconstruction_files[[key]]
  
  if (file.exists(file_path)) {
    reconstructions[[key]] <- readRDS(file_path)
  } else {
    message(paste("Skipping missing file:", file_path))
  }
}



# Fix names to match rest of vars
names(reconstructions) <- gsub("^DENV(\\d+)", "Dengue_\\1", names(reconstructions))


# Update the inferred reconstruction for each (location, ORF)
for (loc in locations) {
  for (orf in names(orf_mapping)) {
    key <- paste0(loc, "_", orf)
 
    if (!is.null(reconstructions[[key]]) &&
        !is.null(idx_mins[[loc]])) {
      current_reconstruction <- reconstructions[[key]]
      current_tree <- tree_DENV_all[[loc]]  # Assuming tree_DENV_all is a named list
      current_idx_min <- idx_mins[[loc]]
      
      # Update inferred codon reconstruction at the MRCA
      current_reconstruction$dataset_with_inferred_reconstruction_codon[current_tree$Nnode + 2, 5:ncol(current_reconstruction$dataset_with_inferred_reconstruction_codon)] <-
        current_reconstruction$dataset_with_inferred_reconstruction_codon[current_idx_min, 5:ncol(current_reconstruction$dataset_with_inferred_reconstruction_codon)]
      
      # Update inferred amino acid reconstruction at the MRCA
      current_reconstruction$dataset_with_inferred_reconstruction_AA[current_tree$Nnode + 2, 5:ncol(current_reconstruction$dataset_with_inferred_reconstruction_AA)] <-
        current_reconstruction$dataset_with_inferred_reconstruction_AA[current_idx_min, 5:ncol(current_reconstruction$dataset_with_inferred_reconstruction_AA)]
      
      # Extract possible SNPs
      current_reconstruction$possible_snps <- colnames(current_reconstruction$dataset_with_inferred_reconstruction_codon)[-(1:4)]
  
    
    
    
    # Store updated reconstruction
    reconstructions[[key]] <- current_reconstruction
    }
  }
}






```

## association_scores

Use the function *association_scores_per_group* to compute the
association score of each mutation to each group. The function takes in
entry:

-   *dataset_with_nodes*
-   *dataset_with_inferred_reconstruction*: a dataframe with the same
    first columns as *dataset_with_nodes* and then one column per snp
    its ancestral reconstruction along all nodes
-   *tree*: timed tree
-   *possible_snps*: the list of mutations that need to be considered
-   *upstream_window*: for each group, how far upstream to consider
    mutations. Not the whole branch to the root, but some amount of time
-   *downstream_window*: for each group, consider nodes from the MRCA up
    to this time. Sometimes there are heterogeneous groups that aquire
    mutations way later (not lineage causing, but still would be lineage
    defining)

The codes to perform this analysis on the SARS-CoV-2 data is available
in the file *2_4_Lineage_Defining_mutations.R*, in the folder
`2_Functions`.

1/25/25: TODO Determining upsteam_window and downstream window. Kept as
sars defaults

```{r SNP_associations}
########################################################################################################################################
## For each SNP, look at defining mutation of each group
########################################################################################################################################

## Play around with these
##Upstream: 5 or 10 years?
##Downstream 100 years? 

## Time window
upstream_window = 10 # years (not going to the root)
downstream_window = 200 # years (considering all the group)


```

```{r scores}
levels <- c("codon", "AA")
scores_all <- list()

for (recon_name in names(reconstructions)) {
  
  for (level in levels) {
    recon_obj <- reconstructions[[recon_name]]
    cat("For:", recon_name, ",level", level, "\n")
    # Extract virus type and gene
    parts <- strsplit(recon_name, "_", fixed = TRUE)[[1]]
    virus <- paste(parts[1], parts[2], sep = "_")  # e.g., "Dengue_1"
    gene <- paste(parts[3:length(parts)], collapse = "_")  # in case gene name has underscore

    scores_all[[paste0(virus, "_", gene, "_", level)]] <- association_scores_per_group(
      dataset_with_nodes = dataset_with_nodes_all[[virus]],
      dataset_with_inferred_reconstruction = recon_obj[[paste0("dataset_with_inferred_reconstruction_", level)]],
      tree = tree_DENV_all[[virus]],
      possible_snps = recon_obj$possible_snps,
      upstream_window = upstream_window,
      downstream_window = downstream_window
    )
  }
}

saveRDS(scores_all, file= here("4_Output_Data", "4_5_DENV", paste0("scores_all_uw_", upstream_window,"_dw_", downstream_window, "_", st,".rds")))

```

## Plot scores

```{r}
tidy_group_scores <- function(group_list, virus = "Dengue_4", gene = "whole_genome", level = "AA") {
  purrr::imap_dfr(group_list, function(group_scores, group_index) {
    tibble(
      group = paste0("Group_", group_index),
      site = as.numeric(sapply(strsplit(names(group_scores), "\\|"), `[`, 2)),
      score = as.numeric(group_scores)
    )
  })
}



  

for (i in 1:4) {
  # Example
  
  temp_score<- paste0("Dengue_", i, "_whole_genome_AA")
  
  
  
  df_AA <- tidy_group_scores(scores_all[[temp_score]])
  
  score_by_group_plot <- ggplot(df_AA, aes(x = site * 3 - 1.5, y = score)) +
    geom_line(color = "black") +
    facet_wrap(~ group, scales = "fixed") +
    labs(
      title = paste0("DENV", i , " SNP Association Scores by Lineage Group"),
      x = "Genome Position (nt)",
      y = "Association Score"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      strip.text = element_text(size = 10),
      axis.text = element_text(size = 8)
    )
  
  # Save the static combined plot
  ggsave(filename = paste0("3_Output_Figures/","DENV_", i, "_", "_AA_genome_density_by_Group_", st, ".svg"), plot = score_by_group_plot, width = 12, height = 6)
  
  score_by_group_plotly <- ggplotly(score_by_group_plot)
  # Save the interactive plot as an HTML file
  htmlwidgets::saveWidget(score_by_group_plotly, file = paste0("3_Output_Figures/","DENV_", i, "_AA_genome_density_by_Group", st,".html"))
  
  }




```

## Find significant snps

When doing reconstruciton, can set some nodes in the branch that have
high levels of support to NA. Since sig_threshold is based on the nodes,
this is a limitaiton. Have to have non-NA on the branch tip

Noise in sig snp determination: \* phylogeny reconstruction: possible
that some members don't have the snp

Lineage defining is: \* present in 80% of groups and not present in
parent nodes. \*

```{r Find significant snps}

########################################################################################################################################
## Find significant snps
########################################################################################################################################
sig_threshold <- 0.70 #TODO Where to set this. This is percentage of all nodes in group that have mutation. Set at 80 because made sense to us.
# Eval 50% - 100%? 
scores_sig_all <- list()

for (score_name in names(scores_all)) {
  # Example: "Dengue_1_NS5_codon"
  parts <- strsplit(score_name, "_")[[1]]
  # Dengue type number (e.g., 1, 2, 3)
  dengue_type <- parts[2]
  
  # Get the appropriate split object for this DENV type
  split_obj <- get(paste0("split_", dengue_type))
  
  # Set up edge lineage tree with node group mapping
  edge_lineage_tree <- split_obj$lineage_tree$edge
  edge_lineage_tree[, 1] <- split_obj$tip_and_nodes_groups[match(edge_lineage_tree[, 1], names(split_obj$tip_and_nodes_groups))]
  edge_lineage_tree[, 2] <- split_obj$tip_and_nodes_groups[match(edge_lineage_tree[, 2], names(split_obj$tip_and_nodes_groups))]
  
  # Get the score vector (one element per node)
  scores_vector <- scores_all[[score_name]]
  
  # Initialize result vector for this gene/level
  snps_sig <- vector("list", length(scores_vector))

  for (i in seq_along(scores_vector)) {
    tmp <- as.numeric(edge_lineage_tree[which(edge_lineage_tree[, 2] == i), ])

    if (is.null(tmp) || length(tmp) == 0 || is.na(tmp[1])) {
      ans <- NULL
      des <- NULL
    } else {
      ans <- if (!is.na(tmp[1]) && tmp[1] <= length(scores_vector)) {
        names(which(scores_vector[[tmp[1]]] > sig_threshold))
      } else NULL

      des <- if (!is.na(tmp[2]) && tmp[2] <= length(scores_vector)) {
        names(which(scores_vector[[tmp[2]]] > sig_threshold))
      } else NULL
    }

    snps_sig[[i]] <- if (!is.null(des)) des[!des %in% ans] else NULL
  }

  # Store the result using the same name
  scores_sig_all[[score_name]] <- snps_sig
}

saveRDS(scores_sig_all, file= here("4_Output_Data", "4_5_DENV", paste0("scores_sig_all_uw_", upstream_window,"_dw_", downstream_window, "_ thresh_", sig_threshold, "_", st,".rds")))

```

## Plot the SNPs density of the full genome

```{r}
########################################################################################################################################
## Full genome, substitutions (NS, ie AA change) with signal
########################################################################################################################################

data_orfs_tb <- gene_df

get_genes_for_serotype <- function(df, serotype) {
  df %>%
    # dplyr::filter(Serotype == serotype) %>%
    dplyr::select(gene = Protein, start = Genome_Start, end = Genome_End) %>%
    dplyr::arrange(start)
}
gene_tables <- list(
  Dengue_1 = get_genes_for_serotype(data_orfs_tb, "DENV1"),
  Dengue_2 = get_genes_for_serotype(data_orfs_tb, "DENV2"),
  Dengue_3 = get_genes_for_serotype(data_orfs_tb, "DENV3"),
  Dengue_4 = get_genes_for_serotype(data_orfs_tb, "DENV4")
)


# Pull only full genome AA entries
full_genome_score_names <- grep("whole_genome_AA$", names(scores_all), value = TRUE)

# Initialize an empty list to hold score vectors for each Dengue type
scores_dengue_full_genome_AA <- list()

for (name in full_genome_score_names) {
  # Extract Dengue type (e.g., "Dengue_1")
  dengue_type <- paste(strsplit(name, "_")[[1]][1:2], collapse = "_")

  # Pull scores and take max per site across nodes
  score_matrix <- do.call(rbind, scores_all[[name]])
  max_scores <- apply(score_matrix, MAR = 2, max)

  # Get site positions from names (assuming name format like "X|123")
  site_positions <- as.numeric(sapply(strsplit(names(max_scores), "\\|"), `[`, 2))
  genome_positions <- site_positions * 3 - 1.5  # Codon to nucleotide midpoint

  names(max_scores) <- genome_positions
  scores_dengue_full_genome_AA[[dengue_type]] <- max_scores
}


plot_dengue_genome_AA_density <- function(scores, genes_table, gene_colors = NULL) {
  genes_table <- dplyr::filter(genes_table, tolower(.data$gene) != "whole_genome")
  
  if (is.null(gene_colors)) {
    gene_colors <- grDevices::rainbow(nrow(genes_table), alpha = 0.25)
  }

  par(
  oma = c(0, 0, 0, 0),        # outer margin
  mar = c(3, 4, 1, 1),        # margin (bottom, left, top, right)
  mgp = c(2, 0.5, 0),         # axis title, labels, line spacing
  family = 'Arial',           
  cex.axis = 0.7,             # axis number size
  cex.lab = 0.8,              # axis label size
  cex.main = 0.9,
  cex.sub = 0.8
)

  dens <- density(as.numeric(names(scores)), weights = scores / sum(scores), bw = 50, n = 3000)

  plot(NULL, 
       xlim = c(0, max(genes_table$end) + 1000), 
       ylim = c(0, 1.5*max(dens$y)), 
       bty = 'n', 
       yaxt = 'n', 
       ylab = "Weighted SNP Density",
       xlab = "Genome Postion (nt)")
  axis(2, las = 2)

   for (i in seq_len(nrow(genes_table))) {
    start <- genes_table$start[i]
    end <- genes_table$end[i]
    midpoint <- (start + end) / 2

    # Draw gene block
    polygon(x = c(start, start, end, end), y = c(0, 1, 1, 0), 
            border = NA, col = gene_colors[i])

    # Add gene name label at the center top of the block
    text(x = midpoint, y = max(dens$y) * 1.25, labels = genes_table$gene[i], 
         cex = 0.4, srt = 90, adj = 0)
  }

  # Draw the density plot over top
  polygon(x = c(dens$x, rev(dens$x)), y = c(dens$y, rep(0, length(dens$y))), 
          border = FALSE, col = 'grey30')
}

# Pick colors using MetBrewer (optional)

gene_colors <- MetBrewer::met.brewer(name = "Derain", n = nrow(gene_df))


# Plot for Dengue_1
plot_dengue_genome_AA_density(
  scores = scores_dengue_full_genome_AA[["Dengue_1"]],
  genes_table = gene_tables[["Dengue_1"]],
  gene_colors = adjustcolor(gene_colors, alpha.f = 0.25)
  
)


# Save
pdf(file = here("3_Output_Figures", "Plot_Dengue1_AA_genome_density.pdf"), width = 10, height = 5)
plot_dengue_genome_AA_density(
  scores = scores_dengue_full_genome_AA[["Dengue_1"]],
  genes_table = gene_tables[["Dengue_1"]],
  gene_colors = adjustcolor(gene_colors, alpha.f = 0.25)
)
dev.off()


# Plot for Dengue_2
plot_dengue_genome_AA_density(
  scores = scores_dengue_full_genome_AA[["Dengue_2"]],
  genes_table = gene_tables[["Dengue_2"]],
  gene_colors = adjustcolor(gene_colors, alpha.f = 0.25)
  
)

# Save
pdf(file = here("3_Output_Figures", "Plot_Dengue2_AA_genome_density.pdf"), width = 10, height = 5)
plot_dengue_genome_AA_density(
  scores = scores_dengue_full_genome_AA[["Dengue_2"]],
  genes_table = gene_tables[["Dengue_2"]],
  gene_colors = adjustcolor(gene_colors, alpha.f = 0.25)
)
dev.off()


# Plot for Dengue_3
plot_dengue_genome_AA_density(
  scores = scores_dengue_full_genome_AA[["Dengue_3"]],
  genes_table = gene_tables[["Dengue_3"]],
  gene_colors = adjustcolor(gene_colors, alpha.f = 0.25)
  
)

# Save
pdf(file = here("3_Output_Figures", "Plot_Dengue3_AA_genome_density.pdf"), width = 10, height = 5)
plot_dengue_genome_AA_density(
  scores = scores_dengue_full_genome_AA[["Dengue_3"]],
    genes_table = gene_tables[["Dengue_3"]],
  gene_colors = adjustcolor(gene_colors, alpha.f = 0.25)
)
dev.off()


# Plot for Dengue_4
plot_dengue_genome_AA_density(
  scores = scores_dengue_full_genome_AA[["Dengue_4"]],
  genes_table = gene_tables[["Dengue_4"]],
  gene_colors = adjustcolor(gene_colors, alpha.f = 0.25)

)

# Save
pdf(file = here("3_Output_Figures", "Plot_Dengue4_AA_genome_density.pdf"), width = 10, height = 5)
plot_dengue_genome_AA_density(
  scores = scores_dengue_full_genome_AA[["Dengue_4"]],
  genes_table = gene_tables[["Dengue_4"]],
  gene_colors = adjustcolor(gene_colors, alpha.f = 0.25)
)
dev.off()



```

## Plot mutations near tree

```{r tree mutations}

# ########################################################################################################################################
# ## Plot mutations next to tree
# ########################################################################################################################################


plot_dengue_gene_heatmap <- function(dengue_to_plot, gene_to_plot) {
  dengue_to_plot_num <- sub("Dengue_", "", dengue_to_plot)
  
  cols <- get(paste0("colors_groups_", dengue_to_plot_num))
  name_groups <- get(paste0("name_groups_", dengue_to_plot_num))
  times_seqs_to_plot <- get(paste0("times_seqs_", dengue_to_plot_num))
  tree_to_plot <- tree_DENV_all[[dengue_to_plot]]
  node_data_to_plot <- dataset_with_nodes_all[[dengue_to_plot]]
  node_data_to_plot$group_str <- as.character(node_data_to_plot$groups)
  
  names(cols) <- as.character(1:max(as.numeric(name_groups)))
  col_AA <- c('white', cols, 'grey')
  names(col_AA) <- c(0, as.character(1:max(as.numeric(name_groups))), 100)
  
  plot_dataset <- reconstructions[[paste0(dengue_to_plot, "_", gene_to_plot)]]$dataset_with_inferred_reconstruction_AA
  plot_score_list <- scores_sig_all[[paste0(dengue_to_plot, "_", gene_to_plot, "_AA")]]
  
  num_seqs <- sum(plot_dataset$is.node == "no")
  seq_names <- plot_dataset$name_seq[plot_dataset$is.node == "no"]
  plot_AA_mat_total <- matrix(rep(100, num_seqs), ncol = 1)
  rownames(plot_AA_mat_total) <- seq_names
  
  colnames(plot_AA_mat_total) <- ""
  
  for (group_to_plot in seq_along(plot_score_list)) {
    AAs <- plot_score_list[[group_to_plot]]
    if (length(AAs) >= 1) {
      AAs_pos <- as.numeric(sapply(AAs, function(x) str_split(x, '\\|')[[1]][2]))
      AAs_ref <- sapply(AAs, function(x) str_split(str_split(x, '\\|')[[1]][1], ':')[[1]][2])
      AAs_var <- sapply(AAs, function(x) str_split(x, '\\|')[[1]][3])
      
      AA_mat <- matrix(
        plot_dataset[plot_dataset$is.node == "no", which(colnames(plot_dataset) == AAs_pos[1])],
        ncol = 1
      )
      AA_mat[AA_mat[,1] != AAs_ref[1] & AA_mat[,1] != AAs_var[1], 1] <- NA
      AA_mat[AA_mat[,1] == AAs_var[1], 1] <- group_to_plot
      AA_mat[AA_mat[,1] == AAs_ref[1], 1] <- 0
      
      if (length(AAs) > 1) {
        for (i in 2:length(AAs)) {
          next_col <- plot_dataset[plot_dataset$is.node == "no", which(colnames(plot_dataset) == AAs_pos[i])]
          next_col[next_col != AAs_ref[i] & next_col != AAs_var[i]] <- NA
          next_col[next_col == AAs_var[i]] <- group_to_plot
          next_col[next_col == AAs_ref[i]] <- 0
          AA_mat <- cbind(AA_mat, next_col)
        }
      }
      
      rownames(AA_mat) <- seq_names
      labels_for_cols <- AAs

      if (tolower(gene_to_plot) == "whole_genome") {
        # Map genome AA position -> (gene, gene-relative AA position)
        gt <- gene_tables[[dengue_to_plot]] |> 
          dplyr::filter(tolower(gene) != "whole_genome") |> 
          dplyr::mutate(
            aa_start = ceiling(start / 3),
            aa_end   = floor(end / 3)
          )
        
        # Build gene-aware labels
        labels_for_cols <- vapply(AAs, FUN.VALUE = character(1), function(tag) {
          parts <- strsplit(tag, "\\|")[[1]]           # c(ref, pos, var)
          aa_pos <- as.numeric(parts[2])
          
          # Find which gene this AA position falls into
          hit <- which(gt$aa_start <= aa_pos & aa_pos <= gt$aa_end)
          if (length(hit) == 1) {
            rel_pos <- aa_pos - gt$aa_start[hit] + 1
            gene_nm <- gt$gene[hit]
            # "S|343|T — prM: S|1|T"
            paste0(
              parts[1], "|", parts[2], "|", parts[3],
              " \u2014 ", gene_nm, ": ",
              parts[1], "|", rel_pos, "|", parts[3]
            )
          } else {
            # Fallback if no gene match (shouldn't happen, but be safe)
            tag
          }
        })
      }
      
      colnames(AA_mat) <- paste0(group_to_plot, ":", labels_for_cols)
      
      # Add spacer after setting colnames
      spacer_col <- matrix(100, nrow = nrow(AA_mat), ncol = 1)
      AA_mat <- cbind(AA_mat, spacer_col)
      colnames(AA_mat)[ncol(AA_mat)] <- group_to_plot
      
      # Append to master matrix
      plot_AA_mat_total <- cbind(plot_AA_mat_total, AA_mat)
    }
  }
  
  plot_tree <- ggtree(tree_to_plot, mrsd = lubridate::date_decimal(max(times_seqs_to_plot)), size = 0.08) %<+%
    node_data_to_plot +
    aes(color = node_data_to_plot$group_str) +
    scale_color_manual(values = cols) +
    ggplot2::ylim(-0.5, NA) +
    coord_cartesian(clip = "off")

  
  title_label <- paste0(dengue_to_plot, " | ", gene_to_plot)
  
  
  heatmap_plot <- gheatmap(plot_tree,
                           plot_AA_mat_total,
                           offset = 0.1,
                           width = 10,
                           colnames = TRUE,
                           colnames_position = "top",
                           colnames_angle = 90,
                           hjust = 0,
                           font.size = 2.5,
                           legend_title = "Group",
                           color = NA) +
    scale_fill_manual(values = col_AA, na.value = NA) +
    coord_cartesian(clip = "off") +
    ggtitle(title_label) +
    theme(
      legend.position = "none",
      plot.title = element_text(hjust = 0.1, size = 12, face = "bold"),
      plot.title.position = "panel"
    )
  
  print(heatmap_plot)
  
  ggsave(
    filename = here("3_Output_Figures", paste0("Tree_", dengue_to_plot, "_defining_", gene_to_plot, "_mutations_", st, ".pdf")),
    plot = heatmap_plot,
    device = 'pdf',
    scale = 1,
    width = 25,
    height = 18,
    units = 'cm'
  )
}

genes_to_plot <- names(gene_positions)
for(gene in genes_to_plot){
  cat("Plotting:" ,gene, "\n")
  try({
plot_dengue_gene_heatmap("Dengue_4", gene)
}, silent = TRUE )
}


plot_dengue_gene_heatmap("Dengue_4", "E")


scores_sig_all[["Dengue_1_E_AA"]]
scores_sig_all[["Dengue_2_E_AA"]]
scores_sig_all[["Dengue_3_E_AA"]]
scores_sig_all[["Dengue_4_E_AA"]]

  
```

Whole genome plot with protien labels

```{r}

```

```{r examining alignments, eval=FALSE}
# The matrix to search
aa_matrix <- list_matrices_ORFs_AA_3$whole_genome

# The motif as a vector of characters
motif <- strsplit("MRCVG", "")[[1]] #The lead of DENV3 E
motif_length <- length(motif)

# Number of columns we can check (to stay within bounds)
max_start <- ncol(aa_matrix) - motif_length + 1

# Store matches: list of positions where motif is found
matches <- list()

for (col_start in 1:max_start) {
  cols_to_check <- col_start:(col_start + motif_length - 1)
  
  # Subset the matrix columns for this window
  window <- aa_matrix[, cols_to_check]
  
  # Check which rows match the motif
  row_match <- apply(window, 1, function(x) all(x == motif))
  
  # If any row matches, save the column index and the row names
  if (any(row_match)) {
    matches[[as.character(col_start)]] <- rownames(aa_matrix)[row_match]
  }
}

# Display results
matches


# Store starting column indices where motif is found in any row
match_columns <- c()

for (col_start in 1:max_start) {
  cols_to_check <- col_start:(col_start + motif_length - 1)
  
  # Subset window of the matrix
  window <- aa_matrix[, cols_to_check]
  
  # Check if any row matches the motif across these columns
  match_found <- apply(window, 1, function(row_seq) all(row_seq == motif))
  
  if (any(match_found)) {
    match_columns <- c(match_columns, col_start)
  }
}

# Result: column indices where "MRCVG" starts
match_columns
```

# Up/downstream window and sig threshohold testing

```{r window and sig test, eval = FALS}

#Saving imporntant variables to resart R and make this faster
saveRDS(reconstructions, file= here("6_Session_Data", paste0("reconstructions", st,".RData")))
saveRDS(dataset_with_nodes_all, file= here("6_Session_Data", paste0("dataset_with_nodes_all", st,".RData")))
saveRDS(tree_DENV_all, file= here("6_Session_Data", paste0("tree_DENV_all", st,".RData")))


run_single_combination <- function(upstream_window, downstream_window,
                                   reconstructions_local, dataset_with_nodes_local, tree_local) {
  tryCatch({

    cat("Running: uw =", upstream_window, ", dw =", downstream_window, "\n")

    levels <- c("codon", "AA")
    scores_all <- list()

    for (recon_name in names(reconstructions_local)) {
      parts <- strsplit(recon_name, "_", fixed = TRUE)[[1]]

      # Validate expected structure: Dengue_1_NS3 → 3+ parts
      if (length(parts) < 3) {
        warning(paste("Skipping malformed recon_name:", recon_name))
        next
      }

      virus <- paste(parts[1:2], collapse = "_")
      gene <- paste(parts[3:length(parts)], collapse = "_")
      recon_obj <- reconstructions_local[[recon_name]]

      for (level in levels) {
        dataset_key <- paste0("dataset_with_inferred_reconstruction_", level)

        if (!dataset_key %in% names(recon_obj)) {
          warning(paste("Missing", dataset_key, "in", recon_name, "- skipping"))
          next
        }

        scores_all[[paste0(virus, "_", gene, "_", level)]] <- association_scores_per_group(
          dataset_with_nodes = dataset_with_nodes_local[[virus]],
          dataset_with_inferred_reconstruction = recon_obj[[dataset_key]],
          tree = tree_local[[virus]],
          possible_snps = recon_obj$possible_snps,
          upstream_window = upstream_window,
          downstream_window = downstream_window,
          virus = virus,
          gene = gene,
          level = level
        )
      }
    }

    saveRDS(scores_all, file = here("4_Output_Data", "4_5_DENV",
      paste0("scores_all_uw_", upstream_window, "_dw_", downstream_window, ".RData")))

  }, error = function(e) {
    message("Error in run uw=", upstream_window, ", dw=", downstream_window, ": ", e$message)
    return(NULL)
  })
}


# 
# run_single_combination(
#     upstream_window = param_list[[1]]$upstream_window,
#     downstream_window = param_list[[1]]$downstream_window,
#     sig_threshold = 0.8,  # placeholder; unused here
#     reconstructions_local = reconstructions,
#     dataset_with_nodes_local = dataset_with_nodes_all,
#     tree_local = tree_DENV_all
#   )
# 



# Generate all combinations of parameter values
param_grid <- expand.grid(
  upstream_window = seq(2, 20, by = 2),           # 2, 4, ..., 20
  downstream_window = seq(10, 100, by = 10)     # 10, 20, ..., 100
  
)


nrow(param_grid)


# Prepare the parameter combinations as a list of lists
param_list <- split(param_grid, seq(nrow(param_grid)))


#Load RData back in
reconstructions <- readRDS("6_Session_Data/reconstructions2025-07-03_0910.RData")
dataset_with_nodes_all <- readRDS("6_Session_Data/dataset_with_nodes_all2025-07-03_0910.RData")
tree_DENV_all <- readRDS("6_Session_Data/tree_DENV_all2025-07-03_0910.RData")


library(future)
library(future.apply)

plan(multisession) 
options(future.globals.maxSize = 2 * 1024^3)  # 2 GB

# future::plan(sequential)
# gc()

library(progressr)
handlers(global = TRUE)
handlers("txtprogressbar")

with_progress({
# Run in parallel with future_lapply
future_lapply(param_list, function(params) {
  run_single_combination(
    upstream_window = params$upstream_window,
    downstream_window = params$downstream_window,
    reconstructions_local = reconstructions,
    dataset_with_nodes_local = dataset_with_nodes_all,
    tree_local = tree_DENV_all
  )
},
future.packages = c("vcfR", "tidyverse", "here"))
})



```

## Testing Sig Thresholds

```{r testing sig thresholds, eval = FALSE}

# Generate all combinations of parameter values, now with sig thresholds
param_grid_sig <- expand.grid(
  upstream_window = seq(2, 20, by = 2),           # 2, 4, ..., 20
  downstream_window = seq(10, 100, by = 10),      # 10, 20, ..., 100
  sig_threshold = seq(0.5, 1.0, by = 0.1)          # 0.5, 0.6, ..., 1.0
)
nrow(param_grid_sig)

#Split to list
param_list_sig <- split(param_grid_sig, seq(nrow(param_grid_sig)))
  
  
run_single_combination_sig <- function(param, splits){
  
  uw <- param$upstream_window
  dw <- param$downstream_window
  sig_threshold <- param$sig_threshold
  
  tryCatch({
    cat("→ Processing uw =", uw, ", dw =", dw, ", threshold =", sig_threshold, "\n")
    
    # Load scores_all file for this window
    scores_path <- here("4_Output_Data", "4_5_DENV", 
                        paste0("scores_all_uw_", uw, "_dw_", dw, ".RData"))
    if (!file.exists(scores_path)) {
      cat("⚠️ File not found:", scores_path, "\n")
      return(NULL)
    }
    
    scores_all <- readRDS(scores_path)
    
    scores_sig_all <- list()
    
    for (score_name in names(scores_all)) {
      # Example: "Dengue_1_NS5_codon"
      parts <- strsplit(score_name, "_")[[1]]
      # Dengue type number (e.g., 1, 2, 3)
      dengue_type <- parts[2]
      
      # Get the appropriate split object for this DENV type
      split_obj <- splits[[dengue_type]]
      
      # Set up edge lineage tree with node group mapping
      edge_lineage_tree <- split_obj$lineage_tree$edge
      edge_lineage_tree[, 1] <- split_obj$tip_and_nodes_groups[match(edge_lineage_tree[, 1], names(split_obj$tip_and_nodes_groups))]
      edge_lineage_tree[, 2] <- split_obj$tip_and_nodes_groups[match(edge_lineage_tree[, 2], names(split_obj$tip_and_nodes_groups))]
      
      # Get the score vector (one element per node)
      scores_vector <- scores_all[[score_name]]
      
      # Initialize result vector for this gene/level
      snps_sig <- vector("list", length(scores_vector))
      
      for (i in seq_along(scores_vector)) {
        tmp <- as.numeric(edge_lineage_tree[which(edge_lineage_tree[, 2] == i), ])
        
        if (is.null(tmp) || length(tmp) == 0 || is.na(tmp[1])) {
          ans <- NULL
          des <- NULL
        } else {
          ans <- if (!is.na(tmp[1]) && tmp[1] <= length(scores_vector)) {
            names(which(scores_vector[[tmp[1]]] > sig_threshold))
          } else NULL
          
          des <- if (!is.na(tmp[2]) && tmp[2] <= length(scores_vector)) {
            names(which(scores_vector[[tmp[2]]] > sig_threshold))
          } else NULL
        }
        
        snps_sig[[i]] <- if (!is.null(des)) des[!des %in% ans] else NULL
      }
      
      # Store the result using the same name
      scores_sig_all[[score_name]] <- snps_sig
    }
    
    saveRDS(scores_sig_all, file = here("4_Output_Data", "4_5_DENV",
                                        paste0("scores_sig_all_uw_", uw, "_dw_", dw, "_thresh_", sig_threshold, "_", st, ".RData")))
    
    return(paste0("Done: uw=", uw, ", dw=", dw, ", th=", sig_threshold))
    
  }, error = function(e) {
    message("Error in run uw=", uw, ", dw=", dw, ", thresh=", sig_threshold, ": ", e$message)
    return(NULL)
  })
  
}




plan(multisession)

# plan(sequential)

splits <- list(
  `1` = split_1,
  `2` = split_2,
  `3` = split_3,
  `4` = split_4
)

results <- future_map(param_list_sig, 
                      run_single_combination_sig,
                      splits = splits,
                      .progress = TRUE)



```

## Compare files

```{r, eval=FALSE}
summarize_scores_sig <- function(scores_sig_all_file) {
  # Always use readRDS regardless of the file extension
  scores_sig_all <- readRDS(scores_sig_all_file)

  param_info <- stringr::str_match(basename(scores_sig_all_file),
                                   "uw_(\\d+)_dw_(\\d+)_thresh_((?:0\\.\\d+)|(?:1(?:\\.0)?))")
  uw <- as.numeric(param_info[2])
  dw <- as.numeric(param_info[3])
  threshold <- as.numeric(param_info[4])

  purrr::imap_dfr(scores_sig_all, function(gene_list, gene_name_full) {
    virus_gene <- stringr::str_match(gene_name_full, "(Dengue_\\d)_(.*?)_(codon|AA)")
    virus <- virus_gene[2]
    gene <- virus_gene[3]
    level <- virus_gene[4]

    purrr::imap_dfr(gene_list, function(snps, group_index) {
      tibble::tibble(
        uw = uw,
        dw = dw,
        threshold = threshold,
        virus = virus,
        gene = gene,
        level = level,
        group = as.integer(group_index),
        n_snps = length(snps)
      )
    })
  })
}



#Run in parallel
plan(multisession)

all_sig_files <- list.files(here("4_Output_Data", "4_5_DENV","UW_DW_SigThres_Testing"),
                          pattern = "^scores_sig_all_uw_.*\\.RData$",
                          full.names = TRUE)

summary_scores_sig <- future_map_dfr(all_sig_files, summarize_scores_sig, .progress = TRUE)



# Save for later
write_csv(summary_scores_sig, "summary_scores_sig_all.csv")


```

## Plot comparisons

```{r}
p_sig_compare <- summary_scores_sig %>%
  group_by(uw, dw, threshold, virus) %>%
  summarise(total_snps = sum(n_snps), .groups = "drop") %>%
  ggplot(aes(x = threshold, y = total_snps, color = virus)) +
  geom_line() +
  facet_grid(uw ~ dw, labeller = label_both) +
  theme_minimal()


p_sig_compare_labeled <- add_facet_labels(p_sig_compare, xfacet_label = "DW", yfacet_label = "UW")


grid.newpage()
grid.draw(p_sig_compare_labeled )




summary_scores_sig %>%
  filter(gene == "whole_genome") %>%
  group_by(virus, threshold, gene) %>%
  summarise(avg_snps = mean(n_snps), .groups = "drop") %>%
  ggplot(aes(x = threshold, y = virus, fill = avg_snps)) +
  geom_tile() +
  facet_wrap(~ gene) +
  scale_fill_viridis_c() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))





summary_by_params_D1_WG <- summary_scores_sig %>%
  filter(gene == "whole_genome", level == "AA", virus == "Dengue_1") %>% 
  group_by(uw, dw, threshold)

  summary_by_params_D1_WG %>% 
  ggplot(aes(x = threshold, y = n_snps, fill = factor(group))) +
  geom_col(position = "dodge") +
  facet_grid(uw ~ dw, labeller = label_both) +
  theme_minimal()
  
  
summary_by_params_D2_WG <- summary_scores_sig %>%
  filter(gene == "whole_genome", level == "AA", virus == "Dengue_2") %>% 
  group_by(uw, dw, threshold)

  summary_by_params_D2_WG %>% 
  ggplot(aes(x = threshold, y = n_snps, fill = factor(group))) +
  geom_col(position = "dodge") +
  facet_grid(uw ~ dw, labeller = label_both) +
  theme_minimal()

  
summary_by_params_D3_WG <- summary_scores_sig %>%
  filter(gene == "whole_genome", level == "AA", virus == "Dengue_3") %>% 
  group_by(uw, dw, threshold)

  summary_by_params_D3_WG %>% 
  ggplot(aes(x = threshold, y = n_snps, fill = factor(group))) +
  geom_col(position = "dodge") +
  facet_grid(uw ~ dw, labeller = label_both) +
  theme_minimal()

  
  
summary_by_params_D4_WG <- summary_scores_sig %>%
  filter(gene == "whole_genome", level == "AA", virus == "Dengue_4") %>% 
  group_by(uw, dw, threshold)

  summary_by_params_D4_WG %>% 
  ggplot(aes(x = threshold, y = n_snps, fill = factor(group))) +
  geom_col(position = "dodge") +
  facet_grid(uw ~ dw, labeller = label_both) +
  theme_minimal()
  
unique(split_2$groups)

unique(split_1$groups)
```

# Save

```{r Save}
save.image(file = here("6_Session_Data", paste0(st, "_Workspace.RData")))
```
