---
title: "Knee Analysis"
author: "Douglas Fritz"
date: "2025-10-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r lib}
library(reticulate)
library(ggplot2)
library(glue)



```


```{r}
# Load in potential splits
potential_splits_1 <- readRDS("4_Output_Data/Max_groups_potential_splits_reasonable_DENV1_timescale7_wind1_2025-10-20_1143.rds")
potential_splits_2 <- readRDS("4_Output_Data/Max_groups_potential_splits_reasonable_DENV2_timescale7_wind1_2025-10-20_1143.rds")
potential_splits_3 <- readRDS("4_Output_Data/Max_groups_potential_splits_reasonable_DENV3_timescale7_wind1_2025-10-20_1143.rds")
potential_splits_4 <- readRDS("4_Output_Data/Max_groups_potential_splits_reasonable_DENV4_timescale7_wind1_2025-10-20_1143.rds")


# pick the env you want to use (this one already exists)
use_virtualenv("~/.virtualenvs/r-kneeliverse", required = TRUE)

py_config()  # should now show .../.virtualenvs/r-kneeliverse/bin/python
py_run_string("import kneeliverse, numpy as np, pandas as pd; print('kneeliverse OK')")



```

# RUN THIS PART MANUALLY STEPPING THROUGH D1-D4

```{r}

output_dir <- "3_Output_Figures/3_5_DENV/3_5_99_Supplement"

tag <- "DENV4"
ps <- potential_splits_reasonable_4
mods <- ps[["best_mod"]]



df <- tibble::tibble(
  step = seq_along(mods),
  # AIC/BIC can still fail on odd models → guard them
  AIC  = map_dbl(mods, possibly(AIC,           NA_real_)),
  BIC  = map_dbl(mods, possibly(BIC,           NA_real_)),
  dev  = {
    if (!is.null(ps$best_summary)) {
      map_dbl(ps$best_summary, ~ (.x$dev.expl %||% NA_real_))
    } else {
      map_dbl(mods, possibly(~ summary(.x)$dev.expl, NA_real_))
    }
  }
) %>%
  mutate(nonexpl = pmax(1 - dev, 0))
```


Consider using Kneeliverse python to detect elbows: 
```{python}
# bring R's df into Python
df = r.df  # reticulate hands us a pandas.DataFrame

# choose which curve to analyze: dev (increasing) or nonexpl (decreasing)
x = df["step"].to_numpy(dtype=float)
y = df["nonexpl"].to_numpy(dtype=float)        # or: df["nonexpl"].to_numpy()

# assemble points (N x 2); x should be increasing
pts = np.column_stack([x, y])
ptslog = np.column_stack([x, np.log10(y)])

# --- kneeliverse: Kneedle (single or multiple knees) ---
import kneeliverse.kneedle as kneedle

# Linear
knees_kneedle_idx = kneedle.knees(pts, t=1.0, sensitivity=1.0, p = 'Kneedle')   # returns integer indices  
knees_sig_idx = kneedle.knees(pts, t=1.0, sensitivity=1.0, p = 'Significant')   # returns integer indices
knees_zscore_idx = kneedle.knees(pts, t=1.0, sensitivity=1.0, p = 'ZScore')   # returns integer indices
multiknee_idx  = kneedle.multi_knee(pts, t1=0.01, t2=3)  # multi_knee variant

# Log
knees_log_idx = kneedle.knees(ptslog, t=1.0, sensitivity=1.0,  p = 'Kneedle')   # returns integer indices
knees_log_sig_idx = kneedle.knees(ptslog, t=1.0, sensitivity=1.0, p = 'Significant')   # returns integer indices
knees_log_zscore_idx = kneedle.knees(ptslog, t=1.0, sensitivity=1.0, p = 'ZScore')   # returns integer indices
multiknee_log_idx  = kneedle.multi_knee(ptslog, t1=0.01, t2=3)  # multi_knee variant

# expose back to R
# Linear knees 
r.knees_kneedle_idx = knees_kneedle_idx.astype(int) 
r.knees_sig_idx = knees_sig_idx.astype(int) 
r.knees_zscore_idx = knees_zscore_idx.astype(int) 
r.multi_knee_idx = multiknee_idx.astype(int) 

# log knees
r.knees_log_idx = knees_log_idx.astype(int) 
r.knees_log_sig_idx = knees_log_sig_idx.astype(int) 
r.knees_log_zscore_idx = knees_log_zscore_idx.astype(int) 
r.multiknee_log_idx = multiknee_log_idx.astype(int) 

```


```{r}

knee_steps_kneedle <- df$step[knees_kneedle_idx]
knee_steps_sig <- df$step[knees_sig_idx]
knee_steps_zscore <- df$step[knees_zscore_idx]
knee_steps_multiknee <- df$step[multi_knee_idx]

knee_log_steps_kneedle <- df$step[knees_log_idx]
knee_log_steps_sig <- df$step[knees_log_sig_idx]
knee_log_steps_zscore <- df$step[knees_log_zscore_idx]
knee_log_steps_multiknee <- df$step[multiknee_log_idx]


Lin_plot <- ggplot(df, aes(step, nonexpl)) +
  geom_line() + geom_point() + 
  geom_vline(xintercept = knee_steps_kneedle + 0.25, linetype = 3, colour = "#C8102E") +
  geom_vline(xintercept = knee_steps_sig, linetype = 1, colour = "#c75265") +
  geom_vline(xintercept = knee_steps_zscore - 0.25, linetype = 2, colour = "#c77381") +
  geom_point(
    data = dplyr::filter(df, step %in% knee_steps_multiknee),
    color = "purple", size = 2
  ) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Explained deviance with knees")


Log_plot <- ggplot(df, aes(step, nonexpl)) +
  geom_line() + geom_point() + 
  geom_vline(xintercept = knee_log_steps_kneedle + 0.25, linetype = 3, colour = "#C8102E") +
  geom_vline(xintercept = knee_log_steps_sig, linetype = 1, colour = "#c75265") +
  geom_vline(xintercept = knee_steps_zscore - 0.25, linetype = 2, colour = "#c77381") +
  geom_point(
    data = dplyr::filter(df, step %in% knee_log_steps_multiknee),
    color = "purple", size = 2
  ) +
  scale_y_continuous(labels = scales::percent, 
                     trans  = "log10") +
  labs(title = "Explained deviance with knees (Log)")



Lin_plot
Log_plot

ggsave(file.path(output_dir, glue("{tag}_knees_linear.png")), Lin_plot, width = 11, height = 7, dpi = 300)
ggsave(file.path(output_dir, glue("{tag}_knees_log.png")), Log_plot, width = 11, height = 7, dpi = 300)


```


### STOP Manual RUN ^^^





#Number of Groups Selection Plots (AIC, BIC, Deviance, Knees)

Look at the deviance explained by the models with different number of
groups.To decide on this many groups, we initially ran the algorithm up to 30
groups.

How many lineages to choose is a practical question not something that
is easy to ground in math. Elbow plot analysis is somewhat arbitrary.
Most plots have two elbows: First elbow is the known dynamics, the second elbow would include groups that are more exploratory.
Stopping at the first linear point on horizontal line.

We also Use knees to detect elbow.

###TODO Why is D3 log less than 1 (log1 = 0)

```{r linlog group plot}
max_groups_found_line_3 = 6 #Knees + 1 say 9
max_groups_found_line_2 = 3 #Knees + 1 say 9
max_groups_found_line_1 = 8 #Knees + 1 say 10
max_groups_found_line_4 = 4 #Knees + 1 say 4

for (i in 1:4) {
  
 svglite(sprintf("3_Output_Figures/3_5_DENV/3_5_99_Supplement/DENV_%d_linlog.svg", i), width = 12, height = 6)
  par(mfrow = c(1, 2), oma = c(2, 2, 1, 1), mar = c(2, 2, 2, 0.5),
      mgp = c(0.75, 0.25, 0), cex.axis = 0.5, cex.lab = 0.5,
      cex.main = 0.7, cex.sub = 0.5)
  ## Dynamically retrieve the potential splits for the current group
  potential_splits <- get(paste0('potential_splits_', i))
  max_groups_found_line_i <- get(paste0('max_groups_found_line_', i))
  
  ## Dynamically create or update the data frame for the current group
  df_explained_dev_i <- data.frame(
    'N_groups' = 0:length(potential_splits$best_dev_explained),
    'Non_explained_deviance' = (1 - c(potential_splits$first_dev, potential_splits$best_dev_explained)),
    'Non_explained_deviance_log' = log10(1 - c(potential_splits$first_dev, potential_splits$best_dev_explained))
  )
  
  ## Linear scale plot
  plot(df_explained_dev_i$N_groups,
       df_explained_dev_i$Non_explained_deviance,
       bty = 'n', ylim = c(0, 1),
       xaxt = 'n', yaxt = 'n', pch = 16, 
       main = paste('Linear scale DENV', i), cex = 0.5,
       ylab = 'Non-explained deviance (%)', xlab = 'Number of groups')
  axis(1, lwd = 0.5, tck = -0.02)
  axis(2, las = 2, at = seq(0, 1, 0.1), labels = seq(0, 1, 0.1) * 100, lwd = 0.5, tck = -0.02)
  abline(v = max_groups_found_line_i, col = "purple", lty = 2, lwd = 0.5)
  
  ## Log scale plot
  plot(df_explained_dev_i$N_groups,
       df_explained_dev_i$Non_explained_deviance,
       log = 'y', # This is log10
       ylim = c(0.0001, 1),
       bty = 'n',
       xaxt = 'n', yaxt = 'n', pch = 16, 
       main = paste('Log scale DENV', i), cex = 0.5,
       ylab = 'Non-explained deviance (%) - log scale', xlab = 'Number of groups')
  axis(1, lwd = 0.5, tck = -0.02)
  yticks <- c(0.0001, 0.001, 0.005, 0.01, 0.05, 0.1, 0.25, 0.5, 1)
  yticks <- yticks[yticks >= 0.0001 & yticks <= 1]
  axis(2, las = 2, at = yticks, labels = yticks * 100, lwd = 0.5, tck = -0.02)
  abline(v = max_groups_found_line_i, col = "purple", lty = 2, lwd = 0.5)
  
  dev.off()
}


```

```{r AIC BIC plots}

for (i in 1:4) {
  
  
  potential_splits <- get(paste0('potential_splits_', i))
  mods <- potential_splits[["best_mod"]]
  output_dir <- "3_Output_Figures/3_5_DENV/3_5_99_Supplement"

df <- tibble(
  step = seq_along(mods),
  AIC  = sapply(mods, AIC),
  BIC = sapply(mods, BIC),
  dev  = sapply(mods, function(m) summary(m)$dev.expl)
)

df <- df %>%
  mutate(
    dev_pct  = dev * 100,
    nonexpl  = pmax(1 - dev, 1e-8)
  )

kmin <- df$step[which.min(df$AIC)]
kmin_BIC <- df$step[which.min(df$BIC)]

p1 <- ggplot(df, aes(step, AIC)) +
  geom_line() + geom_point() +
  geom_vline(xintercept = kmin, linetype = 2, color = "#012169aa") +
  labs(x = "Number of groups (step index)", y = "AIC",
       title = "Model AIC across tested splits",
       subtitle = sprintf("Min AIC at k=%d (AIC=%.1f)", kmin, df$AIC[kmin]))

p2 <- ggplot(df, aes(step, dev_pct)) +
  geom_line() + geom_point() +
  labs(x = "Number of groups (step index)", y = "Deviance explained (%)",
       title = glue("DENV{i} Deviance across tested splits"))


p3 <- ggplot(df, aes(step, BIC)) +
  geom_line() + geom_point() +
  geom_vline(xintercept = kmin_BIC, linetype = 2, color = "#C8102Eaa") +
  labs(x = "Number of groups (step index)", y = "BIC",
       title = glue("DENV{i} Model BIC across tested splits"),
       subtitle = sprintf("Min BIC at k=%d (BIC=%.1f)", kmin_BIC, df$BIC[kmin_BIC])) 

p4 <- ggplot(df, aes(step)) +
  geom_line(aes(y = AIC), colour = "#012169FF") + geom_point(aes(y = AIC), colour = "#012169FF") +
  geom_line(aes(y = BIC), colour = "#C8102EFF") + geom_point(aes(y = BIC), colour = "#C8102EFF") +
  geom_vline(xintercept = kmin, linetype = 2, color = "#012169aa") +
  geom_vline(xintercept = kmin_BIC, linetype = 2, color = "#C8102Eaa") +
  labs(x = "Number of groups (step index)", y = "Value",
       title = glue("DENV{i} Model AIC & BIC across tested splits"),
       subtitle = glue::glue("Min AIC at k={kmin} (AIC={round(df$AIC[kmin],1)})\n",
                      "Min BIC at k={kmin_BIC} (BIC={round(df$BIC[kmin_BIC],1)})") )


p6 <- ggplot(df, aes(step, nonexpl)) +
  geom_line() + geom_point() +
  scale_y_continuous(
    trans  = "log10",
    breaks = scales::log_breaks(n = 6),                       # data-aware log ticks
    labels = function(x) scales::percent(x, accuracy = 0.1)   # show as %
  ) +
  labs(x = "Number of groups (step index)", y = "Non-explained Deviance (%) - log scale",
       title = glue("DENV{i} Non-explained Deviance explained across tested splits (log)"))

p7 <- ggplot(df, aes(step, nonexpl*100)) +
  geom_line() + geom_point() +
  labs(x = "Number of groups (step index)", y = "Non-explained Deviance (%)",
       title = glue("DENV{i} Non-explained Deviance explained across tested splits"))



p8 <- p2 + p6 + p7



p5 <- p4 + p6 + p7

print(p1); print(p2); print(p3); print(p4); print(p5); print(p6); print(p7); print(p8)

# save if you want

# ggsave(glue("{output_dir}/aic_by_step_DENV{i}.png"), p1, width = 6, height = 4, dpi = 300)
# ggsave(glue("{output_dir}/devexp_by_step_DENV{i}.png"), p2, width = 6, height = 4, dpi = 300)
# ggsave(glue("{output_dir}/bic_by_step_DENV{i}.png"), p3, width = 6, height = 4, dpi = 300)
ggsave(glue("{output_dir}/AIC_BIC_devexp_by_step_DENV{i}.png"), p5, width = 8, height = 11, dpi = 300)
ggsave(glue("{output_dir}/Devexp_by_step_DENV{i}.png"), p8, width = 8, height = 11, dpi = 300)

}

```

Optimize the number of groups: set the minimum number of sequences per
group to 30, with a minimum frequency of 1%.

#TODO: Decide if we should do more groups, and do sensitivty analysis on
differnt numbers of groups


