---
author: "Douglas Fritz"
date: "2025-10-13"
# Ref: from Noémie Lefrancq https://www.medrxiv.org/content/10.1101/2023.12.23.23300456v1.full
output: html_document
# output:
#   github_document:
#     toc: false
# editor_options: 
#   markdown: 
#     wrap: 72
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Input

## Load codes and DENV data

#### Load index functions

```{r here setup, echo = F, eval=T}
## Not printed, set directory
here::i_am("pocketD4.Rmd")
st <- format(Sys.time(), "%Y-%m-%d_%H%M")


```

### Load necessary packages

```{r packages, echo = T, eval=T, , results = 'hide', warning=FALSE, message=FALSE}
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("ggtree")

# install.packages("remotes")
# remotes::install_github(repo = "stan-dev/cmdstanr")

library(ape, quiet = T); library(phytools, quiet = T); library(stringr, quiet = T)
library(MetBrewer, quiet = T); library(parallel, quiet = T); library(mgcv, quiet = T)
library(cowplot, quiet = T); library(ggplot2, quiet = T); library(ggtree, quiet = T);
library(cmdstanr, quiet = T); library(binom, quiet = T); library(plotly) ; library(DescTools); 
library(readr, quiet = T); library(glue, quiet = T)
library(grid)
library(png)
library(vDiveR)
library(vcfR)
library(reticulate)
library(tidyverse)
library(foreach)
library(extrafont)
library(dsmisc)
library(phangorn)
library(glue)
library(readr)
library(svglite)
library(here)
library(svglite)
library(teal.modules.general)

# font_import()
loadfonts(device="all") 

```

First, source all the necessary functions:

```{r functions, eval=T}
source(file = '2_Functions/2_1_Index_computation_20240909.R')
source(file = '2_Functions/2_2_Lineage_detection_20240909.R')
source(file = '2_Functions/2_3_Lineage_fitness_20240909.R')
source(file = '2_Functions/DF_edits_create_vcf_from_fasta.R')

# Function to extract the start of date ranges
extract_start_date <- function(date_entry) {
  # Check if the entry contains a range
  if (str_detect(date_entry, "\\(")) {
    # Extract the date before the '('
    start_date <- str_extract(date_entry, "^[^\\s]+")
  } else {
    # It's a single date
    start_date <- date_entry
  }
  
  return(start_date)
}


#Reconstructs ancestral states (e.g., mutation presence/absence) along phylogenetic trees based on SNP data.

reconstruct_node_states = function(tree, dataset_tips, dataset_with_nodes, min_prop, max_prop, names_seqs){
  # Meta-data with nodes 
  dataset_with_inferred_resonstruction = dataset_with_nodes[,1:4]
  possible_snps = names(dataset_tips[,4:ncol(dataset_tips)])
  
  ## Filter SNP on frequency: only keep SNP that are present in:
  ## >1% of dataset
  ## <99% of dataset
  prevalence = t(apply(dataset_tips[,4:ncol(dataset_tips)], MARGIN = 2, function(x)table(factor(x, levels = c("0", "1")))))
  prevalence_prop = prevalence[,2]/(prevalence[,1]+prevalence[,2])
  prevalence = cbind(prevalence, prevalence_prop)
  possible_snps = names(which(prevalence[,3] >= min_prop & prevalence[,3] <= max_prop))
  
  ## Filter dataset_tips accordingly
  a = which(is.na(match(colnames(dataset_tips), possible_snps)) == F)
  dataset_tips = dataset_tips[,c(1:3, a)]
  
  print(paste0('Going though ', length(possible_snps), ' snps'))
  k=1
  for(i in possible_snps){
    print(paste0(k, ' / ', length(possible_snps), ' snps'))
    
    snp_data = dataset_tips[,which(colnames(dataset_tips) == i)]
    snp_data = as.factor(snp_data)
    names(snp_data) = names_seqs
    
    ## Perform reconstruction for this position
    # rec = phytools::fastAnc(tree = tree, x = snp_data, CI = TRUE, vars = T) ## factAnc lots faster than ace.ML
    rec = ape::ace(phy = tree, x = snp_data, method = "pic")
    rec_all = c(snp_data, rec$ace) ## List of all states: first all tips, then all nodes
    
    ## Find first state, to set it to 0, always
    first_state = rec_all[which(dataset_with_inferred_resonstruction$ID == length(names_seqs) + 1)]
    first_state = round(first_state, digits = 0)
    
    ## Write reconstruction in the big dataset
    if(first_state < 1.5){ ## First state is 0: all good
      dataset_with_inferred_resonstruction = cbind(dataset_with_inferred_resonstruction, rec_all - 1)     
    }
    if(first_state > 1.5){ ## First state is 1: have to change to 0
      dataset_with_inferred_resonstruction = cbind(dataset_with_inferred_resonstruction,  2 - rec_all)
    }
    
    ## Set column name to position name
    colnames(dataset_with_inferred_resonstruction)[which(colnames(dataset_tips) == i) +1] = i
    
    k=k+1
  }
  return(list('dataset_with_inferred_resonstruction' = dataset_with_inferred_resonstruction,
              'snp_prevalence' = prevalence,
              'possible_snps' = possible_snps))
}


## SNP defining Mutations Function
association_scores_per_group = function(dataset_with_nodes, dataset_with_inferred_reconstruction, tree, 
                                        possible_snps, upstream_window, downstream_window, 
                                        virus = NULL, gene = NULL, level = NULL){
  ## Set list to store results
  ## Extracts unique group names and initializes an empty list (scores) to store results for each group.
  group_names = levels(as.factor(dataset_with_nodes$groups))
  scores = as.list(rep(NA, length(group_names)-1))
  n_tips = length(tree$tip.label)
  
  
  
  ## For each group (except the initial group, which is the root), look at snp association
  for(j in 1:(length(group_names)-1)){ 
    cat("Working on group", j, "of", length(group_names)-1, "\n")
    
    ## Find members of the group and MRCA
    ## Identifies members of the current group and its MRCA. Ensures the MRCA is included in the list of members.
    members = dataset_with_nodes$ID[which(dataset_with_nodes$groups == group_names[j])]
    
    mrca_node = getMRCA(tree, dataset_with_nodes$name_seq[which(dataset_with_nodes$groups == group_names[j] & dataset_with_nodes$is.node == 'no')])
    if (is.null(mrca_node) || is.na(mrca_node)) {
      cat("⚠️ getMRCA failed for group", j, "in",dataset_with_nodes, "\n")
      next
      }
    members = unique(c(members, mrca_node))
    
    ## Update members list, with strains downstream, within the time window
    ## Identifies members of the current group and its MRCA. Ensures the MRCA is included in the list of members.
    time_mrca = dataset_with_nodes$time[which(dataset_with_nodes$ID == mrca_node)] ## Reference time for window
    # tmp = getDescendants(tree, mrca)
    # tmp = tmp[which(dataset_with_nodes$time[tmp] < time_mrca + downstream_window)]
    # members = unique(c(tmp, members))
    
    ## Update members list, with strains upstream, within the time window
    ancest = Ancestors(x = tree, node = mrca_node, type = 'all')
    
    if (length(ancest) == 0) {
      cat("⚠️ Group", j, "in", virus, gene, level, "has MRCA at root → no ancestors available\n")
      
      # Use just members for downstream analysis (no ancestors to add)
      all_nodes_of_interest <- members
    } else {
      # Normal case: Use members + valid ancestors
      time_ancest = dataset_with_nodes$time[ancest]
      time_ancest[1] = time_mrca
      tmp = which(time_ancest < time_mrca - upstream_window)
      if(length(tmp) > 0) ancest = ancest[-tmp]
      if(length(ancest) == 0) ancest = Ancestors(tree, node = mrca, type = 'all')[1]
      groups_ancest = dataset_with_nodes$groups[ancest]
      gr = min(as.numeric(as.character(groups_ancest)))
      ancest = ancest[which(groups_ancest == gr)]
      time_ancest = dataset_with_nodes$time[ancest]
      
      all_nodes_of_interest <- unique(c(ancest, members))
    }
    
    ## Branches of interest
    branches = tree$edge
    branches = branches[match(all_nodes_of_interest, branches[,2]),] ## Take all tips, from ancests, members
    tmp = which(is.na(match(branches[,1], c(ancest, members))))
    if(length(tmp) > 0){  branches = branches[-tmp, ]} ## Remove nodes that are not in ancest or members
    
    ## Branches time
    branches_time = branches
    branches_time[,1] = dataset_with_nodes$time[branches[,1]]
    branches_time[,2] = dataset_with_nodes$time[branches[,2]]
    
    ## Branches group (checked: ok)
    branches_group = branches
    branches_group[,1] = dataset_with_nodes$groups[branches[,1]]
    branches_group[,2] = dataset_with_nodes$groups[branches[,2]]
    
    ## Make branch group matrix binary: 1=group of interest, 0=other group (eg ancestral)
    branches_group[which(branches_group == j, arr.ind = T)] = 1
    branches_group[which(branches_group > j, arr.ind = T)] = 0
    
    snps = time_diff = snps_props_within = snps_props_whole = names_possibles_snps = NULL
    
    
    cat("👉 DEBUG: group", j, "(", group_names[j], ") in", virus, gene, level, "\n")
    cat("Number of members:", length(members), "\n")
    cat("Number of ancestors:", length(ancest), "\n")
    cat("Branches:", nrow(branches), "\n")
    cat("Possible SNPs:", length(possible_snps), "\n")
    ## Extracts SNP states for the start and end nodes of each branch for the current SNP.
    
    scores[[j]] <- numeric(length(possible_snps))
    
    for (i in 1:length(possible_snps)) {
      snp_name <- possible_snps[i]
      col_idx <- which(colnames(dataset_with_inferred_reconstruction) == snp_name)
      branches_snp <- branches
      branches_snp[, 1] <- dataset_with_inferred_reconstruction[branches[, 1], col_idx]
      branches_snp[, 2] <- dataset_with_inferred_reconstruction[branches[, 2], col_idx]
      
      if (length(ancest) == 0) {
        ancestral_state <- dataset_with_inferred_reconstruction[mrca_node, col_idx]
      } else {
        k <- 1
        ancestral_state <- branches_snp[which(branches[, 1] == rev(ancest)[k]), 1]
        k <- 2
        while (str_detect(ancestral_state, pattern = 'n|X') == TRUE & k <= length(ancest) - 1) {
          current_ancestor <- rev(ancest)[k]
          idx <- which(branches[, 1] == current_ancestor)
          if (length(idx) == 0) {
            cat("⚠️ No match for ancestor node in branches[,1] at SNP", i, "\n")
            break
          }
          ancestral_state <- branches_snp[idx, 1]
          k <- k + 1
        }
        if (str_detect(ancestral_state, pattern = 'n|X') == TRUE) {
          ancestral_state <- branches_snp[which(branches[, 1] == mrca_node), 1][1]
        }
      }
      
      branches_snp[str_detect(branches_snp, pattern = 'n|X')] = ancestral_state
      
      ## Converts SNP states into a binary format indicating differences from the ancestral state. 
      ## Calculates the association score (score3) for the SNP.
      branches_snp_bin = (branches_snp!=ancestral_state)
      
      tmp = which(branches_group[,1] == 1 & branches_group[,2] == 1)
      Px = as.numeric(branches_group[tmp,])
      ## Px = 1 → If the branch belongs to the phylogenetic group.
      ## Px = 0 → If the branch is outside the group.
      Sx = as.numeric(branches_snp_bin[tmp,])
      ## Sx = 1 → If the SNP is mutated at that position.
      ##Sx = 0 → If the SNP retains the ancestral state.
      if (length(Px) == 0) {
        scores[[j]][i] <- NA
        names_possibles_snps <- c(names_possibles_snps, paste0(ancestral_state, '|', snp_name, '|NA'))
        next
      }
      score3 <- (sum((1 - Px)*(1 - Sx), na.rm=TRUE) + sum(Px*Sx, na.rm=TRUE)) / length(Px)
      ## Px: Presence/absence of the phylogenetic group.
      ## Sx: Presence/absence of the SNP.
      ## score3: Measures how well the mutation segregates within the group 
      ## (higher values suggest stronger association).
      ## Entries where the score is 1 indicate a perfect association (mutation is fully linked with the group).
      
      scores[[j]][i] = score3
      
      t <- table(branches_snp)
      if (length(t) == 0 || all(names(t) == ancestral_state)) {
        names_possibles_snps <- c(names_possibles_snps, paste0(ancestral_state, '|', snp_name, '|NA'))
      } else {
        t <- t[names(t) != ancestral_state]
        derived_state <- names(t)[which.max(t)]
        names_possibles_snps <- c(names_possibles_snps, paste0(ancestral_state, '|', snp_name, '|', derived_state))
      }
      
    }
    scores[[j]] = scores[[j]]#- median(scores[[j]])
    names(scores[[j]]) = names_possibles_snps
    ## Each entry for a group has both a name and a value:
    ## A name that follows the pattern "X|Y|Z", which corresponds to:
      ## X: Ancestral state (e.g., 0).
      ## Y: SNP position index (e.g., 1, 2, 3, ...).
      ## Z: Derived state (present for some SNPs, missing for others).
    ## A value that represents a computed score for SNP association
    
    ## For example, in 1.06490532281936|15|1 - Value 1 
    ## The score is 1.06490532281936 (a high association score)
    ## 15 is the SNP position
    ## 1 is the derived state (different from the ancestral state 0)
    ## Value of 1 Suggests Linieage defining mutaiton
    ## As compared to 0.0642177183586905|136|0 - Value 1
    
  }
  return(scores)
  ## calculate association scores between SNPs and phylogenetic groups, considering evolutionary and temporal relationships.
  ## The scores quantify how well specific SNPs distinguish groups of nodes/tips in the tree.
}


read_clean_tree <- function(file) {
  tr <- ape::read.nexus(file)
  tr <- collapse.singles(ladderize(multi2di(tr, random = FALSE), right = FALSE))
  tr
}

```

**Explainer: For the association scores:**

`score3 <- (sum((1 - Px)*(1 - Sx), na.rm=TRUE) + sum(Px*Sx, na.rm=TRUE)) / length(Px)`

$$
\text{score3} = \frac{\sum (1 - P_x)(1 - S_x) + \sum P_x S_x}{\text{length}(P_x)}
$$

**First term: `sum((1 - Px) * (1 - Sx))`**

-   Counts cases where **neither** the group nor the SNP mutation is
    present

**Second term: `sum(Px * Sx)`**

-   Counts cases where **both** the group and the SNP mutation are
    present.

**Denominator: `length(Px)`**

-   Normalizes by the number of branches considered.

##### **What Does `score3` Represent?**

-   If **score = 1** → **Perfect association** (mutation occurs
    exclusively within the group).

-   If **score = 0** → No association (mutation occurs randomly outside
    the group).

-   Intermediate values (e.g., `0.5`) → Partial association (mutation is
    somewhat linked to the group).

## Explainer: Compute the index dynamics

To compute the index of all nodes, use the function *compute.index*, you
will need different inputs:

1.  Data:
    -   the *timed_tree*
    -   *metadata* dataframe (see below the details of the
        *dataset_with_nodes*)
    -   *distance matrix*: can be computed from the timed tree with the
        function *dist.nodes.with.names*
2.  Information on the pathogen genome (these are pathogen specific):
    -   The genome length of the pathogen considered: *genome_length*
        (in bp).
    -   The mutation rate of the pathogen considered: *mutation_rate*
        (in bp/genome/year), an average is fine.
3.  Index parameters (these are both pathogen-specific and
    dataset-specific):
    -   The timescale: *timescale* (in years), this will be used to
        compute the bandwidth, see more details below.
    -   Window of time on which to search for samples in the population:
        *wind*, see more details below.

The function outputs a vector containing the index of each node
(internal and terminal).

**timescale**: The timescale determines the kernel which enables to
track lineage emergence dynamically, focusing on short distances between
nodes (containing information about recent population dynamics) rather
than long distances (containing information about past evolution). The
timescale is tailored to the specific pathogen studied and its choice
depends on the molecular signal, as well as the transmission rate. In
the study, we used timescales ranging from months (typical of RNA
viruses) to years (typical of bacteria). To determine a timescale
suitable for your dataset, we recommend thinking about the generation
time of the pathogen considered, its mutation rate, and the amount of
diversity already accumulated. For example, at the time of the analysis,
SARS-CoV-2 was a new pathogen, spreading quickly and accumulating
diversity at a rate of \~2 mutations per month. Therefore, a small
timescale of less than a year chosen (0.15 years). On the contrary,
*Mycobacterium tuberculosis* is an older and relatively slowly spreading
pathogen, which accumulates mutations at a rate of \~0.2 mutation per
year. A much larger timescale was then chosen (30 years), to reflect
this. Ultimately, the best timescale is one that maximises the
visualisation of population dynamics. We recommend trying different
values.

**wind**: The choice of *wind* will depend on the sampling intensity of
the dataset. It defines the window of time around each node on which to
search for samples in the population. Ultimately it smooths the index
dynamics. As a mean of example, for SARS-CoV-2, we set *wind* to 15
days, as the dataset was intensely sampled. But for *Bordetella
pertussis*, which is more sparsely sampled, we chose a *wind* of 1 year.
If *wind* is too large, then all the nodes are considered to be part of
the same time window. If *wind* is too small, then only the nodes in
direct proximity of the node of interest will be considered in the time
window, which can result in noisy index dynamics. We recommend choosing
a *wind* value that enables to span multiple sampling times, for example
if you have samples and nodes every week, you may choose a *wind* of
\~1-2 months. If you have samples and nodes every year, you may choose a
*wind* of \~2 years.

### Set the index parameters.

```{r index params, eval=T}
## Length genome 
genome_length = 11000 # reference  https://nextstrain.org/dengue/all/genome Could also just use 11kb or something highly specific
## Mutation rate 
mutation_rate = 7.6e-4
  #7.6e-4 # mutation rate #ref https://pmc.ncbi.nlm.nih.gov/articles/PMC9030598/
mutation_rate_sci <- formatC(mutation_rate, format = "e", digits = 2)
## Parameters for the index
timescale = 7 ## Timescale in years; raises index 
## Window of time on which to search for samples in the population
wind = 365 #days 
wind = wind/365 #Wind smooths lines


```


## Find DENGUE clades based on index dynamics

#### Run the lineage detection algorithm on DENGUE data

To run the lineage detection algorithm, use the function
*find.groups.by.index.dynamics*, you will need different inputs:

1.  Data: *timed_tree* (must be the same as the one used in
    compute.index) and *metadata* (*dataset_with_nodes*, with the index
    values of each node)
2.  Lineage detection parameters:
    -   *min_descendants_per_tested_node*: to start the analysis, start
        from nodes that have this minimum number of sequences
    -   *min_group_size*: minimum group size, when creating a new
        potential split
    -   *node_support*: numeric value of support of each node (e.g.
        mutations on the branch leading to the node, or bootstrap
        support)
    -   *threshold_node_support*: threshold on the node support for the
        nodes to be considered in the detection algorithm
    -   *weight_by_time*: size of the window of time on which to compute
        the weights (NULL or numeric, in years)
    -   *weighting_transformation*: type of weighting to use (NULL,
        inv_freq, inv_sqrt, or inv_log)
    -   *max_groups_found*: maximum number of groups to find (Integer)
3.  Technical parameters: they do not necessarily need to be updated
    (see the function documentation for details): *p_value_smooth*,
    *stepwise_deviance_explained_threshold*, *stepwise_AIC_threshold*,
    *k_smooth*, *parallelize_code*, *number_cores*, *plot_screening*,
    *keep_track* and *log_y*.

The function outputs multiple elements in a list:

-   potential_splits: vector of the nodes included in most complex model
    tested
-   best_dev_explained: vector of the deviance explained by the best
    models for each number of groups
-   first_dev: the null deviance of the initial model (when no lineage
    is present)
-   best_AIC : vector of the AIC of the best models for each number of
    groups
-   best_BIC: vector of the BIC of the best models for each number of
    groups
-   best_summary: list of the summaries of the best models for each
    number of groups
-   best_mod: list of the best models for each number of groups
-   best_groups: list of the groups used in the best models for each
    number of groups
-   best_nodes_names: list of the nodes included in the best models for
    each number of groups

Typically, one chooses a value of *max_groups_found* greater than the
expected number of lineages. The algorithm then runs until it finds all
those groups, or until it cannot find any significant split anymore. The
user can then check the deviance explained by all the models with
increasing complexity and choose an adequate number of groups.

Once the split nodes have been defined, the user can then extract the
group ID for each node using the function *merge.groups*. One can choose
to refine these groups if needed, by setting a minimum number of nodes
per group (*group_count_threshold*) or a minimum frequency of each group
(*group_freq_threshold*).

##Parameters for the detection:

Threshold not support: Max lineages is the same as number of lineages. Idea is restrict search space to certain level. Trying to restrict space to only well supported nodes. Each node has support value: 1. bootstrap/posterior (similar to Tb tree) support comes from data Creates sparsing issues in viruses that don't change much. 2. Proxy of how many mutations ahead of branch. If two nodes are separated by a short amount of time then the hypothesis that both nodes are distinct is less likely. Uses branch lenght between nodes. Only permit nodes that are at least 1 mutation apart. Length must be how long it takes for 1 mutation.


Using time_window_initial + increment: sometimes helpful if wanting to look at newer lineages. But elbow plots are harder to interpret as depeneding on time params. Not recommended.  

min_descendants_per_tested_node: only considering nodes with n tip and node descendents. A filter or partition of the tree. 

Index dynamics are made of many splines (each with a k), p values for the spline need to better explain dynamics. k is the number of knots 


How many lineages to choose is a practical question not something that is easy to ground in math. Elbow plot analysis is basically arbitrary. First elbow is the known dynamics, the second elbow is more exploratory. Stopping at the first linear point on horizontal line. 

Weight by time: Seqiences are a time series, this compensates for some times where there are no sequences. Thus the model is trying to overexplain the densely time sampled variants. Penalty for higher samples in higher sampled times. Weight by time is the bucket/slice of time in years. 
weighting_transformation: how the penalty is applied. Can try all of these but can't really compare between compare since AIC and BIC are poor measure of GAM comparison.  


```{r model params, eval =TRUE}
threshold_node_support = 1/(11000*7.6e-4)  #Changed from 1/(29903*0.00081) this is likely SARS bp num times the substitution rate? 
time_window_initial = 2030 #last sequence is in 2015, okay to set at 2030. Just number that consider whole tree. 
max_time_window_increment = 100 #Why is this 100 years? Not being used if window_intial is in the future. 
min_time_window_increment = 0.25 #years
min_descendants_per_tested_node = 30 #Changed from 30
min_group_size = 30 #Changed from 30
p_value_smooth = 0.05 #formerly 0.05
min_stepwise_deviance_explained_threshold = 0
max_stepwise_deviance_explained_threshold = 0.005
min_stepwise_AIC_threshold = 0
max_stepwise_AIC_threshold = 7
weight_by_time = 0.1 #in years and shows the bins of time, consider increasing for dengue
weighting_transformation = c('inv_sqrt') 
k_smooth = -1 # smooting term for basis functions, -1 allows mgcv to set the default (usually 10)
plot_screening = T
parallelize_code = T
number_cores = 8
max_groups_found_default = 30
keep_track = T
log_y = T

```
To ‘force’ more lineages to be detected by phylowave, you have two options (and you can use both at the same time):
- 1: change the lineages you’re proposing: if your current code restricts the lineages to have at least 100 tips+nodes, you could change this to 10. This will effectively allow the model to test smaller lineages. The parameters to change for this are ‘min_descendants_per_tested_node’, and also min_group_size’. You can also set the ‘threshold_node_support’ parameter to NULL, effectively not filtering node based on any support value.
- 2: make it harder for the model to stop the exploration of new lineages. By default, phylowave needs each new lineage set to significantly improve the model fit. The significance of the improvement is mainly set by two parameters: “p_value_smooth” and “stepwise_deviance_explained_threshold”. The first step is making sure each new group (linked with an index dynamics) has a well-supported spline, typically with a p-value of 0.05. You could change this to 1, effectively allowing all splines to be included (so no filter anymore). The second parameter sets the minimum improvement of deviance explained by each model. If this minimum is not reached, than the model stops. You can set this one to 0, which will make your model stop only when it can’t improve the deviance at all.

#load in what you need to debug
```{r}
# potential_splits_4 <- read_rds("potential_splits_4.rds")

dataset_with_nodes_4 <- read_rds("dataset_with_nodes_4.rds")
tree_DENV4 <- read_rds("tree_DENV4.rds")
n_seq_4 <- read_rds("n_seq_4.rds")
# split_4 <- read_rds("split_4.rds")

```


#Potential Splits 

DENV4 Run the detection function (this steps takes approximately \<10
min on 2 cores):

```{r smalltimewindow, eval=FALSE, results = 'hide', warning=FALSE, message=FALSE}
start_time = Sys.time()
ps_smallwindow_4 = find.groups.by.index.dynamics(timed_tree = tree_DENV4,
                                                 metadata = dataset_with_nodes_4,
                                                 node_support = tree_DENV4$edge.length[match((n_seq_4+1):(2*n_seq_4-1),
                                                                                            tree_DENV4$edge[,2])],
                                                 threshold_node_support = threshold_node_support,
                                                 time_window_initial = time_window_initial,
                                                 time_window_increment = min_time_window_increment,
                                                 min_descendants_per_tested_node = min_descendants_per_tested_node,
                                                 min_group_size = min_group_size,
                                                 p_value_smooth = p_value_smooth,
                                                 stepwise_deviance_explained_threshold = min_stepwise_deviance_explained_threshold,
                                                 stepwise_AIC_threshold = min_stepwise_AIC_threshold,
                                                 weight_by_time = weight_by_time,
                                                 weighting_transformation = weighting_transformation,
                                                 k_smooth = k_smooth,
                                                 parallelize_code = parallelize_code,
                                                 number_cores = number_cores, 
                                                 plot_screening = plot_screening,
                                                 max_groups_found =  max_groups_found_default,
                                                 keep_track = keep_track)
end_time = Sys.time()
print(end_time - start_time)


attr(ps_smallwindow_4, "params") <- list(
  head_note = "Run keeping all previous params the same but using smaller runitme windows",
  node_support = tree_DENV4$edge.length[match((n_seq_4+1):(2*n_seq_4-1),
                                              tree_DENV4$edge[,2])],
  time_window_initial = time_window_initial,
  time_window_increment = min_time_window_increment,
  min_descendants_per_tested_node = min_descendants_per_tested_node,
  min_group_size = min_group_size,
  p_value_smooth = p_value_smooth,
  stepwise_deviance_explained_threshold = min_stepwise_deviance_explained_threshold,
  stepwise_AIC_threshold = min_stepwise_AIC_threshold,
  weight_by_time = weight_by_time,
  weighting_transformation = weighting_transformation,
  k_smooth = k_smooth,
  parallelize_code = parallelize_code,
  number_cores = number_cores,
  max_groups_found = max_groups_found_default,
  keep_track = keep_track
)

attr(ps_smallwindow_4, "params")



```
1] "Considering tree up to 2030-------------------------->"
[1] "Testing 40 nodes"
[1] "Better fit found"
[1] "Deviance explained by 45.158% more"
[1] 441
[1] "Testing 39 nodes"
[1] "Better fit found"
[1] "Deviance explained by 7.035% more"
[1] 441 443
[1] "Testing 38 nodes"
[1] "Better fit found"
[1] "Deviance explained by 0.659% more"
[1] 441 443 557
[1] "Testing 37 nodes"
[1] "No better fit found"





```{r}
library(lattice)
b0 <- 10
b1 <- .5
b2 <- .3
g <- expand.grid(x = 1:20, y = 1:20)
g$z <- b0 + b1*g$x + b2*g$y
 
png(file="example%03d.png", width=300, heigh=300)
  for (i in seq(0, 350 , 10)){
    print(wireframe(z ~ x * y, data = g,
              screen = list(z = i, x = -60)))
  }
dev.off()
 
# converting .png file in .gif using ImageMagick
system("/opt/local/bin/convert -delay 40 *.png example_2_reduced.gif")
 
# Remove .png file
file.remove(list.files(pattern=".png"))
```


```{r large timewindow}

#Large time increment 
start_time = Sys.time()
ps_largewindow_4 = find.groups.by.index.dynamics(timed_tree = tree_DENV4,
                                                 metadata = dataset_with_nodes_4,
                                                 node_support = tree_DENV4$edge.length[match((n_seq_4+1):(2*n_seq_4-1),
                                                                                            tree_DENV4$edge[,2])],
                                                 threshold_node_support = threshold_node_support,
                                                 time_window_initial = time_window_initial,
                                                 time_window_increment = max_time_window_increment,
                                                 min_descendants_per_tested_node = min_descendants_per_tested_node,
                                                 min_group_size = min_group_size,
                                                 p_value_smooth = p_value_smooth,
                                                 stepwise_deviance_explained_threshold = min_stepwise_deviance_explained_threshold,
                                                 stepwise_AIC_threshold = min_stepwise_AIC_threshold,
                                                 weight_by_time = weight_by_time,
                                                 weighting_transformation = weighting_transformation,
                                                 k_smooth = k_smooth,
                                                 parallelize_code = parallelize_code,
                                                 number_cores = number_cores, 
                                                 plot_screening = plot_screening,
                                                 max_groups_found =  max_groups_found_default,
                                                 keep_track = keep_track)
end_time = Sys.time()
print(end_time - start_time)


attr(ps_largewindow_4, "params") <- list(
  head_note = "Run keeping all previous params the same but using larger runitme windows",
  node_support = tree_DENV4$edge.length[match((n_seq_4+1):(2*n_seq_4-1),
                                              tree_DENV4$edge[,2])],
  threshold_node_support = threshold_node_support,
  time_window_initial = time_window_initial,
  time_window_increment = max_time_window_increment,
  min_descendants_per_tested_node = min_descendants_per_tested_node,
  min_group_size = min_group_size,
  p_value_smooth = p_value_smooth,
  stepwise_deviance_explained_threshold = min_stepwise_deviance_explained_threshold,
  stepwise_AIC_threshold = min_stepwise_AIC_threshold,
  weight_by_time = weight_by_time,
  weighting_transformation = weighting_transformation,
  k_smooth = k_smooth,
  parallelize_code = parallelize_code,
  number_cores = number_cores, 
  plot_screening = plot_screening,
  max_groups_found =  max_groups_found_default,
  keep_track = keep_track
)
attr(ps_smallwindow_4, "params")
```



```{r more loose params}


start_time = Sys.time()
ps_looseparams_4 = find.groups.by.index.dynamics(timed_tree = tree_DENV4,
                                                 metadata = dataset_with_nodes_4,
                                                 node_support = tree_DENV4$edge.length[match((n_seq_4+1):(2*n_seq_4-1),
                                                                                            tree_DENV4$edge[,2])],
                                                 threshold_node_support = NULL,
                                                 time_window_initial = 2030,
                                                 time_window_increment = 100,
                                                 min_descendants_per_tested_node = 10,
                                                 min_group_size = 10,
                                                 p_value_smooth = 1,
                                                 stepwise_deviance_explained_threshold = 0,
                                                 stepwise_AIC_threshold = 0,
                                                 weight_by_time = 0.1,
                                                 weighting_transformation = c('inv_sqrt'),
                                                 k_smooth = -1,
                                                 parallelize_code = T,
                                                 number_cores = 8, 
                                                 plot_screening = T,
                                                 plot_gif = T,
                                                 gif_output_dir = "3_Output_Figures/3_5_DENV/3_5_99_Supplement",
                                                 gif_filename  = "screening_loose_params.gif",
                                                 max_groups_found =  30,
                                                 keep_track = T)
end_time = Sys.time()
print(end_time - start_time)


attr(ps_looseparams_4, "params") <- list(
  head_note = "Run to force more lineages",
  node_support = tree_DENV4$edge.length[match((n_seq_4+1):(2*n_seq_4-1),
                                              tree_DENV4$edge[,2])],
  threshold_node_support = NULL,
  time_window_initial = 2030,
  time_window_increment = 100,
  min_descendants_per_tested_node = 10,
  min_group_size = 10,
  p_value_smooth = 1,
  stepwise_deviance_explained_threshold = 0,
  stepwise_AIC_threshold = 0,
  weight_by_time = 0.1,
  weighting_transformation = c('inv_sqrt'),
  k_smooth = -1,
  parallelize_code = T,
  number_cores = 8, 
  plot_screening = T,
  max_groups_found =  30,
  keep_track = T
)
attr(ps_looseparams_4, "params")

plot.gam(ps_looseparams_4[["best_mod"]][[1]])


ps_looseparams_4$best_summary
mgcv::gam.check(ps_looseparams_4[["best_mod"]][[1]])
mgcv::concurvity(ps_looseparams_4[["best_mod"]][[1]], full=TRUE)
mgcv::concurvity(ps_looseparams_4[["best_mod"]][[1]], full=FALSE)
plot(ps_looseparams_4[["best_mod"]][[1]], pages=1, shade=TRUE) 


mgcv::k.check(ps_looseparams_4[["best_mod"]][[5]])
mgcv::gam.check(ps_looseparams_4[["best_mod"]][[5]])
mgcv::concurvity(ps_looseparams_4[["best_mod"]][[5]], full=TRUE)
mgcv::concurvity(ps_looseparams_4[["best_mod"]][[5]], full=FALSE)
plot(ps_looseparams_4[["best_mod"]][[1]], pages=1, shade=TRUE)


mgcv::gam.check(ps_looseparams_4[["best_mod"]][[10]])
mgcv::concurvity(ps_looseparams_4[["best_mod"]][[10]], full=TRUE)
 mgcv::concurvity(ps_looseparams_4[["best_mod"]][[10]], full=FALSE)
plot(ps_looseparams_4[["best_mod"]][[10]], shade = TRUE)   


```


```{r linlog group plot}
max_groups_found_line_4 = 3

for (i in 4) {
  
 svglite(sprintf("DENV_%d_linlog.svg", i), width = 12, height = 6)
  par(mfrow = c(1, 2), oma = c(2, 2, 1, 1), mar = c(2, 2, 2, 0.5),
      mgp = c(0.75, 0.25, 0), cex.axis = 0.5, cex.lab = 0.5,
      cex.main = 0.7, cex.sub = 0.5)
  ## Dynamically retrieve the potential splits for the current group
  potential_splits <- get(paste0('ps_looseparams_', i))
  max_groups_found_line_i <- get(paste0('max_groups_found_line_', i))
  
  ## Dynamically create or update the data frame for the current group
  df_explained_dev_i <- data.frame(
    'N_groups' = 0:length(potential_splits$best_dev_explained),
    'Non_explained_deviance' = (1 - c(potential_splits$first_dev, potential_splits$best_dev_explained)),
    'Non_explained_deviance_log' = log(1 - c(potential_splits$first_dev, potential_splits$best_dev_explained))
  )
  
  ## Adjust the log values
  df_explained_dev_i$Non_explained_deviance_log <- df_explained_dev_i$Non_explained_deviance_log - 
                                                   min(df_explained_dev_i$Non_explained_deviance_log)
  
  ## Linear scale plot
  plot(df_explained_dev_i$N_groups,
       df_explained_dev_i$Non_explained_deviance,
       bty = 'n', ylim = c(0, 1),
       xaxt = 'n', yaxt = 'n', pch = 16, 
       main = paste('Linear scale DENV', i), cex = 0.5,
       ylab = 'Non-explained deviance (%)', xlab = 'Number of groups')
  axis(1, lwd = 0.5, tck = -0.02)
  axis(2, las = 2, at = seq(0, 1, 0.1), labels = seq(0, 1, 0.1) * 100, lwd = 0.5, tck = -0.02)
  abline(v = max_groups_found_line_i, col = "purple", lty = 2, lwd = 0.5)
  
  ## Log scale plot
  plot(df_explained_dev_i$N_groups,
       df_explained_dev_i$Non_explained_deviance,
       log = 'y',
       ylim = c(0.01, 1),
       bty = 'n',
       xaxt = 'n', yaxt = 'n', pch = 16, 
       main = paste('Log scale DENV', i), cex = 0.5,
       ylab = 'Non-explained deviance (%) - log scale', xlab = 'Number of groups')
  axis(1, lwd = 0.5, tck = -0.02)
  axis(2, las = 2, 
       at = c(0.01, 0.1, 0.25, 0.5, 1),
       labels = c(0.01, 0.1, 0.25, 0.5, 1) * 100, 
       lwd = 0.5, tck = -0.02)
  abline(v = max_groups_found_line_i, col = "purple", lty = 2, lwd = 0.5)
  
  dev.off()
} 

```


#TODO: ASK WHY NOT MIN AIC/BIC

```{r}
mods <- ps_looseparams_4[["best_mod"]]

df <- tibble(
  step = seq_along(mods),
  AIC  = sapply(mods, AIC),
  BIC = sapply(mods, BIC),
  dev  = sapply(mods, function(m) summary(m)$dev.expl)
)

kmin <- df$step[which.min(df$AIC)]
kmin_BIC <- df$step[which.min(df$BIC)]

p1 <- ggplot(df, aes(step, AIC)) +
  geom_line() + geom_point() +
  geom_vline(xintercept = kmin, linetype = 2, color = "purple") +
  labs(x = "Number of groups (step index)", y = "AIC",
       title = "Model AIC across tested splits",
       subtitle = sprintf("Min AIC at k=%d (AIC=%.1f)", kmin, df$AIC[kmin]))

p2 <- ggplot(df, aes(step, dev * 100)) +
  geom_line() + geom_point() +
  labs(x = "Number of groups (step index)", y = "Deviance explained (%)",
       title = "Deviance explained across tested splits") 

p3 <- ggplot(df, aes(step, BIC)) +
  geom_line() + geom_point() +
  geom_vline(xintercept = kmin_BIC, linetype = 2, color = "purple") +
  labs(x = "Number of groups (step index)", y = "BIC",
       title = "Model BIC across tested splits",
       subtitle = sprintf("Min BIC at k=%d (BIC=%.1f)", kmin_BIC, df$BIC[kmin_BIC])) 

p4 <- ggplot(df, aes(step)) +
  geom_line(aes(y = AIC), colour = "#012169FF") + geom_point(aes(y = AIC), colour = "#012169FF") +
  geom_line(aes(y = BIC), colour = "#C8102EFF") + geom_point(aes(y = BIC), colour = "#C8102EFF") +
  geom_vline(xintercept = kmin, linetype = 2, color = "#C8102Eaa") +
  geom_vline(xintercept = kmin_BIC, linetype = 2, color = "#012169aa") +
  labs(x = "Number of groups (step index)", y = "Value",
       title = "Model AIC & BIC across tested splits",
       subtitle = glue::glue("Min AIC at k={kmin} (AIC={round(df$AIC[kmin],1)})\n",
                      "Min BIC at k={kmin_BIC} (BIC={round(df$BIC[kmin_BIC],1)})") )





p5 <- p4 + p2

print(p1); print(p2); print(p3); print(p4); print(p5)

# save if you want
ggsave("aic_by_step_DENV4.png", p1, width = 6, height = 4, dpi = 300)
ggsave("devexp_by_step_DENV4.png", p2, width = 6, height = 4, dpi = 300)
ggsave("bic_by_step_DENV4.png", p3, width = 6, height = 4, dpi = 300)
ggsave("AIC_BIC_devexp_by_step_DENV4.png", p5, width = 6, height = 4, dpi = 300)

```

Optimize the number of groups: set the minimum number of sequences per
group to 30, with a minimum frequency of 1%.

The idea of merging groups functions is to merge small lineages into bigger ones and to enforce the minimum frequency and merging that with geneticially similar groups. Bascially to change the numbers of groups without rerunning the model. Fitting logistic model on infrequent lineages is not super relevant 

```{r minimum groups, eval = T}

for (i in 4) {
  ## Dynamically retrieve the relevant timed tree and dataset_with_nodes
  timed_tree <- get(paste0('tree_DENV', i))
  dataset_with_nodes <- get(paste0('dataset_with_nodes_', i))
  potential_splits <- get(paste0('potential_splits_', i))
  
  ## Call the merge.groups function
  split <- merge.groups(
    timed_tree = timed_tree,
    metadata = dataset_with_nodes,
    initial_splits = potential_splits$potential_splits,
    group_count_threshold = 30,
    group_freq_threshold = 0.01 
  )
  
  ## Dynamically assign the result to split_1, split_2, etc.
  assign(paste0('split_', i), split)
}

```

Label sequences with these new groups, and assign a color to each of
them.

```{r, eval = T}

for (i in 4) {
  ## Dynamically retrieve the dataset and split for the current iteration
  dataset_with_nodes <- get(paste0('dataset_with_nodes_', i))
  split <- get(paste0('split_', i))
  
  ## Label sequences with new groups
  dataset_with_nodes$groups <- as.factor(split$groups)
  
  ## Reorder labels by time of emergence
  name_groups <- levels(dataset_with_nodes$groups)
  time_groups_world <- NULL
  for (j in 1:length(name_groups)) {
    time_groups_world <- c(
      time_groups_world, 
      min(dataset_with_nodes$time[which(dataset_with_nodes$groups == name_groups[j] & 
                                        dataset_with_nodes$is.node == 'no')])
    )
  }
  
  ## Update group levels
  levels(dataset_with_nodes$groups) <- match(name_groups, order(time_groups_world, decreasing = TRUE))
  dataset_with_nodes$groups <- as.numeric(as.character(dataset_with_nodes$groups))
  dataset_with_nodes$groups <- as.factor(dataset_with_nodes$groups)
  
  ## Update names in the split list
  split$tip_and_nodes_groups <- match(split$tip_and_nodes_groups, order(time_groups_world, decreasing = TRUE))
  names(split$tip_and_nodes_groups) <- 1:length(split$tip_and_nodes_groups)
  split$groups <- as.factor(split$groups)
  levels(split$groups) <- match(name_groups, order(time_groups_world, decreasing = TRUE))
  split$groups <- as.numeric(as.character(split$groups))
  
  ## Choose color palette
  n_groups <- length(name_groups)
  colors_groups <- met.brewer(name = "Cross", n = n_groups, type = "continuous")
  
  ## Color each group
  dataset_with_nodes$group_color <- dataset_with_nodes$groups
  levels(dataset_with_nodes$group_color) <- colors_groups
  dataset_with_nodes$group_color <- as.character(dataset_with_nodes$group_color)
  
  ## Dynamically save group names and colors for this iteration
  assign(paste0('name_groups_', i), name_groups)
  assign(paste0('colors_groups_', i), colors_groups)
  
  ## Save the updated variables dynamically
  assign(paste0('dataset_with_nodes_', i), dataset_with_nodes)
  assign(paste0('split_', i), split)
  assign(paste0('colors_groups_', i), colors_groups)
}



```
