---
title: "Code for the paper 'Learning the fitness dynamics of pathogens from phylogenies'"
# author: Noémie Lefrancq
# date: 2024-09-10
output:
  github_document:
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load codes and SARS-CoV-2 data

#### Load index functions

```{r, echo = F, eval=T}
## Not printed, set directory
#setwd('~/Dropbox/THD/Index_paper_methods_Sc2_H3N2_BP_TB/paper-index-fitness-dynamics-trees/')
```

First, source all the necessary functions:
```{r, eval=T}
source(file = '2_Functions/2_1_Index_computation_20240909.R')
source(file = '2_Functions/2_2_Lineage_detection_20240909.R')
source(file = '2_Functions/2_3_Lineage_fitness_20240909.R')
```

Load necessary packages
```{r, echo = T, eval=T, , results = 'hide', warning=FALSE, message=FALSE}
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("ggtree")

# install.packages("remotes")
# remotes::install_github(repo = "stan-dev/cmdstanr")

library(ape, quiet = T); library(phytools, quiet = T); library(stringr, quiet = T)
library(MetBrewer, quiet = T); library(parallel, quiet = T); library(mgcv, quiet = T)
library(cowplot, quiet = T); library(ggplot2, quiet = T); library(ggtree, quiet = T);
library(cmdstanr, quiet = T); library(binom, quiet = T); library(plotly)
```

Versions:

Packages:
ape v5.7-1, phytools v1.9-16, stringr v1.5.0, MetBrewer v0.2.0, parallel v4.1.2, mgcv v1.8-42, cowplot v1.1.1, ggplot2 v3.4.3, ggtree v3.2.1, cmdstanr v0.5.2, binom v1.1

R: 4.1.2
 
#### Load data

Load the NexStrain SARS-CoV-2 tree, in which all the tip name include: collection time, location and Pango lineage 
```{r, eval=T}
tree_sars_cov2 = read.nexus('1_Data/1_1_SARS_CoV_2/Tree_SARSCoV2_global_alltime_nextstrain_20230414.nexus')
## Make sure the tree is binary, and ladderized
tree_sars_cov2 = collapse.singles(ladderize(multi2di(tree_sars_cov2, random = F), right = F))
## Names all sequences
names_seqs = tree_sars_cov2$tip.label
n_seq = length(names_seqs)
## Collection times of all sequences
times_seqs = as.numeric(sapply(names_seqs, function(x)tail(str_split(x, pattern = '/')[[1]],2)[1]))
## Nextstrain clades of all sequences
clades_seqs = sapply(names_seqs, function(x)tail(str_split(x, pattern = '/')[[1]],1))
```

Set the index parameters.
```{r, eval=T}
## Length genome 
genome_length = 29903 # reference nextstrain https://www.ncbi.nlm.nih.gov/nuccore/MN908947
## Mutation rate 
mutation_rate = 8.1e-4 # mutation rate used by nextstrain https://github.com/nextstrain/ncov
## Parameters for the index
timescale = 0.15 ## Timescale
## Window of time on which to search for samples in the population
wind = 15 #days
wind = wind/365
```


#### Compute pairwise distance matrix
Compute distance between each pair of sequences and internal nodes in the tree
```{r, eval=T}
genetic_distance_mat = dist.nodes.with.names(tree_sars_cov2)
```

Get the time of each internal node
```{r, eval=T}
nroot = length(tree_sars_cov2$tip.label) + 1 ## Root number
distance_to_root = genetic_distance_mat[nroot,]
root_height = times_seqs[which(names_seqs == names(distance_to_root[1]))] - distance_to_root[1]
nodes_height = root_height + distance_to_root[n_seq+(1:(n_seq-1))]
```

#### Preparation data tips and nodes

Prepare the main dataframe, where the index and lineages of all nodes (internal and terminal) are going to be stored.
```{r, eval=T}
# Meta-data with all nodes 
dataset_with_nodes = data.frame('ID' = c(1:n_seq, n_seq+(1:(n_seq-1))),
                                'name_seq' = c(names_seqs, n_seq+(1:(n_seq-1))),
                                'time' = c(times_seqs, nodes_height),
                                'is.node' = c(rep('no', n_seq), rep('yes', (n_seq-1))),
                                'Nextstrain_clade' = c(clades_seqs, rep(NA, n_seq-1)))
```


#### Compute index of every tip and node
```{r, eval=T}
dataset_with_nodes$index = compute.index(time_distance_mat = genetic_distance_mat, 
                                         timed_tree = tree_sars_cov2, 
                                         time_window = wind,
                                         metadata = dataset_with_nodes, 
                                         mutation_rate = mutation_rate,
                                         timescale = timescale,
                                         genome_length = genome_length)
```

#### Plot tree & index below, with colors from NextStrain clades

First, generate the color key, based on the Nextstrain clade of each sequence.
```{r, eval=T}
## Color key for Nextstrain clades
colors_clade = met.brewer(name="Cross", n=length(levels(as.factor(dataset_with_nodes$Nextstrain_clade))), type="continuous")

## Color of each node, based on the key
dataset_with_nodes$Nextstrain_clade_color = as.factor(dataset_with_nodes$Nextstrain_clade)
clade_labels = levels(dataset_with_nodes$Nextstrain_clade_color)
levels(dataset_with_nodes$Nextstrain_clade_color) = colors_clade
dataset_with_nodes$Nextstrain_clade_color = as.character(dataset_with_nodes$Nextstrain_clade_color)
```

Then plot the tree and index:
```{r, eval=T}
par(mfrow = c(2,1), oma = c(0,0,0,0), mar = c(4,4,0,0))

min_year = 2020
max_year = 2023.5

## Tree
plot(tree_sars_cov2, show.tip.label = FALSE, 
     edge.color = 'grey', edge.width = 0.25,
     x.lim = c(min_year, max_year)-root_height)
tiplabels(pch = 16, col = dataset_with_nodes$Nextstrain_clade_color, cex = 0.3)
axisPhylo_NL(side = 1, root.time = root_height, backward = F,
             at_axis = seq(min_year, max_year, 0.5)-root_height,
             lab_axis = seq(min_year, max_year, 0.5), lwd = 0.5)

## Index
plot(dataset_with_nodes$time, 
     dataset_with_nodes$index, 
     col = adjustcolor(dataset_with_nodes$Nextstrain_clade_color, alpha.f = 1),
     bty = 'n', xlim = c(min_year, max_year), cex = 0.4,
     pch = 16, bty = 'n', ylim = c(0, 1), 
     main = paste0(''), 
     ylab = 'Index', xlab = 'Time (years)', xaxt = 'n', yaxt = 'n')
axis(2, las = 2, lwd = 0.5)
axis(1, lwd = 0.5)

# Color key
legend('topright', 
       legend = clade_labels,
       fill = colors_clade, border = colors_clade,
       cex = 0.5, bty = 'n', ncol = 5)
```


## Find SARS-CoV-2 clades based on index dynamics 

#### Run the lineage detection algorithm on SARS-CoV-2 data

Parameters for the detection: 
```{r, eval = T, eval=FALSE}
time_window_initial = 2030;
time_window_increment = 100;
p_value_smooth = 0.05
weight_by_time = 0.1
k_smooth = -1
plot_screening = F
min_descendants_per_tested_node = 30
min_group_size = 30
weighting_transformation = c('inv_sqrt')

parallelize_code = T
number_cores = 2

max_stepwise_deviance_explained_threshold = 0
max_groups_found = 13
stepwise_AIC_threshold = 0

keep_track = T
```


Run the detection function (this steps takes approximately <10 min on 2 cores):
```{r, eval=FALSE, results = 'hide', warning=FALSE, message=FALSE}
start_time = Sys.time()
potential_splits = find.groups.by.index.dynamics(timed_tree = tree_sars_cov2,
                                                 metadata = dataset_with_nodes,
                                                 node_support = tree_sars_cov2$edge.length[match((n_seq+1):(2*n_seq-1), tree_sars_cov2$edge[,2])],
                                                 threshold_node_support = 1/(29903*0.00081),
                                                 time_window_initial = time_window_initial,
                                                 time_window_increment = time_window_increment,
                                                 min_descendants_per_tested_node = min_descendants_per_tested_node,
                                                 min_group_size = min_group_size,
                                                 p_value_smooth = p_value_smooth,
                                                 stepwise_deviance_explained_threshold = max_stepwise_deviance_explained_threshold,
                                                 stepwise_AIC_threshold = stepwise_AIC_threshold,
                                                 weight_by_time = weight_by_time,
                                                 weighting_transformation = weighting_transformation,
                                                 k_smooth = k_smooth,
                                                 parallelize_code = parallelize_code,
                                                 number_cores = number_cores, 
                                                 plot_screening = plot_screening,
                                                 max_groups_found = max_groups_found, 
                                                 keep_track = keep_track)
end_time = Sys.time()
print(end_time - start_time)
```

Instead, you may wish to load the results:
```{r, eval = T}
potential_splits = readRDS('README_files/potential_splits.rds')
```

Look at the deviance explained by the models with different number of groups. Here for simplicity we directly chose max_groups_found = 13, which is the number of groups used in the original analysis. To decide on 13 groups, we initially ran the algorithm up to 30 groups.
```{r, eval = T}
df_explained_dev = data.frame('N_groups' = 0:length(potential_splits$best_dev_explained),
                              'Non_explained_deviance' = (1-c(potential_splits$first_dev, potential_splits$best_dev_explained)),
                              'Non_explained_deviance_log' = log(1-c(potential_splits$first_dev, potential_splits$best_dev_explained)))
df_explained_dev$Non_explained_deviance_log = df_explained_dev$Non_explained_deviance_log-min(df_explained_dev$Non_explained_deviance_log)

par(mfrow = c(1,2), oma = c(2,2,1,1), mar = c(2,2,2,0.5), mgp = c(0.75,0.25,0), cex.axis=0.5, cex.lab=0.5, cex.main=0.7, cex.sub=0.5)
plot(df_explained_dev$N_groups,
     df_explained_dev$Non_explained_deviance,
     bty = 'n', ylim = c(0, ceiling(10*max(df_explained_dev$Non_explained_deviance))/10),
     xaxt = 'n', yaxt = 'n', pch = 16, main = 'linear scale', cex = 0.5, 
     ylab = 'Non-explained deviance (%)', xlab = 'Number of groups')
axis(1, lwd = 0.5, tck=-0.02)
axis(2, las = 2, at = seq(0,ceiling(10*max(df_explained_dev$Non_explained_deviance))/10,0.1),
     labels = seq(0, ceiling(10*max(df_explained_dev$Non_explained_deviance))/10,0.1)*100, lwd = 0.5, tck=-0.02)

plot(df_explained_dev$N_groups,
     (df_explained_dev$Non_explained_deviance),
     log = 'y',
     ylim = c(0.01, 1),
     bty = 'n',
     xaxt = 'n', yaxt = 'n', pch = 16, main = 'log scale', cex = 0.5, 
     ylab = 'Non-explained deviance (%) - log scale', xlab = 'Number of groups')
axis(1, lwd = 0.5, tck=-0.02)
axis(2, las = 2, at = c(0.01, 0.1, 0.25, 0.5, 1),
     labels = c(0.01, 0.1, 0.25, 0.5, 1)*100, lwd = 0.5, tck=-0.02)
```


Optimize the number of groups: set the minimum number of sequences per group to 30, with a minimum frequency of 1%.
```{r, eval = T}
split = merge.groups(timed_tree = tree_sars_cov2, metadata = dataset_with_nodes, 
                     initial_splits = potential_splits$potential_splits, 
                     group_count_threshold = 30, group_freq_threshold = 0.01)
```

Label sequences with these new groups, and assign a color to each of them.
```{r, eval = T}
## Label sequences with new groups
dataset_with_nodes$groups = as.factor(split$groups)
## Reorder labels by time of emergence
name_groups = levels(dataset_with_nodes$groups)
time_groups_world = NULL
for(i in 1:length(name_groups)){
  time_groups_world = c(time_groups_world, min(dataset_with_nodes$time[which(dataset_with_nodes$groups == name_groups[i] &
                                                                                   dataset_with_nodes$is.node == 'no')]))
}
levels(dataset_with_nodes$groups) = match(name_groups, order(time_groups_world, decreasing = T))
dataset_with_nodes$groups = as.numeric(as.character(dataset_with_nodes$groups))
dataset_with_nodes$groups = as.factor(dataset_with_nodes$groups)
## Update names in split list
split$tip_and_nodes_groups = match(split$tip_and_nodes_groups, order(time_groups_world, decreasing = T))
names(split$tip_and_nodes_groups) = 1:length(split$tip_and_nodes_groups)
split$groups = as.factor(split$groups)
levels(split$groups) = match(name_groups, order(time_groups_world, decreasing = T))
split$groups = as.numeric(as.character(split$groups))
## Choose color palette
n_groups <- length(name_groups)
colors_groups = (met.brewer(name="Cross", n=n_groups, type="continuous"))
## Color each group
dataset_with_nodes$group_color = dataset_with_nodes$groups
levels(dataset_with_nodes$group_color) = colors_groups
dataset_with_nodes$group_color = as.character(dataset_with_nodes$group_color)
```


#### Plot tree & index below, with colors from index-defined groups

Plot the tree and index colored with the new groups:
```{r, eval = T}
par(mfrow = c(2,1), oma = c(0,0,0,0), mar = c(4,4,0,0))

## Tree
plot(tree_sars_cov2, show.tip.label = FALSE, 
     edge.color = 'grey', edge.width = 0.25,
     x.lim = c(min_year, max_year)-root_height)
tiplabels(pch = 16, col = dataset_with_nodes$group_color, cex = 0.3)
axisPhylo_NL(side = 1, root.time = root_height, backward = F,
             at_axis = seq(min_year, max_year, 0.5)-root_height,
             lab_axis = seq(min_year, max_year, 0.5), lwd = 0.5)

## Index colored by group
plot(dataset_with_nodes$time, 
     dataset_with_nodes$index, 
     col = adjustcolor(dataset_with_nodes$group_color, alpha.f = 1),
     bty = 'n', xlim = c(min_year, max_year), cex = 0.5,
     pch = 16, bty = 'n', #ylim = c(0, 1), 
     main = paste0(''), #log = 'y',
     ylab = 'Index', xlab = 'Time (years)', yaxt = 'n')
axis(2, las = 2)
# Color key
legend('topright', 
       legend = name_groups,
       fill = colors_groups, border = colors_groups,
       cex = 0.5, bty = 'n', ncol = 5)
```

#### Compare NextStrain groups and groups called with the index

Generate SARS-CoV-2 trees colored with each set of groups next to each other:
```{r, eval = T, results = 'hide', warning=FALSE, message=FALSE}
## Tree with index-defined groups
groups = matrix(dataset_with_nodes$groups[which(dataset_with_nodes$is.node == 'no')], ncol = 1)
colnames(groups) = 'groups'
rownames(groups) = dataset_with_nodes$name_seq[which(dataset_with_nodes$is.node == 'no')]
cols = as.character(colors_groups)
names(cols) = as.character(1:max(as.numeric(name_groups)))
plot_tree_sars_world_groups <- ggtree(tree_sars_cov2, mrsd=lubridate::date_decimal(max(times_seqs)), size = 0.10,
                   aes(color = as.character(dataset_with_nodes$groups))) + 
  scale_color_manual(values = cols)+theme_tree2()
plot_tree_sars_world_groups = gheatmap(plot_tree_sars_world_groups, groups, offset=0.1, width=0.10, 
                    colnames=FALSE, legend_title="Group", color=NA) +
  scale_fill_manual(values = (cols))+scale_y_continuous(expand=c(0, 0.3))+theme(legend.position = 'none')

## Tree with NextStrain clades
Nextstrain = matrix(dataset_with_nodes$Nextstrain_clade[which(dataset_with_nodes$is.node == 'no')], ncol = 1)
colnames(Nextstrain) = 'groups'
rownames(Nextstrain) = dataset_with_nodes$name_seq[which(dataset_with_nodes$is.node == 'no')]
cols_NextStrain = as.character(colors_clade)
names(cols_NextStrain) = clade_labels
plot_tree_sars_world_Nextstrain <- ggtree(tree_sars_cov2, mrsd=lubridate::date_decimal(max(times_seqs)), size = 0.10,
                                          aes(color = as.character(dataset_with_nodes$Nextstrain_clade))) + 
  scale_color_manual(values = cols_NextStrain)+
  theme_tree2(legend = 'none')
plot_tree_sars_world_Nextstrain = gheatmap(plot_tree_sars_world_Nextstrain, Nextstrain, offset=0.1, width=0.10, 
                     colnames=FALSE, legend_title="Group", color=NA) +
  scale_fill_manual(values = cols_NextStrain, na.value = 'white')+
  scale_x_reverse() + 
  scale_y_continuous(expand=c(0, 0.3))+
  theme(legend.position = 'none')
```

Plot the generated SARS-CoV-2 trees:
```{r, eval = T}
plot_grid(plot_tree_sars_world_groups, plot_tree_sars_world_Nextstrain,
          rel_widths = c(1, 1), labels = c('Automatic clades', 'NextStrain clades'), label_size = 10, label_x = c(0.1, 0.25), ncol = 2)
```


## Quantify the fitness of detected SARS-CoV-2 lineage

#### Run the fitness model
Quantify the fitness of each group you can run the code (this steps takes approximately <5 min on 3 cores):
```{r, eval = F, results = 'hide', warning=FALSE, message=FALSE}
start_time = Sys.time()
## Load and compile stan code (this can take a few minutes)
## If you get an error about path mapping to a previous user 
##make sure to go into the model/scripts folder and delete the old unix model before making this one
model_compiled <- cmdstan_model(stan_file = '2_Functions/Model_multinomial_logistic_birthdeath_lineage_fitness_20231220.stan')
## Run model on SARS-CoV-2 groups
res_fitness = estimate_rel_fitness_groups_with_branches(dataset_with_nodes = dataset_with_nodes,
                                                        tree = tree_sars_cov2,
                                                        min_year = 2020, 
                                                        window = 30/365,
                                                        model_compiled = model_compiled,
                                                        iter_warmup = 250, iter_sampling = 500, refresh = 50, seed = 1)
end_time = Sys.time()
print(end_time - start_time)
```
You might encounter a warning saying that '*alpha_true_GA*' has a missing init value - this is normal as those groups (ancestral groups that are not present at the start of the time series) do not always exist and therefore there is no default initial value. This is does not impact the model run. The seed has been set to 1 so allow for reproducible results.

To save some time, you may wish to load the results:
```{r, eval = T}
res_fitness = readRDS('README_files/res_fitness.rds')
```

#### Plot the fits and estimated parameters
Plot the fits:
```{r}
order_colors = order(as.numeric(split$tip_and_nodes_groups))
colour_lineage = colors_groups[match(split$tip_and_nodes_groups[order_colors], name_groups)]

plot_fit_data_new(data = res_fitness$data,
                  Chains = res_fitness$chains,
                  colour_lineage = colour_lineage,
                  xmin = 2020, xmax = 2023.5)
```

Plot the predicted vs observed proportions:
```{r}
plot_observed_vs_predicted(data = res_fitness$data,
                           Chains = res_fitness$chains,
                           colour_lineage = colour_lineage)
```

Plot raw fitness estimates:
```{r}
plot_estimated_fitness_ref_ancestral(data = res_fitness$data,
                                     Chains = res_fitness$chains,
                                     colour_lineage = colour_lineage, 
                                     gentime = 1)
```






